### 2020-03-24 Analysis/mokken.Rmd
mokken

---
title: "Mokken scale analysis"
output: 
  html_document:
    code_folding: "hide"
    theme: cosmo
    toc: true
    toc_float: true
    tables: TRUE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(drake)
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
library(citippy)
library(mokken)
citippy::citippy_path(here::here("Analysis/references.bib"))
citippy::citippy_options(pandoc = FALSE)

Item <- c("My mouth is dry",
          "My lips are dry",
          "My tongue is thick",
          "My saliva is thick",
          "My throat is dry",
          "I have a bad taste in my mouth",
          "I want to drink water")
```

## Preparation

First we will load in a dataframe of items included in the PTDS.

```{r }
reactable::reactable(data = readd(ptds_factor),
                     resizable = TRUE, 
              showPageSizeOptions = TRUE, 
              onClick = "expand", 
              highlight = TRUE,
              filterable=TRUE
              )
```

### Correlation plot

```{r fig.align="center"}
## convert to numeric and drop one participant with missing data
readd(corr_plot)
```

## Automatic item selection procedure

The first step in Mokken scale analysis is to determine how many scales there may be in the bank of items that were assessed.  The `aisp` fuction is used to determine which items cluster with each other. We will use a lower bound of 0.5 to determine clustering of items.

```{r}
knitr::kable(readd(aisp), caption = "aisp for increasing H thresholds (*c*)")
```

Based on the aisp table, the item `bad_taste` is excluded by selecting the remaining items which show unidimensionality at a threshold level of .50. No multi-dimensional solution is apparent from this table (no groups of items identified as ‘leaving to form another scale’ at the same homogeneity threshold).

The 6 items are further examined below for MSA criteria.

## Scalability coefficient

The next step is to calculate the scalability coefficient **H**, which is an index of homogeneity for item pairs, using the `coefH` function.

```{r results='asis'}
knitr::kable(readd(H_values_ptds6)$Hi, caption = "Homogeneity values (and standard errors) for items")
```

The complete item set has a homogeneity value H(se) = `r readd(H_values_ptds6)$H`. These results indicate a strong Mokken scale for the items in the PTDS.

## Monotonicity

Results indicate that there were no significant violations of the assumption of monotonicity (z-sig). As such, no items were removed.

```{r}
knitr::kable(readd(monotonicity_df_ptds6), caption = "monotonicity with default minsize")
```

## Inspection of item pairs

To assess intersection and proximity, plots of item-pairs were produced. We can see that only one of the item-pair plots displays intersection (dry throat and dry lips).

```{r fig.align='center'}
plot(readd(iio_ptds6),  plot.ci=FALSE, ask = FALSE)
```

## Invariant item ordering

Analysis of invariant item ordering using the `check.iio` command revealed violations for the dry throat/dry lips item pair, but there were no *significant* intersections between items. Crit values less than 40 are considered acceptable.`r citippy(c('molenaar2000mps5'))` Due to the intersection between the dry throat and dry lips items, crit values were above 40. For this reason, we will exclude one of these items (dry throat), and re-assess the items for the assumptions of Mokken scaling.

```{r}
knitr::kable(summary(readd(iio_ptds6))$item.summary)
```

With the dry throat item removed, we now see that none of the items violate the assumption of invariant item ordering.

```{r}
knitr::kable(summary(readd(iio_ptds5))$item.summary)
```

## Accuracy

The accuracy of the item ordering was found to be acceptable with a H~T~ coefficient of `r readd(accuracy_ptds5)`. Values below 0.3 are unacceptable and above 0.4 indicates moderate invariant item ordering.`r citippy('Ligtvoet_2010')`

## Reliability

Scale reliability was adequate with a coeffient score (Rho) of `r readd(reliability_ptds5)`.

## Item difficulty hierarchy

```{r results = 'asis'}
knitr::kable(readd(item_hierarchy_df_ptds5), caption = "Item hierarchy")
```

The 5-item PTDS has a homogeneity value H(se) = `r readd(H_values_ptds5)$H`.

### 2020-03-24 Analysis/grant_plots.R

@@ -1,0 +1,163 @@
+library(dplyr)
+library(redcapAPI)
+library(lubridate)
+library(ggplot2)
+library(gridExtra)
+library(tidyr)
+library(ggtext)
+library(devtools)
+
+### Install development version of ggplot and run code in developer mode
+# move to development mode
+dev_mode(TRUE)
+# # Run if not already installed 
+# install_github("tidyverse/ggplot2")
+
+### Load in data and prepare for manual coding
+rcon <- redcapConnection(url="https://redcap.thetacollaborative.ca/api/", token="CABFAFF8E4259517B23F344818EE682F")
+myData <- exportRecords(rcon, labels = FALSE)
+
+# Export data to csv for qualitative coding
+fasting_descriptions <- myData %>% select(id, fasting_description)
+write.csv(fasting_descriptions, file="./Analysis/fasting_descriptions.csv")
+
+### Import latest version of manually coded data
+coded_descriptions <- read.csv("./Analysis/fasting_descriptions_coded20200130.csv")
+
+### Data cleaning
+# Remove ambiguous codings
+coded_descriptions <- coded_descriptions %>% filter(is.na(ambiguous))
+
+# Add summary variables
+coded_descriptions <- coded_descriptions %>%
+  mutate(fasting_type =  case_when(npo_midnight == 1 ~ 1,
+                                   npo_morning == 1 ~ 2,
+                                   npo_hrsprior == 1 ~ 3,
+                                   npo_dinner == 1 ~ 1,
+                                   separate_fooddrink == 1 ~ 4,
+                                   unclear_patient == 1 ~ 6,
+                                   missing == 1 ~ 7)) %>%
+  mutate(fasting_type_text = case_when(fasting_type %in% 1 ~ "\"No food or drink after midnight\"",
+                                       fasting_type == 2 ~ "\"No food or drink after __ AM\"",
+                                       fasting_type == 3 ~ "\"No food or fluids __ hours prior\"",
+                                       fasting_type == 4 ~ "\"No food __ hours prior and no\ndrink after __ AM this morning\"",
+                                       fasting_type == 6 ~ "Unclear instructions",
+                                       fasting_type == 7 ~ "Missing"))
+
+# Join coded data to original data
[…]

### 2020-03-24 R/functions.r

@@ -1,0 +1,275 @@
+#' @param data dataframe from csv downloaded from redcap
+
+#' @export
+label_data <- function(data){
+
+#Setting Units
+
+
+#Setting Factors(will create new variable for factors)
+data$redcap_event_name_factor = factor(data$redcap_event_name,levels=c("t3_arm_1","t1_arm_2","t2_arm_2","t3_arm_2","screening_arm_3"))
+data$datetime_v1_complete_factor = factor(data$datetime_v1_complete,levels=c("0","1","2"))
+data$sex_factor = factor(data$sex,levels=c("1","2","3"))
+data$procedure_factor = factor(data$procedure,levels=c("1","2","3","4","5","6","7","8","9","10","11","12","13"))
+data$instructions_factor = factor(data$instructions,levels=c("1","2","3","4","5","6"))
+data$baseline_v1_complete_factor = factor(data$baseline_v1_complete,levels=c("0","1","2"))
+data$ptds_dry_mouth_factor = factor(data$ptds_dry_mouth,levels=c("0","1","2"))
+data$ptds_dry_lips_factor = factor(data$ptds_dry_lips,levels=c("0","1","2"))
+data$ptds_thick_tongue_factor = factor(data$ptds_thick_tongue,levels=c("0","1","2"))
+data$ptds_thick_saliva_factor = factor(data$ptds_thick_saliva,levels=c("0","1","2"))
+data$ptds_dry_throat_factor = factor(data$ptds_dry_throat,levels=c("0","1","2"))
+data$ptds_bad_taste_factor = factor(data$ptds_bad_taste,levels=c("0","1","2"))
+data$ptds_water_factor = factor(data$ptds_water,levels=c("0","1","2"))
+data$global_thirst_factor = factor(data$global_thirst,levels=c("1","2","3","4","5","6"))
+data$thirst_intensity_factor = factor(data$thirst_intensity,levels=c("0","1","2","3","4","5","6","7","8","9","10"))
+data$pain_factor = factor(data$pain,levels=c("0","1","2","3","4","5","6","7","8","9","10"))
+data$o2_therapy_factor = factor(data$o2_therapy,levels=c("1","0"))
+data$perioperative_thirst_discomfort_scale_v1_complete_factor = factor(data$perioperative_thirst_discomfort_scale_v1_complete,levels=c("0","1","2"))
+data$thirst_change_factor = factor(data$thirst_change,levels=c("1","2","3","4","5"))
[…]

### 2020-03-24 R/plan.R

@@ -1,0 +1,65 @@
+library(drake)
+library(tidyverse)
+library(here)
+library(mokken)
+library(gtsummary)
+
+conflicted::conflict_prefer("select", "dplyr")
+conflicted::conflict_prefer("filter", "dplyr")
+
+source("R/functions.R")
+
+plan <- drake_plan(
+  original_data = readr::read_csv(here("data/ThirstDiscomfortScal_DATA_2020-03-24_2016.csv")),
+  data_labelled = label_data(original_data), # Label original data
+  rel1 = data_labelled %>%
+    filter(str_detect(id, 'R')) %>% #remove screening data
+    filter(redcap_event_name!="t2_arm_2", redcap_event_name!="t3_arm_2"),# get participant data for this arm
+  participation_df = data_labelled %>%
+    filter(str_detect(id, 'R')|str_detect(id, 'V')) %>% #remove screening data
+    filter(redcap_event_name!="t1_arm_2") %>% # remove additional timepoints for test-retest (only n=5)
+    filter(redcap_event_name!="t2_arm_2") %>%
+    coalesce_join(rel1, by = "id") %>% # join participant data from t1 for reliability arm
+    drop_na(starts_with("ptds")) %>% # remove one participant who missed one item
+    mutate(fluids_duration = round(as.numeric(difftime(date_time,fluids, units = "hours"))),1) %>% #time since last fluids
+    mutate(food_duration = round(as.numeric(difftime(date_time,food, units = "hours"))),1),
+  #mokken scaling
+  ptds_factor = participation_df %>%
+    select(starts_with("ptds")) %>%
+    select(ends_with("factor"))  %>%
+    rename_all(.funs = ~str_remove_all(., "ptds_")) %>%
+    rename_all(.funs = ~str_remove_all(., "_factor")),
+  ptds7  =  ptds_factor %>%
+    mutate_if(is.factor, ~ as.numeric(.x)),
+  corr_plot = ggcorrplot::ggcorrplot(cor(ptds7),
+                                     colors = c( "#E46726", "white","#6D9EC1")),
+  aisp = moscales.for.lowerbounds(x= as.data.frame( ptds7)),
+  ptds6 = ptds7 %>%
+    select(-bad_taste),
+  H_values_ptds6 = create_H_values(data = ptds6),
+  monotonicity_df_ptds6 = create_monotonicity_df( data = ptds6),
[…]

### 2020-03-24 Screening_app/app.R

@@ -1,0 +1,624 @@
+library(shiny)
+library(shinydashboard)
+library(redcapAPI)
+library(tidyverse)
+library(stringr)
+library(lubridate)
+library(flexdashboard)
+library(ggpubr)
+library(bizdays)
+library(qwraps2)
+library(purrr)
+library(kableExtra)
+library(Hmisc)
+library(plotly)
+library(waiter)
+
+
+
+
+
+###########################  UI  ###############
+
+ui <- 
+  
+  
+  dashboardPage(
+    dashboardHeader(title = "Thirst discomfort scale"),
+    dashboardSidebar(
+      sidebarMenu(
+        br(),
+        # menu item tabs
+        menuItem("Recruitment", tabName="recruitment", icon=icon("search")),
+        menuItem("Fasting duration", tabName="fasting-duration", icon=icon("exclamation-circle")),
+        menuItem("Thirst discomfort", tabName="thirst-discomfort", icon=icon("chart-area"))
+      )
+       
+    ),
+    dashboardBody(
+      
+      use_waiter(include_js = FALSE), # do not include js
+      
+      tags$head(tags$style(HTML('
+        .skin-blue .main-header .logo {
+          background-color: #3c8dbc;
+        }
+        .skin-blue .main-header .logo:hover {
+          background-color: #3c8dbc;
+        }
+      '))),
+      
+      tabItems(
+        tabItem(tabName="recruitment",
+          fluidRow(
+              column(width = 4,
+               flexdashboard::valueBoxOutput("sysdate"),
+               
+               
+               flexdashboard::valueBoxOutput("Screened"),
+               
+               
+               flexdashboard::valueBoxOutput( "Eligible"),
+               
+               
+               flexdashboard::valueBoxOutput("Consented"),
+               
+               
+               flexdashboard::valueBoxOutput("estdatecompletion")
+               
+              ),
+              column(width = 4,
+               box( width = NULL,
+                    flexdashboard::gaugeOutput("gaugeEligibility"), height = 150, title = "Eligibility Rate", solidHeader = TRUE,
+                    status = "primary"),
+               
+               box( width = NULL,
+                    flexdashboard::gaugeOutput("gaugeRecruitment"),height = 150,title = "Recruitment Rate", solidHeader = TRUE,
+                    status = "primary"),
+               
+               box(width = NULL,
[…]

### 2020-06-08 manuscript/index.Rmd
manuscript

---
title: "Validating the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures"
author:
  - Kristina Chang MScN:
      institute: [PMCC]
      correspondence: false
  - Megan Bittner BSc:
      institute: [LBFON]
      correspondence: false
  - Dan Phan BSc:
      institute: [LBFON]
      correspondence: false
  - Navpreet Kamboj BScN:
      institute: [LBFON]
      correspondence: false
  - Matteo Parotto MD, PhD:
      institute: [TGH, UoT]
      correspondence: false
  - Amanda Matthews:
      institute: [PMCC]
      correspondence: false
  - Sherryl Alexander:
      institute: [PMCC]
      correspondence: false
  - Pauline Glaves:
      institute: [PMCC]
      correspondence: false
  - Shahvand Masihi:
      institute: [IR]
      correspondence: false
  - Sebastian Mafeld:
      institute: [IR]
      correspondence: false
  - Aaron Conway PhD:
      email: aaron.conway@utoronto.ca
      phone: +1 416 946 7112
      address: 585 University Ave | Toronto, ON  M5G 2N2
      institute: [PMCC, LBFON, QUT]
      correspondence: true
institute:
  - PMCC: Peter Munk Cardiac Centre, University Health Network, Toronto, Canada
  - LBFON: Lawrence S. Bloomberg Faculty of Nursing, University of Toronto, Toronto, Canada
  - QUT: School of Nursing, Queensland University of Technology (QUT), Brisbane, Australia
  - TGH: Department of Anesthesia and Pain Management, Toronto General Hospital, UHN, Toronto, Canada
  - UoT: Department of Anesthesia and Interdepartmental Division of Critical Care Medicine, University of Toronto, Toronto, Canada
  - IR: Interventional Radiology, JDMI, Toronto General Hospital
bibliography: Extras/references.bib
csl: Extras/sp-basic.csl
output:
  officedown::rdocx_document:
    reference_docx: Extras/manuscript_template.docx
    pandoc_args:
      - --lua-filter=Extras/scholarly-metadata.lua
      - --lua-filter=Extras/author-info-blocks.lua
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

library(drake)
library(magrittr)
library(flextable)
```

\pagebreak

### Abstract

**Introduction**

**Methods**

**Results**

**Conclusion**

Clinical trial number: Not applicable

Keywords: fasting, aspiration, patient safety, thirst, anesthesia, nursing, Mokken scale, scale development, item response theory

\pagebreak

## Declarations

Availability of data and material (data transparency): All data used is available [here](https://aconway.dev/ptds).

Code availability: Code to reproduce the analyses is available [here](https://github.com/awconway/ptds)<!-- and is archived [here]()-->.

Abbreviated title: Perioperative thirst discomfort scale

Funding statement: Support was provided solely from institutional and/or departmental sources. MP is supported by an Early Investigator Merit Award from the Department of Anesthesia, University of Toronto and Toronto General Hospital.

Conflicts of interest: The authors declare no competing interests.

Author contributions: 
KC: Study design, data analysis and wrote manuscript; AC: Study design, analysis, wrote manuscript;  MB: Data collection, revised the manuscript for important intellectual content; DP: Data collection, revised the manuscript for important intellectual content; NK: Study design, revised the manuscript for important intellectual content; AM: Study design, revised the manuscript for important intellectual content; MP: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; PG: Study design, revised the manuscript for important intellectual content; SA: Study design, revised the manuscript for important intellectual content.

\pagebreak

# Introduction

Pre-procedure fasting has been historically implemented to reduce the risk of vomiting and aspiration pneumonia with general anaesthesia, or in case emergency intubation is required due to unexpected cardiac arrest.[@hamid2014pre; @osborne2002preoperative] However, prolonged fluid restriction causes thirst symptoms to develop (e.g., dry mouth, swollen tongue), which can lead to great discomfort.[@madsen1998perioperative] Current guidelines related to pre-procedure fasting for elective procedures recommend a minimum fasting period of 2 hours nil-per-os (NPO) for clear fluids.[@dobsonGuidelinesPracticeAnesthesia2018] Despite these recommendations, current practice is for patients undergoing surgical and other medical procedures that require sedation or anesthesia to receive standardized 'nil-by-mouth' fasting instructions at a pre-specified time interval before procedures. For example, 'no eating or drinking after midnight' is most common. This 'standardized' instruction will not be changed regardless of whether or not there are alterations in scheduling throughout the day that result in significant delays in procedure start time. As a result, fasting durations far exceed the recommended requirement for most patients undergoing medical and surgical procedures.[@de2014actual; @sorita2015frequency; @spitz2017impact] For example, in a recent study of 3641 fasting orders at a large academic institution in the USA, it was found that the median fasting duration was 12.8 hours, averaging 2 missed meals.[@sorita2015frequency]

As a direct result of prolonged pre-procedure fasting, symptoms of thirst discomfort have been reported as common and severe. In a qualitative study where 12 participants were interviewed from a tertiary hospital in Australia, surgical patients and patients who adhered to prolonged fasting instructions (e.g., NPO for two days) for tests, procedures or symptom management (e.g., bowel obstruction) described the discomfort from thirst symptoms to be more unbearable than hunger.[@carey2015qualitative] Similarly, Madsen et al. interviewed a convenience sample of 50 adult surgical patients who reported that thirst symptoms caused more discomfort than hunger, sleep, or anxiety related to the procedure.[@madsen1998perioperative] However, thirst and its symptoms continue to be undervalued, under-reported and infrequently assessed by health care providers, including the nursing team.[@milani2016thirst]

Despite the relevance and value of assessing thirst-discomfort of patients, the subjective experience of thirst presents challenges in developing a valid and reliable tool to succinctly and accurately measure its symptoms and level of discomfort prior to procedures. A thirst-discomfort scale for perioperative use was developed at a surgical centre affiliated with an accredited public hospital in Brazil.[@Martins_2017] The perioperative thirst-discomfort tool was developed in three stages, including face and content validation and reliability, and based on the Consensus-based Standards for the selection of health status Measurement Instruments (COSMIN) checklist.[@mokkink2010cosmin] Inter-rater reliability was tested through inter-observer equivalence where a pair of nurses independently, but simultaneously administered the scale among 70 patients. Six items on the scale had a weighted kappa coefficient of 1, while the item "I feel like drinking water" was 0.97.[@Martins_2017] Cronbach's alpha was used to test the internal consistency of the scale with a value of 0.91.[@Martins_2017]

Although the Perioperative Thirst Discomfort Scale was developed in a different context (e.g., surgical use), it demonstrates strong potential for accurately assessing thirst-discomfort in patients before medical procedures more generally. For this reason, the aim of this study was to test the validity and reliability of the Perioperative Thirst Discomfort Scale for measuring thirst discomfort prior to procedures.

# Methods

## Study Design

This study used a prospective, observational design. No changes to usual clinical practice were made for this study in regard to pre-procedure fasting. Patients who were scheduled for a morning procedure were typically asked to remain nil-per-os (NPO) from midnight. Patients with afternoon procedures can have a small breakfast but must remain NPO from 6:00am on the day of the procedure. Ethics approved by....

## Participants

### Inclusion criteria

Adult patients undergoing an elective procedure in the Cardiac Catheterization Laboratories or in Interventional Radiology at the Toronto General Hospital were included.

### Exclusion criteria

1. Under 16 years of age
2. Undergoing an emergency procedure
3. Unable to understand or speak English and a translator is not immediately available
4. Nurse in pre-procedural bay considered that there was insufficient time prior to anticipated commencement of the procedure for participation in the study.

## Data collection

A Research Assistant administered a brief questionnaire prior to procedures. It comprised the following components:

- the Perioperative Thirst Discomfort Scale (7 items);
- a one-item global thirst discomfort rating;
- a one-item thirst intensity rating;
- an item to determine if the participant is currently on oxygen therapy;
- an item to evaluate the presence of pain; 
- the time the patient last had any clear fluids;
- the time the patient last had any food;
- age, sex of patient and type of procedure to be performed.

*Seven-item Perioperative Thirst Discomfort Scale (PTDS-7)*

The PTDS-7 is a 7-item self-reported, composite score evaluating the severity of dry mouth, lips, and throat, thick tongue and saliva, a bad taste in the mouth, and a desire to drink water. The total score can range from 0 to 14, with 14 representing the "most extreme" intensity of discomfort related to perioperative thirst. This instrument was developed and validated for perioperative use in a surgical centre at a public hospital in Brazil. Face validation was conducted by group consisting of nurses with PhDs, new graduate nurses, and patients. Content validation was conducted by a team consisting of specialists with perioperative nursing experience related to thirst or instrument validation. The reliability of the scale was assessed through inter-observer equivalence by a research nurse and a staff nurse from the Post Anesthesia Care Unit among 70 patients pre and postoperatively in an anesthetic recovery room. The Cronbach alpha score was 0.91. The time taken to complete the Perioperative Thirst Discomfort Scale was not evaluated.

*Global thirst discomfort score*

Participants were asked to rate their overall level of thirst discomfort on a rating scale with score ranging from "extreme thirst discomfort" to "extreme comfort".

*Thirst intensity rating*

Participants will be asked to rate the intensity of their thirst on a scale ranging from 0 to 10, where 0 represents "No thirst" and 10 represents "Most intense thirst".

*Pain scale rating*

The experience of pain may influence the participant's perception of thirst.[@Pierotti_2018] Participants were asked to rate their current level of pain on a scale of 0 to 10, where 0 represents "no pain" and 10 represents that their pain as "unimaginable/unspeakable".

*Use of oxygen therapy*

The use of oxygen therapy (e.g., via nasal prongs) may contribute to the experience of thirst symptoms (e.g., dry mouth, dry throat).[@conchon2015perioperative]. The RA recorded "yes" or "no" as to whether the participant was currently receiving oxygen therapy at the time the questionnaire was administered.

## Statistical analysis

Mokken scaling analyses properties of individual items as described by the item characteristic curve. Item characteristic curves show the association between the score on an item to the level of the latent trait being measured. Mokken scaling makes no assumptions about this association, apart from the assumption of monotone homogeneity. This means that when the trait increases, so does the item score. "Difficulty", is a term used in Mokken Scale Analysis to refer to the extent to which items are endorsed by respondents. Items at the upper end of the range of the latent trait being more difficult. Items in a Mokken Scale are arranged along the latent trait in terms of their difficulty. The properties of items can be measured using the scalability coefficient, *H* (Loevinger's coefficient). *H* measures the extent to which items are arranged as expeceted by their mean values along the latent trait. *H* > 0.3 is the minimum acceptable value, with values > 0.4 indicating a moderate scale and > 0.5 indicates a strong scale. Invariant item ordering, where where the order of items along the latent trait is the same for all respondents at all levels of the latent trait, is a desirable property of a Mokken Scale. It can be assessed mathematically to look for significant violations of this property. The accuracy of the  invariant item ordering can also be assessed by calculating the H~TRANS~ (H~T~) coefficient. Values of H~T~ exceeding 0.3 indicates acceptable accuracy of invariant item ordering.

Mokken scale analysis was conducted in this study to explore whether there were hierarchical properties in participant ratings of thirst discomfort and to explore the dimensions of the PTDS. Mokken scale analysis proceeded by first checking the PTDS-7 scalability coefficients. If H~i~ was below 0.3 or if the lower limit of the 95% CI (confidence interval) for H~i~ was below 0.3, the item was excluded. Then, scale partitioning was carried out to explore the dimensions of the PTDS through increasing c (Lower bound c defines the minimum value of coefficients Hi in the Mokken scale by 0.05 increments).[@molenaar2000mps5] Monotone homogeneity and invariant item ordering were investigated at the whole scale level, as no sub-scales were identified.

To assess invariant item ordering, crit values <40 were considered acceptable.[@molenaar2000mps5]

Accuracy - Values below 0.3 are unacceptable and above 0.4 indicates moderate invariant item ordering.[@Ligtvoet_2010]

Construct validity was evaluated by using correlations to identify associations between scores on the Perioperative Thirst-Discomfort Scale and fasting duration. To demonstrate construct validity we hypothesised that higher scores on the Perioperative Thirst Discomfort Scale would be associated with greater fasting duration. Convergent validity was assessed by using Kendall's tau to compare the Perioperative Thirst Discomfort Scale score with the global thirst discomfort score. Convergent validity was assessed by using the Spearman's rank correlation to compare the Perioperative Thirst Discomfort Scale with the thirst intensity rating and pain scale rating. Chi-square analysis was used to evaluate the association between Perioperative Thirst Discomfort Scale score and use of oxygen supplementation.

# Results

## Participants

From November 2019 to March 2020, we screened 203 patients for inclusion in the study. A total of 198 were eligible and 193 chose to participate. A summary of participant characteristics is displayed in Table \@ref(tab:tab1). The median age was 62 (IQR 48 - 72) and 58% (n=111) were female. 43% of patients were scheduled to have their procedure in the cardiac catheterization laboratory. A biopsy was the most common Interventional Radiology procedure (n=62; 32%), but a wide range of other procedures were included.

## Mokken scale analysis

No items required removal from the PTDS-7 based on the condition that scalability coefficients (H~i~ and lower bound of 95% CI) should be over 0.3 (Appendix Table \@ref(tab:Hvaluesptds7)). We used a lower bound for *c*, starting from 0.05 and increasing to 0.80 in 0.05 increments, to explore the dimensionality of the PTDS. From 0.05 to 0.45, all of the items formed a single scale. The item, *'I have a bad taste in my mouth'*, was excluded by selecting the remaining items which showed uni-dimensionality at a threshold level of .50. The 6 remaining items (PTDS-6) were further examined for the criteria of a Mokken scale and item invariant ordering. The PTDS-6 item set has a homogeneity value H(se) of `r readd(H_values_ptds6)$H`, which indicated a strong Mokken scale. Results indicated that there were no significant violations of the assumption of monotonicity (z-sig). As such, no items were removed based on this criteria.

### Invariant item ordering

Analysis of invariant item ordering for the PTDS-6 revealed violations for the dry throat/dry lips item pair (Appendix Table \@ref(tab:iioptds6)). Although there were no *significant* intersections between items, due to the intersection between the dry throat and dry lips items, crit values were above 40. For this reason, we excluded one of these items (dry throat), and re-assessed the remaining 5 items for the assumptions of a Mokken scale. With the dry throat item removed, none of the items violated the assumption of invariant item ordering (Appendix Table \@ref(tab:iioptds5)).

### Accuracy and reliability

The accuracy of the item ordering was found to be acceptable with a H~T~ coefficient of `r drake::readd(accuracy_ptds5)`. Scale reliability was adequate with a coefficient score (Rho) of `r drake::readd(reliability_ptds5)`.

### Item difficulty hierarchy

The most highly endorsed symptom (i.e. the least severe symptom of thirst discomfort in the PTDS-5) was being bothered by a desire to drink water (Table \@ref(tab:tab2)). The least endorsed symptom (i.e. the most severe symptom of thirst discomfort in the PTDS-5) was being bothered by the perception of a 'thick tongue'. The items in the PTDS-5 form a strong Mokken scale (H~s~ = `r drake::readd(H_values_ptds5)$H[1]`), meaning it is a reliable and precise way to order patients according to their thirst discomfort.

## Construct validity

PTDS-5 and Age

PTDS-5 and Food/fluid duration

PTDS-5 and global thirst

PTDS-5 and thirst intensity

PTDS-5 and pain

# Discussion

The items in the PTDS-5 form a strong Mokken scale (H~s~ = `r drake::readd(H_values_ptds5)$H[1]`), meaning it is a reliable and precise way to order patients according to their thirst discomfort. In addition, an acceptable accuracy for invariant item ordering was identified, meaning patients are likely to rank the PTDS-5 symptoms  of thirst discomfort in the same way. Discuss clinical usefulness of the scale... Comment on the results for the item difficulty hierarchy (least severe symptom to most severe symptom).

Contrast with results of previous study...The perioperative thirst-discomfort tool has previously been used to determine the relationship between thirst intensity and thirst-discomfort in the immediate postoperative period during anesthesia recovery.[@Pierotti_2018]  In a non-probabilistic sample of 203 adult participants awaiting elective, urgent and emergency surgeries, the investigators found no association between the fasting time, or patient age, and the intensity of thirst and degree of discomfort.[@Pierotti_2018] However, the mean fasting time among participants was quite high at 16.2 hours (SD 8.7 hours), which may have prevented the researchers from discerning correlation between fasting time and thirst as even with 8 hours of fasting all participants reported thirst.[@aroni2012assessment; @Pierotti_2018] There was no correlation between age, and intensity of thirst and thirst-discomfort, but few elderly patients were recruited to the study with a mean age of 41.5 years old (SD 15.8 years).

- The dimensionality of the PTDS had not been performed prior to our study. A multi-dimensional solution was not apparent, although this was to be expected with the small number of items that were evaluated.

- Lack of association between PTDS-5 and fasting duration. As such it is important to assess thirst discomfort periodically, regardless of the duration of fasting. If thirst discomfort is present, and fasting is still indicated, interventions to alleviate symptoms should be implemented. It should be noted, though, that there is a lack of evidence regarding the efficacy of any such interventions...

## Limitations

- convenience sample (selection bias can not be ruled out)
- generalize to other procedural settings?
- Low amount of missing data 
- Further research required to evaluate the minimal detectable difference in the PTDS-5 and the minimal clinically important difference...

## Conclusion

\pagebreak

# References

<div id="refs"></div>

\pagebreak

# Figure legend

Fig. \@ref(fig:intensity) Association between PTDS-5 and thirst intensity

Fig. \@ref(fig:globalthirst) Association between PTDS-5 and global thirst rating

Fig. \@ref(fig:fasting) Association between PTDS-5 and duration of fasting

\pagebreak

```{r tab.cap="Participant characteristics", tab.id = 'tab1'}
drake::loadd(summary_table)
gtsummary::as_flextable(summary_table)
```

\pagebreak

```{r tab.id = "tab2", tab.cap= "Item hierarchy of the PTDS-5"}
flextable::flextable(drake::readd(item_hierarchy_df_ptds5)) %>% 
  autofit()
```

<!---BLOCK_MULTICOL_START--->

<!---BLOCK_STOP--->

<!--BLOCK_LANDSCAPE_START--->

```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 5 to 15 with higher scores indicating worse thirst discomfort."}
knitr::include_graphics("figures/thirst-intensity.png")
```

\pagebreak

```{r globalthirst, fig.cap="Association between PTDS-5 and global thirst rating"}
knitr::include_graphics("figures/global-thirst.png")
```

\pagebreak

```{r fasting, fig.cap = "Association between PTDS-5 and duration of fasting"}
knitr::include_graphics("figures/fasting.png")
```

\pagebreak

# Appendix

```{r tab.id='Hvaluesptds7', tab.cap.pre="Appendix Table ", tab.cap = "Homogeneity values (and standard errors) for all items in the PTDS"}
drake::readd(H_values_ptds7_table) %>% 
  flextable::flextable() %>% 
  autofit()
```

\pagebreak

```{r tab.id='monotonicity', tab.cap.pre="Appendix Table ", tab.cap="monotonicity with default minsize"}

flextable::flextable(drake::readd(monotonicity_df_ptds6_table)) %>%  
  autofit()
```

\pagebreak

```{r tab.id='aisp', tab.cap.pre="Appendix Table ", tab.cap = "aisp for increasing H thresholds (*c*)"}
flextable::flextable(drake::readd(aisp)) %>% 
  autofit()
```

\pagebreak

```{r tab.id='iioptds6', tab.cap.pre="Appendix Table ", tab.cap = "Invariant item ordering for PTDS-6"}
drake::readd(iio_ptds6_table) %>% 
  flextable() %>% 
  autofit()
```

\pagebreak

```{r tab.id='iioptds5', tab.cap.pre="Appendix Table ", tab.cap = "Invariant item ordering for PTDS-5"}
drake::readd(iio_ptds5_table) %>%
    flextable() %>%
    autofit()
```

<!---BLOCK_LANDSCAPE_STOP--->

### 2020-06-08 R/add_ptds_total.R

@@ -1,0 +1,12 @@
+
+##' @title
+
+##' @export
+#' @importFrom dplyr rowwise c_across everything summarise
+add_ptds_total <- function(ptds_df, original_df) {
+  ptds_total <- ptds_df %>%
+    rowwise() %>%
+    summarise(ptds = sum(c_across(everything())))
+
+  original_df %>% cbind(ptds_total)
+}

### 2020-06-08 R/coalesce_join.R

@@ -1,0 +1,31 @@
+##' @title Coalesce Join
+##' @description https://alistaire.rbind.io/blog/coalescing-joins/
+##' @return
+##' @export
+
+#' @importFrom dplyr union coalesce bind_cols full_join
+#' @importFrom purrr map_dfc
+coalesce_join <- function(x, y,
+                          by = NULL, suffix = c(".x", ".y"),
+                          join = full_join, ...) {
+  joined <- join(x, y, by = by, suffix = suffix, ...)
+  # names of desired output
+  cols <- union(names(x), names(y))
+
+  to_coalesce <- names(joined)[!names(joined) %in% cols]
+  suffix_used <- suffix[ifelse(endsWith(to_coalesce, suffix[1]), 1, 2)]
+  # remove suffixes and deduplicate
+  to_coalesce <- unique(substr(
+    to_coalesce,
+    1,
+    nchar(to_coalesce) - nchar(suffix_used)
+  ))
+
+  coalesced <- map_dfc(to_coalesce, ~ coalesce(
+    joined[[paste0(.x, suffix[1])]],
+    joined[[paste0(.x, suffix[2])]]
+  ))
+  names(coalesced) <- to_coalesce
+
+  bind_cols(joined, coalesced)[cols]
+}

### 2020-06-08 R/create_DT.R

@@ -1,0 +1,15 @@
+##' @title
+
+##' @return
+##' @author Aaron Conway
+##' @export
+#' @importFrom dplyr select starts_with ends_with
+create_DT <- function(data) {
+  data %>%
+    select(
+      -starts_with("redcap"), -ends_with("complete"), -thirst_change,
+      -starts_with("screen"),
+      -ends_with("complete_factor"),
+      -studyid, -`1`
+    )
+}

### 2020-06-08 R/create_H_values.R

@@ -1,0 +1,10 @@
+##' @title
+##' @param data dataframe
+##' @return
+##' @author Aaron Conway
+##' @importFrom mokken coefH
+
+##' @export
+create_H_values <- function(data = ptds6) {
+  coefH(as.matrix(data))
+}

### 2020-06-08 R/create_H_values_ptds7_table.R

@@ -1,0 +1,34 @@
+##' @title
+
+##' @return
+##' @author Aaron Conway
+##' @importFrom dplyr tibble
+##' @importFrom stringr str_remove_all
+
+##' @export
+
+create_H_values_ptds7_table <- function(H_values_ptds7){
+
+  tibble(Item = c(
+  "My mouth is dry",
+  "My lips are dry",
+  "My tongue is thick",
+  "My saliva is thick",
+  "My throat is dry",
+  "I have a bad taste in my mouth",
+  "I want to drink water"
+), Hi = c(H_values_ptds7$Hi[1],
+          H_values_ptds7$Hi[2],
+          H_values_ptds7$Hi[3],
+          H_values_ptds7$Hi[4],
+          H_values_ptds7$Hi[5],
+          H_values_ptds7$Hi[6],
+          H_values_ptds7$Hi[7]),
+    SE = c(str_remove_all(H_values_ptds7$Hi[8], "[()]"),
+           str_remove_all(H_values_ptds7$Hi[9], "[()]"),
+                          str_remove_all(H_values_ptds7$Hi[10], "[()]"),
+                                         str_remove_all(H_values_ptds7$Hi[11], "[()]"),
+                                                        str_remove_all(H_values_ptds7$Hi[12], "[()]"),
+                                                                       str_remove_all(H_values_ptds7$Hi[13], "[()]"),
+                                                                                      str_remove_all(H_values_ptds7$Hi[14], "[()]")))
+}

### 2020-06-08 R/create_age_ptds_plot.R

@@ -1,0 +1,27 @@
+
+##' @title
+##' @return
+##' @author Aaron Conway
+##' @export
+#' @importFrom ggplot2 ggplot aes geom_jitter
+#' geom_smooth theme_minimal labs theme ggsave
+create_age_ptds_plot <- function(data) {
+  data %>%
+    ggplot(aes(y = ptds, x = age)) +
+    geom_jitter(alpha = 0.5) +
+    geom_smooth(method = "lm") +
+    theme_minimal() +
+    labs(
+      x = "\nAge",
+      y = "5-item peri-operative thirst discomfort scale (PTDS-5)\n",
+      title = "Association between PTDS-5 and age",
+      subtitle = "PTDS-5 score ranges from 5 to 15 with higher scores indicating worse thirst discomfort"
+    ) +
+    theme(
+      legend.position = "none",
+      plot.title.position = "plot"
+    )
+
+  ggsave(device = "png", filename = here("manuscript/figures/age.png"),
+         width = 174, units = "mm")
+}

### 2020-06-08 R/create_fasting_duration_ptds_plot.R

@@ -1,0 +1,84 @@
+##' @title
+##' @return
+##' @author Aaron Conway
+##' @export
+#' @importFrom ggplot2 ggplot aes geom_jitter geom_smooth
+#' theme_minimal labs theme ggsave
+#' @importFrom patchwork plot_layout
+#' @importFrom ggtext element_markdown geom_richtext
+#' @importFrom dplyr summarize mutate
+#' @importFrom glue glue
+create_fasting_duration_ptds_plot <- function(data) {
+
+  ptds_fluids_cor <- data %>%
+    drop_na(ptds, fluids_duration) %>%
+    summarize(r_square = cor(ptds, fluids_duration)^2) %>%
+    mutate(
+      # location of each text label in data coordinates
+      ptds = 14,
+      fluids_duration = 20,
+      # text label containing r^2 value
+      label = glue("*r*<sup>2</sup> = {round(r_square, 2)}")
+    )
+
+  fluids <- data %>%
+    ggplot(aes(y = ptds, x = fluids_duration)) +
+    geom_jitter(alpha = 0.5,
+                colour = "dodgerblue") +
+    geom_smooth(method = "lm") +
+    theme_minimal() +
+    labs(
+      x = "\nDuration of fasting from <span style='color:dodgerblue'>fluids</span> in hours",
+      y = "PTDS-5 (5-15; higher scores indicate worse thirst discomfort)\n"
+    ) +
+    theme(
+      legend.position = "none",
+      plot.title.position = "plot",
+      axis.title.x = element_markdown()
+
+    ) +
+    ggtext::geom_richtext(
+      data = ptds_fluids_cor,
+      aes(label = label),
+      hjust = 1, vjust = 1
+    )
+
+  ptds_food_cor <- data %>%
+    drop_na(ptds, food_duration) %>%
+    summarize(r_square = cor(ptds, food_duration)^2) %>%
+    mutate(
+      # location of each text label in data coordinates
+      ptds = 14,
+      food_duration = 25,
+      # text label containing r^2 value
+      label = glue("*r*<sup>2</sup> = {round(r_square, 2)}")
+    )
+
+ food <-  data %>%
+   ggplot(aes(y = ptds, x = food_duration)) +
+   geom_jitter(alpha = 0.5,
+               colour = "orangered") +
+   geom_smooth(method = "lm") +
+   theme_minimal() +
+   labs(
[…]

### 2020-06-08 R/create_global_thirst_plot.R

@@ -1,0 +1,25 @@
+##' @title
+##' @return
+##' @author Aaron Conway
+##' @export
+#' @importFrom ggbeeswarm geom_beeswarm
+#' @importFrom dplyr desc
+#' @importFrom ggplot2 ggplot aes theme_minimal
+#' labs element_blank theme ggsave
+create_global_thirst_plot <- function(data) {
+  data %>%
+    ggplot(aes(x = ptds, y = reorder(global_thirst_factor, desc(global_thirst_factor)),
+               colour = global_thirst_factor)) +
+    geom_beeswarm(groupOnX = FALSE, alpha = 0.5, size = 2) +
+    theme_minimal() +
+    labs(
+      y = element_blank(),
+      x = "\nPTDS-5 (5-15; higher scores indicate worse thirst discomfort)"
+    ) +
+    theme(
+      legend.position = "none",
+      plot.title.position = "plot"
+    )
+  ggsave(device = "png", filename = here("manuscript/figures/global-thirst.png"),
+         width = 174, units = "mm")
+}

### 2020-06-08 R/create_iio.R

@@ -1,0 +1,10 @@
+##' @title
+##' @param data dataframe
+##' @return
+##' @author Aaron Conway
+##' @importFrom mokken check.iio
+
+##' @export
+create_iio <- function(data) {
+  summary(check.iio(as.matrix(data), item.selection = FALSE))
+}

### 2020-06-08 R/create_iioptds5_table.R

@@ -1,0 +1,17 @@
+##' @title
+##' @return
+##' @author Aaron Conway
+##' @export
+#' @importFrom dplyr as_tibble select everything mutate
+
+create_iioptds5_table <- function(iio_ptds5) {
+  iio_ptds5$item.summary %>%
+  as_tibble() %>%
+    mutate(item = c("I want to drink water",
+                    "My mouth is dry",
+                    "My lips are dry",
+                    "My saliva is thick",
+                    "My tongue is thick")) %>%
+    select(item, everything())
+
+}

### 2020-06-08 R/create_iioptds6_table.R

@@ -1,0 +1,18 @@
+##' @title
+##' @return
+##' @author Aaron Conway
+##' @export
+#' @importFrom dplyr as_tibble select everything mutate
+
+create_iioptds6_table <- function(iio_ptds6) {
+  iio_ptds6$item.summary %>%
+    as_tibble() %>%
+    mutate(item = c("I want to drink water",
+                    "My mouth is dry",
+                    "My lips are dry",
+                    "My throat is dry",
+                    "My saliva is thick",
+                    "My tongue is thick")) %>%
+    select(item, everything())
+
+}

### 2020-06-08 R/create_item_hierarchy_df.R

@@ -1,0 +1,44 @@
+##' @title
+##' @return
+##' @author Aaron Conway
+##' @export
+
+#' @importFrom dplyr summarise across everything mutate
+#' tribble arrange case_when
+#' @importFrom stringr str_remove_all
+#' @importFrom tidyr pivot_longer
+create_item_hierarchy_df <- function(data, Hvalues) {
+  Hi <- Hvalues$Hi
+  scalability <- data %>%
+    summarise(across(everything(), mean)) %>%
+    pivot_longer(
+      cols = everything(),
+      names_to = "item", values_to = "mean"
+    ) %>%
+    mutate(mean = round(mean, 2))
+
+  Hi_df <- tribble(
+    ~Hi, ~se,
+    Hi[1], Hi[6],
+    Hi[2], Hi[7],
+    Hi[3], Hi[8],
+    Hi[4], Hi[9],
+    Hi[5], Hi[10]
+  )
+
+  scalability <- scalability %>% cbind(Hi_df)
+
+  scalability <- scalability %>%
+    mutate(se = str_remove_all(se, "\\(")) %>%
+    mutate(se = str_remove_all(se, "\\)")) %>%
+    mutate(se = as.numeric(se)) %>%
+    mutate(Hi = round(as.numeric(Hi), 2)) %>%
+    arrange(-mean)
+
+  scalability  %>%
+    mutate(item = case_when(item == "water" ~ "I want to drink water",
+                            item == "dry_mouth" ~ "My mouth is dry",
+                            item == "dry_lips" ~ "My lips are dry",
+                            item == "thick_saliva" ~ "My saliva is thick",
+                            item == "thick_tongue" ~ "My tongue is thick"))
+}

### 2020-06-08 R/create_monotonicity_df.R

@@ -1,0 +1,8 @@
+##' @title
+##' @param data dataframe
+##' @importFrom mokken check.monotonicity
+##' @author Aaron Conway
+##' @export
+create_monotonicity_df <- function(data) {
+  summary(check.monotonicity(as.matrix(data)))
+}

### 2020-06-08 R/create_monotonicity_df_ptds6_table.R

@@ -1,0 +1,20 @@
+##' @title
+
+##' @return
+##' @author Aaron Conway
+##' @importFrom dplyr as_tibble select everything mutate
+
+##' @export
+create_monotonicity_df_ptds6_table <- function(monotonicity_df_ptds6) {
+  monotonicity_df_ptds6 %>%
+    as_tibble() %>%
+    mutate(item = c(
+      "My mouth is dry",
+      "My lips are dry",
+      "My tongue is thick",
+      "My saliva is thick",
+      "My throat is dry",
+      "I want to drink water"
+    )) %>%
+    select(item, everything())
+}

### 2020-06-08 R/create_reliability.R

@@ -1,0 +1,10 @@
+
+##' @title
+#' @importFrom mokken check.reliability
+
+##' @return
+##' @author Aaron Conway
+##' @export
+create_reliability <- function(data) {
+  round(check.reliability(as.matrix(data))$MS, 2)
+}

### 2020-06-08 R/create_summary_table.R

@@ -1,0 +1,20 @@
+##' @title
+##' @return
+##' @author Aaron Conway
+##' @export
+#' @importFrom dplyr select
+#' @importFrom gtsummary tbl_summary italicize_labels as_flextable
+create_summary_table <- function(data_ptds) {
+  data_ptds %>%
+    select(age, sex_factor, procedure_factor, fluids_duration, food_duration) %>%
+    tbl_summary(
+      missing = "no",
+      label = list(
+        sex_factor ~ "Gender",
+        procedure_factor ~ "Procedure",
+        fluids_duration ~ "Time since last clear fluids (hours)",
+        food_duration ~ "Time since last food (hours)"
+      )
+    ) %>%
+    italicize_labels()
+}

### 2020-06-08 R/create_thirst_intensity_plot.R

@@ -1,0 +1,28 @@
+
+##' @title
+##' @return
+##' @author Aaron Conway
+##' @export
+#' @importFrom ggbeeswarm geom_beeswarm
+#' @importFrom here here
+#' @importFrom ggplot2 ggplot aes geom_smooth theme_minimal labs theme
+#' ggsave
+create_thirst_intensity_plot <- function(data) {
+  data %>%
+    # mutate(thirst_intensity = as_factor(thirst_intensity)) %>%
+    ggplot(aes(x = ptds, y = factor(thirst_intensity),
+               colour = factor(thirst_intensity))) +
+    geom_beeswarm(alpha = 0.5, size = 2, groupOnX = FALSE) +
+    theme_minimal() +
+    labs(
+      y = "\nThirst intensity (0-10; higher scores indicate greater thirst intensity)",
+      x = "PTDS-5 (5-15; higher scores indicate worse thirst discomfort)\n"
+    ) +
+    theme(
+      legend.position = "none",
+      plot.title.position = "plot"
+    )
+
+  ggsave(device = "png", filename = here("manuscript/figures/thirst-intensity.png"),
+         width = 174, units = "mm")
+}

### 2020-06-08 R/functions.r

@@ -1,275 +1,0 @@
-#' @param data dataframe from csv downloaded from redcap
-
-#' @export
-label_data <- function(data){
-
-#Setting Units
-
-
-#Setting Factors(will create new variable for factors)
-data$redcap_event_name_factor = factor(data$redcap_event_name,levels=c("t3_arm_1","t1_arm_2","t2_arm_2","t3_arm_2","screening_arm_3"))
-data$datetime_v1_complete_factor = factor(data$datetime_v1_complete,levels=c("0","1","2"))
-data$sex_factor = factor(data$sex,levels=c("1","2","3"))
-data$procedure_factor = factor(data$procedure,levels=c("1","2","3","4","5","6","7","8","9","10","11","12","13"))
-data$instructions_factor = factor(data$instructions,levels=c("1","2","3","4","5","6"))
-data$baseline_v1_complete_factor = factor(data$baseline_v1_complete,levels=c("0","1","2"))
-data$ptds_dry_mouth_factor = factor(data$ptds_dry_mouth,levels=c("0","1","2"))
-data$ptds_dry_lips_factor = factor(data$ptds_dry_lips,levels=c("0","1","2"))
-data$ptds_thick_tongue_factor = factor(data$ptds_thick_tongue,levels=c("0","1","2"))
-data$ptds_thick_saliva_factor = factor(data$ptds_thick_saliva,levels=c("0","1","2"))
-data$ptds_dry_throat_factor = factor(data$ptds_dry_throat,levels=c("0","1","2"))
-data$ptds_bad_taste_factor = factor(data$ptds_bad_taste,levels=c("0","1","2"))
-data$ptds_water_factor = factor(data$ptds_water,levels=c("0","1","2"))
-data$global_thirst_factor = factor(data$global_thirst,levels=c("1","2","3","4","5","6"))
-data$thirst_intensity_factor = factor(data$thirst_intensity,levels=c("0","1","2","3","4","5","6","7","8","9","10"))
-data$pain_factor = factor(data$pain,levels=c("0","1","2","3","4","5","6","7","8","9","10"))
-data$o2_therapy_factor = factor(data$o2_therapy,levels=c("1","0"))
-data$perioperative_thirst_discomfort_scale_v1_complete_factor = factor(data$perioperative_thirst_discomfort_scale_v1_complete,levels=c("0","1","2"))
-data$thirst_change_factor = factor(data$thirst_change,levels=c("1","2","3","4","5"))
[…]

### 2020-06-08 R/label_data.R

@@ -1,0 +1,113 @@
+##' @title Label data
+##' @param data dataframe from csv downloaded from redcap
+##' @return
+##' @author Aaron Conway
+##' @export
+
+label_data <- function(data) {
+
+  # Setting Units
+
+
+  # Setting Factors(will create new variable for factors)
+  data$redcap_event_name_factor <- factor(data$redcap_event_name, levels = c("t3_arm_1", "t1_arm_2", "t2_arm_2", "t3_arm_2", "screening_arm_3"))
+  data$datetime_v1_complete_factor <- factor(data$datetime_v1_complete, levels = c("0", "1", "2"))
+  data$sex_factor <- factor(data$sex, levels = c("1", "2", "3"))
+  data$procedure_factor <- factor(data$procedure, levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13"))
+  data$instructions_factor <- factor(data$instructions, levels = c("1", "2", "3", "4", "5", "6"))
+  data$baseline_v1_complete_factor <- factor(data$baseline_v1_complete, levels = c("0", "1", "2"))
+  data$ptds_dry_mouth_factor <- factor(data$ptds_dry_mouth, levels = c("0", "1", "2"))
+  data$ptds_dry_lips_factor <- factor(data$ptds_dry_lips, levels = c("0", "1", "2"))
+  data$ptds_thick_tongue_factor <- factor(data$ptds_thick_tongue, levels = c("0", "1", "2"))
+  data$ptds_thick_saliva_factor <- factor(data$ptds_thick_saliva, levels = c("0", "1", "2"))
+  data$ptds_dry_throat_factor <- factor(data$ptds_dry_throat, levels = c("0", "1", "2"))
+  data$ptds_bad_taste_factor <- factor(data$ptds_bad_taste, levels = c("0", "1", "2"))
+  data$ptds_water_factor <- factor(data$ptds_water, levels = c("0", "1", "2"))
+  data$global_thirst_factor <- factor(data$global_thirst, levels = c("1", "2", "3", "4", "5", "6"))
+  data$thirst_intensity_factor <- factor(data$thirst_intensity, levels = c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10"))
+  data$pain_factor <- factor(data$pain, levels = c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10"))
+  data$o2_therapy_factor <- factor(data$o2_therapy, levels = c("1", "0"))
[…]

### 2020-06-08 R/make_corr_plot.R

@@ -1,0 +1,12 @@
+
+##' @title Make a correlation plot
+##' @param ptds7 data frame
+##' @return
+##' @author Aaron Conway
+##' @importFrom ggcorrplot ggcorrplot
+##' @export
+make_corr_plot <- function(ptds7) {
+  ggcorrplot(cor(ptds7),
+    colors = c("#E46726", "white", "#6D9EC1")
+  )
+}

### 2020-06-08 R/mokken_variables.R

@@ -1,0 +1,16 @@
+##' @title
+##' @return
+##' @export
+#' @importFrom dplyr select starts_with ends_with rename_all mutate_if
+#' @importFrom stringr str_remove_all
+mokken_variables <- function(participation_df) {
+
+  # variables for mokken scaling
+  participation_df %>%
+    select(starts_with("ptds")) %>%
+    select(ends_with("factor")) %>%
+    rename_all(.funs = ~ str_remove_all(., "ptds_")) %>%
+    rename_all(.funs = ~ str_remove_all(., "_factor")) %>%
+    mutate_if(is.factor, ~ as.numeric(.x)) %>%
+    as.data.frame()
+}

### 2020-06-08 R/moscales_lowerbounds.R

@@ -1,0 +1,21 @@
+##' @title Bounds
+##' @param x data frame
+##' @return
+##' @export
+#' @importFrom mokken aisp
+moscales_lowerbounds <- function(x, lowerbounds = seq(from = 0.05, to = 0.80, by = 0.05)) {
+  ret.value <- NULL
+  for (lowerbound in lowerbounds)
+  {
+    tmp <- aisp(x, lowerbound = lowerbound)
+    if (is.null(ret.value)) {
+      ret.value <- data.frame("Item" = rownames(tmp), "Scales." = tmp[, 1])
+    }
+    else {
+      ret.value <- cbind(ret.value, "Scales." = tmp[, 1])
+    }
+    names(ret.value)[ncol(ret.value)] <- sprintf("%.2f", lowerbound)
+  }
+  rownames(ret.value) <- NULL
+  ret.value
+}

### 2020-06-08 R/plan.R

@@ -1,65 +1,54 @@
-library(drake)
-library(tidyverse)
-library(here)
-library(mokken)
-library(gtsummary)
+#' @title drake plan
+#' @rdname get_analysis_plan
+#' @description Targets and functions for analyses
+#' @export
+#' @importFrom readr read_csv
+#' @importFrom dplyr select
+#' @importFrom here here
+#' @importFrom drake drake_plan knitr_in
+#'
+get_analysis_plan <- function() {
+  drake_plan(
+    original_data = read_csv(here("data/ThirstDiscomfortScal_DATA_2020-03-24_2016.csv")),
+    # Label original data
+    data_labelled = label_data(data = original_data),
+    participation_df = preprocess_df(data_labelled),
+    ptds7 = mokken_variables(participation_df),
+    corr_plot = make_corr_plot(ptds7),
+    aisp = moscales_lowerbounds(x = ptds7),
+    ptds6 = ptds7 %>%
+      select(-bad_taste),
+    H_values_ptds7 = create_H_values(data = ptds7),
+    H_values_ptds7_table = create_H_values_ptds7_table(H_values_ptds7),
+    H_values_ptds6 = create_H_values(data = ptds6),
+    monotonicity_df_ptds6 = create_monotonicity_df(data = ptds6),
+    monotonicity_df_ptds6_table = create_monotonicity_df_ptds6_table(monotonicity_df_ptds6),
+    iio_ptds6 = create_iio(data = ptds6),
+    iio_ptds6_table = create_iioptds6_table(iio_ptds6),
+    ptds5 = ptds6 %>%
+      select(-dry_throat),
+    iio_ptds5 = create_iio(data = ptds5),
+    iio_ptds5_table = create_iioptds5_table(iio_ptds5),
+    accuracy_ptds5 = round(iio_ptds5$HT, 3),
+    reliability_ptds5 = create_reliability(ptds5),
+    H_values_ptds5 = create_H_values(data = ptds5),
+    item_hierarchy_df_ptds5 = create_item_hierarchy_df(
+      data = ptds5,
+      Hvalues = H_values_ptds5
+    ),
+    data_ptds = add_ptds_total(ptds5, participation_df), # data with total ptds score added
+    summary_table = create_summary_table(data_ptds),
+    global_thirst_plot = create_global_thirst_plot(data_ptds),
+    thirst_intensity_plot = create_thirst_intensity_plot(data_ptds),
[…]

### 2020-06-08 R/preprocess_df.R

@@ -1,0 +1,33 @@
+##' @title Preprocessing
+##' @param data_labelled labelled data from redcap
+##' @return
+##' @author Aaron Conway
+##' @export
+#' @importFrom dplyr filter starts_with mutate select
+#' ends_with rename_all mutate_if
+#' @importFrom stringr str_detect str_remove_all
+#' @importFrom tidyr drop_na
+preprocess_df <- function(data_labelled) {
+  rel1 <- data_labelled %>%
+    filter(str_detect(id, "R")) %>% # remove screening data
+    filter(
+      redcap_event_name != "t2_arm_2",
+      redcap_event_name != "t3_arm_2"
+    ) # get participant data for this arm
+  participation_df <- data_labelled %>%
+    filter(str_detect(id, "R") | str_detect(id, "V")) %>% # remove screening data
+    filter(redcap_event_name != "t1_arm_2") %>% # remove additional timepoints for test-retest (only n=5)
+    filter(redcap_event_name != "t2_arm_2") %>%
+    coalesce_join(rel1, by = "id") %>% # join participant data from t1 for reliability arm
+    drop_na(starts_with("ptds")) %>% # remove one participant who missed one item
+    mutate(fluids_duration = round(
+      as.numeric(
+        difftime(date_time, fluids, units = "hours")
+      )
+    ), 1) %>% # time since last fluids
+    mutate(food_duration = round(
+      as.numeric(
+        difftime(date_time, food, units = "hours")
+      )
+    ), 1)
+}

### 2020-06-08 _drake.R

@@ -1,0 +1,9 @@
+## Load functions in R folder
+devtools::load_all()
+
+## _drake.R must end with a call to drake_config().
+## The arguments to drake_config() are basically the same as those to make().
+## lock_envir allows functions that alter the random seed to be used. The biggest
+## culprits of this seem to be interactive graphics e.g. plotly and mapdeck.
+drake::drake_config(get_analysis_plan(),
+                    lock_envir = FALSE)

### 2020-07-02 manuscript/index.Rmd
first draft to onedrive for review

@@
 author:
+  - Aaron Conway PhD:
+      email: aaron.conway@utoronto.ca
+      phone: +1 416 946 7112
+      address: 585 University Ave | Toronto, ON  M5G 2N2
+      institute: [PMCC, LBFON, QUT]
+      correspondence: true
   - Kristina Chang MScN:
@@
       correspondence: false
-  - Aaron Conway PhD:
-      email: aaron.conway@utoronto.ca
-      phone: +1 416 946 7112
-      address: 585 University Ave | Toronto, ON  M5G 2N2
-      institute: [PMCC, LBFON, QUT]
-      correspondence: true
 institute:
@@
 library(flextable)
+
+food  <-  glue::glue("{round(mean(drake::readd(data_ptds)$food_duration, na.rm=T), 1)} (SD {round(sd(drake::readd(data_ptds)$food_duration, na.rm=T), 1)})")
+fluids  <-  glue::glue("{round(mean(drake::readd(data_ptds)$fluids_duration, na.rm=T), 1)} (SD {round(sd(drake::readd(data_ptds)$fluids_duration, na.rm=T), 1)})")
+ptds5 <-  glue::glue("{round(median(drake::readd(data_ptds)$ptds, na.rm=T), 1)} (IQR {quantile(drake::readd(data_ptds)$ptds, 0.25)}, {quantile(drake::readd(data_ptds)$ptds, 0.75)})")
+drake::loadd(global_thirst_corr)
+drake::loadd(intensity_corr)
+
 ```
@@
 
-### Abstract
+## Abstract
 
-**Introduction**
+### Introduction
 
-**Methods**
+Symptoms of thirst discomfort are common in patients who are required to undergo long periods of fasting prior to medical and surgical procedures.
+The aim of this study was to examine the validity and reliability of the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures.
 
-**Results**
+### Methods
 
-**Conclusion**
+A prospective observational study was conducted.
+Patients who were fasted and scheduled for an elective cardiac or interventional radiology procedure were included.
+Mokken scaling analysis was conducted to investigate the unidimensionality and hierarchical nature of the scale.
+Correlations between perioperative thirst discomfort scale scores and fasting durations were calculated to evaluate construct validity and convergent validity was evaluated by comparing scores with global thirst discomfort and thirst intensity items.
 
-Clinical trial number: Not applicable
+### Results
+
+Responses from 193 participants were included.
+The mean duration of fasting from food and non-clear fluids was `r food` hours and clear fluids was `r fluids` hours.
+Five items from the perioperative thirst discomfort scale (PTDS-5) formed a Mokken scale with evidence of invariant item ordering.
+The median  PTDS-5 score was `r ptds5`  from a possible range of 5 to 15.
+The hierarchical pattern of responses to PTDS-5 items is interpretable in terms of respondents most easily endorsing an item related to discomfort associated with the desire to drink water through to first endorsing symptoms related to dryness of the mouth and lips before those related to the abnormal sensations of 'thick' saliva and a 'thick' tongue.
+Scale reliability was adequate with a coefficient score (*rho* =`r drake::readd(reliability_ptds5)`).
+There was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (*tau* = `r round(global_thirst_corr$tau, 2)`; 95% CI = `r round(global_thirst_corr$CI_low, 2)` to `r round(global_thirst_corr$CI_high, 2)`), as well as thirst intensity (*tau* = `r round(intensity_corr$tau, 2)`; 95% CI = `r round(intensity_corr$CI_low, 2)` to `r round(intensity_corr$CI_high, 2)`).
+Duration of fasting was not associated with PTDS-5 scores.
+
+### Conclusion
+
+The items in the PTDS-5 form a strong Mokken scale, meaning it is a reliable and precise way to order patients according to their thirst discomfort.
+Nurses may use this simple scale to assess the severity of thirst discomfort in patients who are fasted prior to procedures and initiate interventions according to individual needs.
 
@@
 Author contributions: 
-KC: Study design, data analysis and wrote manuscript; AC: Study design, analysis, wrote manuscript;  MB: Data collection, revised the manuscript for important intellectual content; DP: Data collection, revised the manuscript for important intellectual content; NK: Study design, revised the manuscript for important intellectual content; AM: Study design, revised the manuscript for important intellectual content; MP: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; PG: Study design, revised the manuscript for important intellectual content; SA: Study design, revised the manuscript for important intellectual content.
+AC: Study design, data analysis and wrote manuscript; KC: Study design, revised the manuscript for important intellectual content;  MB: Data collection, revised the manuscript for important intellectual content; DP: Data collection, revised the manuscript for important intellectual content; NK: Study design, revised the manuscript for important intellectual content; AM: Study design, revised the manuscript for important intellectual content; MP: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; PG: Study design, revised the manuscript for important intellectual content; SA: Study design, revised the manuscript for important intellectual content.
 
@@
 
-# Introduction
+## Introduction
 
-Pre-procedure fasting has been historically implemented to reduce the risk of vomiting and aspiration pneumonia with general anaesthesia, or in case emergency intubation is required due to unexpected cardiac arrest.[@hamid2014pre; @osborne2002preoperative] However, prolonged fluid restriction causes thirst symptoms to develop (e.g., dry mouth, swollen tongue), which can lead to great discomfort.[@madsen1998perioperative] Current guidelines related to pre-procedure fasting for elective procedures recommend a minimum fasting period of 2 hours nil-per-os (NPO) for clear fluids.[@dobsonGuidelinesPracticeAnesthesia2018] Despite these recommendations, current practice is for patients undergoing surgical and other medical procedures that require sedation or anesthesia to receive standardized 'nil-by-mouth' fasting instructions at a pre-specified time interval before procedures.
+Pre-procedure fasting is used to reduce the risk of vomiting and aspiration pneumonia during sedation and general anaesthesia, or in case emergency intubation is required due to unexpected cardiac arrest.[@hamid2014pre; @osborne2002preoperative] However, prolonged fluid restriction causes thirst symptoms to develop (e.g., dry mouth, swollen tongue), which can lead to great discomfort.[@madsen1998perioperative] Current guidelines related to pre-procedure fasting for elective procedures recommend a minimum fasting period of 2 hours nil-per-os (NPO) for clear fluids.[@dobsonGuidelinesPracticeAnesthesia2018] Despite these recommendations, current practice is for patients undergoing surgical and other medical procedures that require sedation or anesthesia to receive standardized 'nil-by-mouth' fasting instructions at a pre-specified time interval before procedures.
 For example, 'no eating or drinking after midnight' is most common.
@@
 As a direct result of prolonged pre-procedure fasting, symptoms of thirst discomfort have been reported as common and severe.
-In a qualitative study where 12 participants were interviewed from a tertiary hospital in Australia, surgical patients and patients who adhered to prolonged fasting instructions (e.g., NPO for two days) for tests, procedures or symptom management (e.g., bowel obstruction) described the discomfort from thirst symptoms to be more unbearable than hunger.[@carey2015qualitative] Similarly, Madsen et al.
+In a qualitative study where 12 participants were interviewed from a tertiary hospital in Australia, surgical patients and patients who adhered to prolonged fasting instructions described the discomfort from thirst symptoms to be the worst physical effect of fasting.[@carey2015qualitative] Similarly, Madsen et al.
 interviewed a convenience sample of 50 adult surgical patients who reported that thirst symptoms caused more discomfort than hunger, sleep, or anxiety related to the procedure.[@madsen1998perioperative] However, thirst and its symptoms continue to be undervalued, under-reported and infrequently assessed by health care providers, including the nursing team.[@milani2016thirst]
@@
 Despite the relevance and value of assessing thirst-discomfort of patients, the subjective experience of thirst presents challenges in developing a valid and reliable tool to succinctly and accurately measure its symptoms and level of discomfort prior to procedures.
-A thirst-discomfort scale for perioperative use was developed at a surgical centre affiliated with an accredited public hospital in Brazil.[@Martins_2017] The perioperative thirst-discomfort tool was developed in three stages, including face and content validation and reliability, and based on the Consensus-based Standards for the selection of health status Measurement Instruments (COSMIN) checklist.[@mokkink2010cosmin] Inter-rater reliability was tested through inter-observer equivalence where a pair of nurses independently, but simultaneously administered the scale among 70 patients.
+A thirst-discomfort scale for perioperative use was developed at a surgical centre affiliated with an accredited public hospital in Brazil.[@Martins_2017] The seven-item perioperative thirst-discomfort scale (PTDS-7) was developed in three stages, including face and content validation and reliability, and based on the Consensus-based Standards for the selection of health status Measurement Instruments (COSMIN) checklist.[@mokkink2010cosmin] Inter-rater reliability was tested through inter-observer equivalence where a pair of nurses independently, but simultaneously administered the scale among 70 patients.
 Six items on the scale had a weighted kappa coefficient of 1, while the item "I feel like drinking water" was 0.97.[@Martins_2017] Cronbach's alpha was used to test the internal consistency of the scale with a value of 0.91.[@Martins_2017]
 
-Although the Perioperative Thirst Discomfort Scale was developed in a different context (e.g., surgical use), it demonstrates strong potential for accurately assessing thirst-discomfort in patients before medical procedures more generally.
-For this reason, the aim of this study was to test the validity and reliability of the Perioperative Thirst Discomfort Scale for measuring thirst discomfort prior to procedures.
+Although the initial validation of the PTDS-7 was conducted in a different context (i.e.
+post-surgical patients), it demonstrates strong potential for accurately assessing thirst-discomfort more generally for peri-operative use.
+The aim of this study was to test the validity and reliability of the PTDS-7 for measuring thirst discomfort for patients who are fasting prior to medical procedures.
 
-# Methods
+## Methods
 
-## Study Design
+### Study Design
 
@@
 Patients with afternoon procedures can have a small breakfast but must remain NPO from 6:00am on the day of the procedure.
-Ethics approved by....
+Ethical approval for this study (Ethical Committee N° 19-5585) was provided by the University Health Network Research Ethics Board, Toronto, Canada on October 31 2019.
 
-## Participants
+### Participants
 
-### Inclusion criteria
+#### Inclusion criteria
 
-Adult patients undergoing an elective procedure in the Cardiac Catheterization Laboratories or in Interventional Radiology at the Toronto General Hospital were included.
+Adult patients who were fasted and scheduled for an elective procedure in the Cardiac Catheterization Laboratories or in Interventional Radiology at an academic teaching hospital in Canada were included.
 
-### Exclusion criteria
+#### Exclusion criteria
 
@@
 
-## Data collection
+### Data collection
 
@@
 
-*Seven-item Perioperative Thirst Discomfort Scale (PTDS-7)*
+#### Measures
 
-The PTDS-7 is a 7-item self-reported, composite score evaluating the severity of dry mouth, lips, and throat, thick tongue and saliva, a bad taste in the mouth, and a desire to drink water.
+The seven-item Perioperative Thirst Discomfort Scale (PTDS-7) is a 7-item self-reported, composite score evaluating the severity of dry mouth, lips, and throat, thick tongue and saliva, a bad taste in the mouth, and a desire to drink water.
 The total score can range from 0 to 14, with 14 representing the "most extreme" intensity of discomfort related to perioperative thirst.
@@
 The Cronbach alpha score was 0.91.
-The time taken to complete the Perioperative Thirst Discomfort Scale was not evaluated.
+The time taken to complete the PTDS-7 was not evaluated.
 
-*Global thirst discomfort score*
-
-Participants were asked to rate their overall level of thirst discomfort on a rating scale with score ranging from "extreme thirst discomfort" to "extreme comfort".
-
-*Thirst intensity rating*
-
-Participants will be asked to rate the intensity of their thirst on a scale ranging from 0 to 10, where 0 represents "No thirst" and 10 represents "Most intense thirst".
-
-*Pain scale rating*
-
+Participants used a global thirst discomfort score to rate their overall level of thirst discomfort on a rating scale with scores ranging from "extreme thirst discomfort" to "extreme comfort".
+Intensity of their thirst was rated on a scale ranging from 0 to 10, where 0 represents "No thirst" and 10 represents "Most intense thirst".
 The experience of pain may influence the participant's perception of thirst.[@Pierotti_2018] Participants were asked to rate their current level of pain on a scale of 0 to 10, where 0 represents "no pain" and 10 represents that their pain as "unimaginable/unspeakable".
-
-*Use of oxygen therapy*
-
 The use of oxygen therapy (e.g., via nasal prongs) may contribute to the experience of thirst symptoms (e.g., dry mouth, dry throat).[@conchon2015perioperative].
@@
 
-## Statistical analysis
+### Statistical analysis
 
-Mokken scaling analyses properties of individual items as described by the item characteristic curve.
+Mokken scaling is a method for establishing if items in a scale conform to a cumulative, hierarchical structure.[@perng2012construct] An intentionally simple example of a hierarchical scale is a set of items enquiring about height that increases in increments of 10cm from 100cm to 200cm.
+In this example, taller people would endorse more items than those who were shorter.[@watson2008hierarchy]
+
+To determine if items conform to a hierarchical structure, Mokken scaling analyses properties of individual items as described by the item characteristic curve.
 Item characteristic curves show the association between the score on an item to the level of the latent trait being measured.
-Mokken scaling makes no assumptions about this association, apart from the assumption of monotone homogeneity.
-This means that when the trait increases, so does the item score.
+Mokken scaling makes only the assumption of monotone homogeneity about the association between the score on an item and the level of the latent trait.
+This means that as the trait increases, so does the item score.
 "Difficulty", is a term used in Mokken Scale Analysis to refer to the extent to which items are endorsed by respondents.
-Items at the upper end of the range of the latent trait being more difficult.
+Items at the upper end of the range of the latent trait are characterized as being more difficult.
 Items in a Mokken Scale are arranged along the latent trait in terms of their difficulty.
 The properties of items can be measured using the scalability coefficient, *H* (Loevinger's coefficient).
-*H* measures the extent to which items are arranged as expeceted by their mean values along the latent trait.
+*H* measures the extent to which items are arranged as expected by their mean values along the latent trait.
 *H* > 0.3 is the minimum acceptable value, with values > 0.4 indicating a moderate scale and > 0.5 indicates a strong scale.
-Invariant item ordering, where where the order of items along the latent trait is the same for all respondents at all levels of the latent trait, is a desirable property of a Mokken Scale.
+
+Invariant item ordering, where the order of items along the latent trait is the same for all respondents at all levels of the latent trait, is a desirable property of a Mokken Scale.
 It can be assessed mathematically to look for significant violations of this property.
-The accuracy of the  invariant item ordering can also be assessed by calculating the H~TRANS~ (H~T~) coefficient.
-Values of H~T~ exceeding 0.3 indicates acceptable accuracy of invariant item ordering.
+The accuracy of the  invariant item ordering can also be assessed by calculating the *H~TRANS~* (**H~T~**) coefficient.
+Values of **H~T~** exceeding 0.3 indicates acceptable accuracy of invariant item ordering.
 
@@
 Mokken scale analysis proceeded by first checking the PTDS-7 scalability coefficients.
-If H~i~ was below 0.3 or if the lower limit of the 95% CI (confidence interval) for H~i~ was below 0.3, the item was excluded.
-Then, scale partitioning was carried out to explore the dimensions of the PTDS through increasing c (Lower bound c defines the minimum value of coefficients Hi in the Mokken scale by 0.05 increments).[@molenaar2000mps5] Monotone homogeneity and invariant item ordering were investigated at the whole scale level, as no sub-scales were identified.
-
-To assess invariant item ordering, crit values <40 were considered acceptable.[@molenaar2000mps5]
-
-Accuracy - Values below 0.3 are unacceptable and above 0.4 indicates moderate invariant item ordering.[@Ligtvoet_2010]
+If **H~i~** was below 0.3 or if the lower limit of the 95% CI (confidence interval) for *H~i~* was below 0.3, the item was to be excluded.
+Then, scale partitioning was carried out to explore the dimensions of the PTDS through increasing c (Lower bound c defines the minimum value of coefficients **H~i~** in the Mokken scale by 0.05 increments).[@molenaar2000mps5] Monotone homogeneity and invariant item ordering were investigated at the whole scale level, as no sub-scales were identified.
+To assess for violations of invariant item ordering, crit values <40 were considered acceptable.[@molenaar2000mps5] Provided invariant item ordering holds for a set of items, the coefficient **H~T~** expresses the accuracy of the ordering, with values below 0.3 being unacceptable and above 0.4 indicating moderate invariant item ordering.[@Ligtvoet_2010] The coefficient score (rho), which is similar to Cronbach's alpha, was used to assess reliability of the scale.
 
 Construct validity was evaluated by using correlations to identify associations between scores on the Perioperative Thirst-Discomfort Scale and fasting duration.
-To demonstrate construct validity we hypothesised that higher scores on the Perioperative Thirst Discomfort Scale would be associated with greater fasting duration.
-Convergent validity was assessed by using Kendall's tau to compare the Perioperative Thirst Discomfort Scale score with the global thirst discomfort score.
-Convergent validity was assessed by using the Spearman's rank correlation to compare the Perioperative Thirst Discomfort Scale with the thirst intensity rating and pain scale rating.
-Chi-square analysis was used to evaluate the association between Perioperative Thirst Discomfort Scale score and use of oxygen supplementation.
+To demonstrate construct validity we hypothesized that higher scores on the PTDS would be associated with greater fasting duration.
+Convergent validity was assessed by using Kendall's *tau* to compare the PTDS score with the global thirst discomfort score.
+Convergent validity was assessed by using the Spearman's rank correlation to compare the PTDS with the thirst intensity rating and pain scale rating.
 
-# Results
-
-## Participants
+## Results
 
@@
 A biopsy was the most common Interventional Radiology procedure (n=62; 32%), but a wide range of other procedures were included.
+The mean duration of fasting from food and non-clear fluids was `r food` hours and clear fluids was `r fluids`.
 
-## Mokken scale analysis
+### Mokken scale analysis
 
-No items required removal from the PTDS-7 based on the condition that scalability coefficients (H~i~ and lower bound of 95% CI) should be over 0.3 (Appendix Table \@ref(tab:Hvaluesptds7)).
+No items required removal from the PTDS-7 based on the condition that scalability coefficients (*H~i~* and lower bound of 95% CI) should be over 0.3 (Appendix Table \@ref(tab:Hvaluesptds7)).
 We used a lower bound for *c*, starting from 0.05 and increasing to 0.80 in 0.05 increments, to explore the dimensionality of the PTDS.
@@
 The 6 remaining items (PTDS-6) were further examined for the criteria of a Mokken scale and item invariant ordering.
-The PTDS-6 item set has a homogeneity value H(se) of `r readd(H_values_ptds6)$H`, which indicated a strong Mokken scale.
+The PTDS-6 item set has a homogeneity value *H(se)* of `r readd(H_values_ptds6)$H`, which indicated a strong Mokken scale.
 Results indicated that there were no significant violations of the assumption of monotonicity (z-sig).
@@
 
-### Invariant item ordering
+#### Invariant item ordering
 
@@
 
-### Accuracy and reliability
+#### Accuracy and reliability
 
-The accuracy of the item ordering was found to be acceptable with a H~T~ coefficient of `r drake::readd(accuracy_ptds5)`.
+The accuracy of the item ordering was found to be acceptable with a *H~T~* coefficient of `r drake::readd(accuracy_ptds5)`.
 Scale reliability was adequate with a coefficient score (Rho) of `r drake::readd(reliability_ptds5)`.
 
-### Item difficulty hierarchy
+#### Item difficulty hierarchy
 
+The item difficult hierarchy for the PTDS-5 is presented in Table \@ref(tab:tab2).
 The most highly endorsed symptom (i.e.
-the least severe symptom of thirst discomfort in the PTDS-5) was being bothered by a desire to drink water (Table \@ref(tab:tab2)).
+the least severe symptom of thirst discomfort in the PTDS-5) was being bothered by a desire to drink water.
 The least endorsed symptom (i.e.
 the most severe symptom of thirst discomfort in the PTDS-5) was being bothered by the perception of a 'thick tongue'.
-The items in the PTDS-5 form a strong Mokken scale (H~s~ = `r drake::readd(H_values_ptds5)$H[1]`), meaning it is a reliable and precise way to order patients according to their thirst discomfort.
+The items in the PTDS-5 form a strong Mokken scale (H~s~ = `r drake::readd(H_values_ptds5)$H[1]`).
 
-## Construct validity
+### Construct and convergent validity
 
-PTDS-5 and Age
+The median PTDS-5 score was `r ptds5`.
+The distribution of PTDS-5 scores was right-skewed (Figure \@ref(fig:distribution)).
+There was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (Figure \@ref(fig:globalthirst)), as well as thirst intensity (Figure \@ref(fig:intensity)).
+In contrast, a longer duration of fasting was not linearly associated with PTDS-5 scores.
+The associations between PTDS-5 scores with food and fluid duration is displayed in Figure \@ref(fig:fasting).
+Likewise, the correlation between pain and PTDS-5 scores was not significant (Spearman's rho = `r round(drake::readd(pain_corr)$rho, 2)`; 95% CI = `r round(drake::readd(pain_corr)$CI_low, 2)` to `r round(drake::readd(pain_corr)$CI_high, 2)`).
 
-PTDS-5 and Food/fluid duration
+## Discussion
 
-PTDS-5 and global thirst
+Result indicate that the PTDS-5 is a reliable and precise way to order patients according to their thirst discomfort.
+In addition, an acceptable accuracy for invariant item ordering was identified, meaning patients are likely to rank the PTDS-5 symptoms  of thirst discomfort in the same way.
+As such, the item difficulty hierarchy, presented in Table \@ref(tab:tab2) can be used as a guide to easily determine which patients are expreriencing worse thirst discomfort.
+The hierarchical pattern of responses to items in the Mokken scale is interpretable in terms of respondents more easily endorsing items related to a general desire to drink water through to first endorsing more specific symptoms related to dryness of the mouth and lips prior to endorsing discomfort associated with the sensations of 'thick' saliva and 'thick' tongue.
+In more practical terms, a patient who reports feeling bothered by a sensation that saliva is thick, could be interpreted as experiencing worse thirst discomfort than a patient who did not endorse any of the items lower down on the item difficulty hierarchy (i.e.
+'my lips are dry' and 'my mouth is dry').
 
-PTDS-5 and thirst intensity
+The original (7-item) perioperative thirst-discomfort scale has previously been used to determine the relationship between thirst intensity and thirst-discomfort in the immediate postoperative period during anesthesia recovery.[@Pierotti_2018]  In a non-probabilistic sample of 203 adult participants in post-operative recovery following elective, urgent and emergency surgeries, the investigators found no association between fasting duration, or patient age, and the intensity of thirst and degree of discomfort.[@Pierotti_2018] However, the mean fasting duration among participants was uniformly high (mean 16.2 [SD 8.7] hours), which may have prevented the researchers from discerning a correlation between fasting duration.
+Even in the subgroup of participant who reported 8 hours of fasting, all reported thirst.[@aroni2012assessment; @Pierotti_2018]
 
-PTDS-5 and pain
+It should be noted that an investigation of the dimensionality of the 7-item PTDS had not been performed prior to our study.
+A uni-dimensional solution was ientified in our analysis.
+This should be expected considering the small number of items that were evaluated.
 
-# Discussion
+We did not observe an association between PTDS-5 and fasting duration for either food or fluids.
+This would suggest that it is important to assess thirst discomfort periodically, regardless of the duration of fasting.
+If thirst discomfort is present, and fasting is still indicated, interventions to alleviate symptoms should be implemented.
+There is a lack of evidence regarding the efficacy of any such interventions.
 
-The items in the PTDS-5 form a strong Mokken scale (H~s~ = `r drake::readd(H_values_ptds5)$H[1]`), meaning it is a reliable and precise way to order patients according to their thirst discomfort.
-In addition, an acceptable accuracy for invariant item ordering was identified, meaning patients are likely to rank the PTDS-5 symptoms  of thirst discomfort in the same way.
-Discuss clinical usefulness of the scale...
-Comment on the results for the item difficulty hierarchy (least severe symptom to most severe symptom).
+### Limitations
 
-Contrast with results of previous study...The perioperative thirst-discomfort tool has previously been used to determine the relationship between thirst intensity and thirst-discomfort in the immediate postoperative period during anesthesia recovery.[@Pierotti_2018]  In a non-probabilistic sample of 203 adult participants awaiting elective, urgent and emergency surgeries, the investigators found no association between the fasting time, or patient age, and the intensity of thirst and degree of discomfort.[@Pierotti_2018] However, the mean fasting time among participants was quite high at 16.2 hours (SD 8.7 hours), which may have prevented the researchers from discerning correlation between fasting time and thirst as even with 8 hours of fasting all participants reported thirst.[@aroni2012assessment; @Pierotti_2018] There was no correlation between age, and intensity of thirst and thirst-discomfort, but few elderly patients were recruited to the study with a mean age of 41.5 years old (SD 15.8 years).
+Selection bias should be considered because a convenience sample was used for this study.
+Also, it is uncertain if similar results would be obtained for patients who are fasted prior to undergoing procedures in other procedural settings.
+Only a very small number of participants had to be excluded from analyses due to missing data, so it is unlikely that attrition bias would have exerted a major influence on the results.
+Further research is required to evaluate the minimal detectable difference for the PTDS-5 as well as the minimal clinically important difference prior to use of this scale as an outcome measure in research.
 
-- The dimensionality of the PTDS had not been performed prior to our study.
-A multi-dimensional solution was not apparent, although this was to be expected with the small number of items that were evaluated.
+### Conclusion
 
-- Lack of association between PTDS-5 and fasting duration.
-As such it is important to assess thirst discomfort periodically, regardless of the duration of fasting.
-If thirst discomfort is present, and fasting is still indicated, interventions to alleviate symptoms should be implemented.
-It should be noted, though, that there is a lack of evidence regarding the efficacy of any such interventions...
-
-## Limitations
-
-- convenience sample (selection bias can not be ruled out)
-- generalize to other procedural settings?
-- Low amount of missing data 
-- Further research required to evaluate the minimal detectable difference in the PTDS-5 and the minimal clinically important difference...
-
-## Conclusion
+The items in the PTDS-5 form a reliable and precise way to order patients according to their thirst discomfort.
+Nurses may use this scale to assess the severity of thirst discomfort in patients who are fasted prior to procedures and initiate interventions according to individual needs.
 
@@
 
-# References
+## References
 
@@
 
-# Figure legend
+## Figure legend
+
+Fig. \@ref(fig:distribution) Distribution of PTDS-5 scores
 
@@
 
+```{r distribution, fig.cap = "Distribution of PTDS-5 scores"}
+knitr::include_graphics("figures/distribution.png")
+```
+
+\pagebreak
+
 ```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 5 to 15 with higher scores indicating worse thirst discomfort."}
@@
 
-# Appendix
+## Appendix
 

### 2020-07-02 R/create_distribution_plot.r

@@ -1,0 +1,23 @@
+#' @title
+#' @return
+#' @export
+#' @importFrom ggplot2 ggplot aes theme_minimal scale_x_continuous
+#' labs element_blank theme ggsave geom_histogram
+create_distribution_plot <- function(data_ptds) {
+    data_ptds %>%
+        ggplot(aes(x = ptds)) +
+        geom_histogram(binwidth = 0.5) +
+        theme_minimal() +
+        labs(
+            y = "Count",
+            x = "\nPTDS-5 (range is 5 to 15; higher scores indicate worse thirst discomfort)"
+        ) +
+        scale_x_continuous(
+            breaks = c(seq(5, 15, by = 1))
+        )
+
+    ggsave(
+        device = "png", filename = here("manuscript/figures/distribution.png"),
+        width = 174, units = "mm"
+    )
+}

### 2020-07-02 R/create_fasting_duration_ptds_plot.R

@@ -8,17 +8,21 @@
 #' @importFrom ggtext element_markdown geom_richtext
 #' @importFrom dplyr summarize mutate
 #' @importFrom glue glue
-create_fasting_duration_ptds_plot <- function(data) {
+#' @importFrom tibble tibble
+create_fasting_duration_ptds_plot <- function(data = data_ptds,
+  fluids_corr, food_corr) {
 
-  ptds_fluids_cor <- data %>%
-    drop_na(ptds, fluids_duration) %>%
-    summarize(r_square = cor(ptds, fluids_duration)^2) %>%
-    mutate(
+rho <- round(fluids_corr$rho, 2)
+rho_low <- round(fluids_corr$CI_low, 2)
+rho_high <- round(fluids_corr$CI_high, 2)
+
+  ptds_fluids_cor <- tibble(
       # location of each text label in data coordinates
-      ptds = 14,
+      ptds = 15,
       fluids_duration = 20,
-      # text label containing r^2 value
-      label = glue("*r*<sup>2</sup> = {round(r_square, 2)}")
+      # text label containing rho value
+      label = glue("*rho* = {rho}
+      (95% CI = {rho_low} to {rho_high})")
     )
 
   fluids <- data %>%
@@ -29,7 +33,7 @@ create_fasting_duration_ptds_plot <- function(data) {
     theme_minimal() +
     labs(
       x = "\nDuration of fasting from <span style='color:dodgerblue'>fluids</span> in hours",
-      y = "PTDS-5 (5-15; higher scores indicate worse thirst discomfort)\n"
+      y = "PTDS-5\n"
     ) +
     theme(
       legend.position = "none",
@@ -40,7 +44,7 @@ create_fasting_duration_ptds_plot <- function(data) {
     ggtext::geom_richtext(
       data = ptds_fluids_cor,
       aes(label = label),
-      hjust = 1, vjust = 1
+      hjust = 1, vjust = 2
     )
 
   ptds_food_cor <- data %>%
@@ -54,6 +58,19 @@ create_fasting_duration_ptds_plot <- function(data) {
       label = glue("*r*<sup>2</sup> = {round(r_square, 2)}")
     )
 
+    rho <- round(food_corr$rho, 2)
+rho_low <- round(food_corr$CI_low, 2)
+rho_high <- round(food_corr$CI_high, 2)
+
+  ptds_food_cor <- tibble(
+      # location of each text label in data coordinates
+      ptds = 15,
[…]

### 2020-07-02 R/create_fluids_corr.r

@@ -1,0 +1,12 @@
+#' @title Correlation ptds and fluids fasting
+#' @return
+#' @author Aaron Conway
+#' @export
+#' @importFrom correlation correlation
+#' @importFrom dplyr mutate select
+
+create_fluids_corr <- function(data_ptds) {
+  data_ptds %>%
+    select(ptds, fluids_duration) %>%
+    correlation(method = "spearman")
+}

### 2020-07-02 R/create_food_corr.r

@@ -1,0 +1,12 @@
+#' @title Correlation ptds and food fasting
+#' @return
+#' @author Aaron Conway
+#' @export
+#' @importFrom correlation correlation
+#' @importFrom dplyr mutate select
+
+create_food_corr <- function(data_ptds) {
+  data_ptds %>%
+    select(ptds, food_duration) %>%
+    correlation(method = "spearman")
+}

### 2020-07-02 R/create_global_thirst_corr.R

@@ -1,0 +1,15 @@
+#' @title Correlation ptds and global thirst
+#' @param data_ptds dataframe
+#' @return
+#' @author Aaron Conway
+#' @export
+#' @importFrom correlation correlation
+#' @importFrom forcats fct_rev
+#' @importFrom dplyr mutate select
+
+create_global_thirst_corr <- function(data_ptds) {
+  data_ptds %>%
+    mutate(global = as.numeric(fct_rev(global_thirst_factor))) %>%
+    select(ptds, global) %>%
+    correlation(method = "kendall")
+}

### 2020-07-02 R/create_global_thirst_plot.R

@@ -1,15 +1,32 @@
-##' @title
-##' @return
-##' @author Aaron Conway
-##' @export
+#' @title
+#' @return
+#' @author Aaron Conway
+#' @export
 #' @importFrom ggbeeswarm geom_beeswarm
 #' @importFrom dplyr desc
 #' @importFrom ggplot2 ggplot aes theme_minimal
 #' labs element_blank theme ggsave
-create_global_thirst_plot <- function(data) {
-  data %>%
-    ggplot(aes(x = ptds, y = reorder(global_thirst_factor, desc(global_thirst_factor)),
-               colour = global_thirst_factor)) +
+#' @importFrom ggtext geom_richtext
+create_global_thirst_plot <- function(data,
+                                      global_thirst_corr) {
+
+tau <- round(global_thirst_corr$tau, 2)
+tau_low <- round(global_thirst_corr$CI_low, 2)
+tau_high <- round(global_thirst_corr$CI_high, 2)
+
+  cor <- tibble(
+    ptds = 12,
+    global_thirst_factor = "Extremely uncomfortable",
+    # text label containing r^2 value
+    label = glue("*tau* = {tau}; (95% CI = {tau_low} to {tau_high})")
+  )
+
+data %>%
+    ggplot(aes(
+      x = ptds, y = reorder(global_thirst_factor, desc(global_thirst_factor)
+      ),
+      colour = global_thirst_factor
+    )) +
     geom_beeswarm(groupOnX = FALSE, alpha = 0.5, size = 2) +
     theme_minimal() +
     labs(
@@ -19,7 +36,15 @@ create_global_thirst_plot <- function(data) {
     theme(
       legend.position = "none",
       plot.title.position = "plot"
+    ) +
+    geom_richtext(
+      data = cor,
+      aes(label = label),
+      hjust = 1, vjust = 2
     )
-  ggsave(device = "png", filename = here("manuscript/figures/global-thirst.png"),
-         width = 174, units = "mm")
+
+  ggsave(
+    device = "png", filename = here("manuscript/figures/global-thirst.png"),
+    width = 174, units = "mm"
+  )
 }

### 2020-07-02 R/create_intensity_corr.R

@@ -1,0 +1,14 @@
+#' @title Correlation ptds and global thirst
+#' @param data_ptds dataframe
+#' @return
+#' @author Aaron Conway
+#' @export
+#' @importFrom correlation correlation
+#' @importFrom forcats fct_rev
+#' @importFrom dplyr mutate select
+
+create_intensity_corr <- function(data_ptds) {
+  data_ptds %>%
+    select(ptds, thirst_intensity) %>%
+    correlation(method = "kendall")
+}

### 2020-07-02 R/create_pain_corr.r

@@ -1,0 +1,12 @@
+#' @title Correlation ptds and pain
+#' @return
+#' @author Aaron Conway
+#' @export
+#' @importFrom correlation correlation
+#' @importFrom dplyr mutate select
+
+create_pain_corr <- function(data_ptds) {
+    data_ptds %>%
+        select(ptds, pain) %>%
+        correlation(method = "spearman")
+}

### 2020-07-02 R/create_thirst_intensity_plot.R

@@ -7,7 +7,23 @@
 #' @importFrom here here
 #' @importFrom ggplot2 ggplot aes geom_smooth theme_minimal labs theme
 #' ggsave
-create_thirst_intensity_plot <- function(data) {
+#' @importFrom ggtext geom_richtext
+
+create_intensity_plot <- function(data,
+intensity_corr) {
+  
+tau <- round(intensity_corr$tau, 2)
+tau_low <- round(intensity_corr$CI_low, 2)
+tau_high <- round(intensity_corr$CI_high, 2)
+
+  cor <- tibble(
+    ptds = 14.5,
+    thirst_intensity = 2,
+    # text label containing r^2 value
+    label = glue("*tau* = {tau}; (95% CI = {tau_low} to {tau_high})")
+  )
+  
+  
   data %>%
     # mutate(thirst_intensity = as_factor(thirst_intensity)) %>%
     ggplot(aes(x = ptds, y = factor(thirst_intensity),
@@ -21,6 +37,11 @@ create_thirst_intensity_plot <- function(data) {
     theme(
       legend.position = "none",
       plot.title.position = "plot"
+    )+
+    geom_richtext(
+      data = cor,
+      aes(label = label),
+      hjust = 1, vjust = 2
     )
 
   ggsave(device = "png", filename = here("manuscript/figures/thirst-intensity.png"),

### 2020-07-02 R/label_data.R

@@ -10,57 +10,176 @@ label_data <- function(data) {
 
 
   # Setting Factors(will create new variable for factors)
-  data$redcap_event_name_factor <- factor(data$redcap_event_name, levels = c("t3_arm_1", "t1_arm_2", "t2_arm_2", "t3_arm_2", "screening_arm_3"))
-  data$datetime_v1_complete_factor <- factor(data$datetime_v1_complete, levels = c("0", "1", "2"))
-  data$sex_factor <- factor(data$sex, levels = c("1", "2", "3"))
-  data$procedure_factor <- factor(data$procedure, levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13"))
-  data$instructions_factor <- factor(data$instructions, levels = c("1", "2", "3", "4", "5", "6"))
-  data$baseline_v1_complete_factor <- factor(data$baseline_v1_complete, levels = c("0", "1", "2"))
-  data$ptds_dry_mouth_factor <- factor(data$ptds_dry_mouth, levels = c("0", "1", "2"))
-  data$ptds_dry_lips_factor <- factor(data$ptds_dry_lips, levels = c("0", "1", "2"))
-  data$ptds_thick_tongue_factor <- factor(data$ptds_thick_tongue, levels = c("0", "1", "2"))
-  data$ptds_thick_saliva_factor <- factor(data$ptds_thick_saliva, levels = c("0", "1", "2"))
-  data$ptds_dry_throat_factor <- factor(data$ptds_dry_throat, levels = c("0", "1", "2"))
-  data$ptds_bad_taste_factor <- factor(data$ptds_bad_taste, levels = c("0", "1", "2"))
-  data$ptds_water_factor <- factor(data$ptds_water, levels = c("0", "1", "2"))
-  data$global_thirst_factor <- factor(data$global_thirst, levels = c("1", "2", "3", "4", "5", "6"))
-  data$thirst_intensity_factor <- factor(data$thirst_intensity, levels = c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10"))
-  data$pain_factor <- factor(data$pain, levels = c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10"))
-  data$o2_therapy_factor <- factor(data$o2_therapy, levels = c("1", "0"))
-  data$perioperative_thirst_discomfort_scale_v1_complete_factor <- factor(data$perioperative_thirst_discomfort_scale_v1_complete, levels = c("0", "1", "2"))
[…]

### 2020-07-02 R/plan.R

@@ -38,11 +38,26 @@ get_analysis_plan <- function() {
     ),
     data_ptds = add_ptds_total(ptds5, participation_df), # data with total ptds score added
     summary_table = create_summary_table(data_ptds),
-    global_thirst_plot = create_global_thirst_plot(data_ptds),
-    thirst_intensity_plot = create_thirst_intensity_plot(data_ptds),
-    fasting_duration_ptds_plot = create_fasting_duration_ptds_plot(data_ptds),
+    global_thirst_plot = create_global_thirst_plot(
+      data = data_ptds,
+      global_thirst_corr = global_thirst_corr
+    ),
+    thirst_intensity_plot = create_intensity_plot(data = data_ptds,
+    intensity_corr = intensity_corr),
+    fasting_duration_ptds_plot = create_fasting_duration_ptds_plot(data_ptds,
+    fluids_corr, food_corr),
     age_plot = create_age_ptds_plot(data_ptds),
     DT = create_DT(data_ptds),
+    distribution_plot = create_distribution_plot(data_ptds),
+
+    # correlations
+
+    global_thirst_corr = create_global_thirst_corr(data_ptds),
+intensity_corr = create_intensity_corr(data_ptds),
+fluids_corr = create_fluids_corr(data_ptds),
+food_corr = create_food_corr(data_ptds),
+pain_corr = create_pain_corr(data_ptds),
+
 
     manuscript = target(
       command = {

### 2020-07-23 manuscript/index.Rmd
draft sent to co-authors

@@
 Five items from the perioperative thirst discomfort scale (PTDS-5) formed a Mokken scale with evidence of invariant item ordering.
-The median  PTDS-5 score was `r ptds5`  from a possible range of 5 to 15.
+The median  PTDS-5 score was `r ptds5`  from a possible range of 0 to 10.
 The hierarchical pattern of responses to PTDS-5 items is interpretable in terms of respondents most easily endorsing an item related to discomfort associated with the desire to drink water through to first endorsing symptoms related to dryness of the mouth and lips before those related to the abnormal sensations of 'thick' saliva and a 'thick' tongue.
@@
 
-Availability of data and material (data transparency): All data used is available [here](https://aconway.dev/ptds).
+Availability of data and material (data transparency): All data used is available [here](https://github.com/awconway/ptds).
 
@@
 Although the initial validation of the PTDS-7 was conducted in a different context (i.e.
-post-surgical patients), it demonstrates strong potential for accurately assessing thirst-discomfort more generally for peri-operative use.
-The aim of this study was to test the validity and reliability of the PTDS-7 for measuring thirst discomfort for patients who are fasting prior to medical procedures.
+*post*-surgical patients), it demonstrates strong potential for accurately assessing thirst-discomfort more generally for peri-operative use.
+The aim of this study was to examine the validity and reliability of the PTDS-7 for measuring thirst discomfort for patients who are fasting *before* medical procedures.
 
@@
 The use of oxygen therapy (e.g., via nasal prongs) may contribute to the experience of thirst symptoms (e.g., dry mouth, dry throat).[@conchon2015perioperative].
-The RA recorded "yes" or "no" as to whether the participant was currently receiving oxygen therapy at the time the questionnaire was administered.
+The research assistant recorded "yes" or "no" as to whether the participant was currently receiving oxygen therapy at the time the questionnaire was administered.
 
@@
 If **H~i~** was below 0.3 or if the lower limit of the 95% CI (confidence interval) for *H~i~* was below 0.3, the item was to be excluded.
-Then, scale partitioning was carried out to explore the dimensions of the PTDS through increasing c (Lower bound c defines the minimum value of coefficients **H~i~** in the Mokken scale by 0.05 increments).[@molenaar2000mps5] Monotone homogeneity and invariant item ordering were investigated at the whole scale level, as no sub-scales were identified.
+Then, scale partitioning was carried out to explore the dimensions of the PTDS through increasing c (lower bound c defines the minimum value of coefficients **H~i~** in the Mokken scale by 0.05 increments).[@molenaar2000mps5] Monotone homogeneity and invariant item ordering were investigated at the whole scale level, as no sub-scales were identified.
 To assess for violations of invariant item ordering, crit values <40 were considered acceptable.[@molenaar2000mps5] Provided invariant item ordering holds for a set of items, the coefficient **H~T~** expresses the accuracy of the ordering, with values below 0.3 being unacceptable and above 0.4 indicating moderate invariant item ordering.[@Ligtvoet_2010] The coefficient score (rho), which is similar to Cronbach's alpha, was used to assess reliability of the scale.
 
-Construct validity was evaluated by using correlations to identify associations between scores on the Perioperative Thirst-Discomfort Scale and fasting duration.
-To demonstrate construct validity we hypothesized that higher scores on the PTDS would be associated with greater fasting duration.
+Construct validity was evaluated by using correlations to identify associations between scores on the PTDS and fasting duration.
+To investigate construct validity we hypothesized that higher scores on the PTDS would be associated with greater fasting duration.
 Convergent validity was assessed by using Kendall's *tau* to compare the PTDS score with the global thirst discomfort score.
@@
 From November 2019 to March 2020, we screened 203 patients for inclusion in the study.
-A total of 198 were eligible and 193 chose to participate.
+A total of 198 were eligible and 194 participated.
+One participant did not complete 1 of the PTDS-7 items and was excluded from the analysis.
 A summary of participant characteristics is displayed in Table \@ref(tab:tab1).
@@
 43% of patients were scheduled to have their procedure in the cardiac catheterization laboratory.
-A biopsy was the most common Interventional Radiology procedure (n=62; 32%), but a wide range of other procedures were included.
-The mean duration of fasting from food and non-clear fluids was `r food` hours and clear fluids was `r fluids`.
+A biopsy was the most common interventional radiology procedure (n=62; 32%), but a wide range of other procedures were included.
+The mean duration of fasting from food and non-clear fluids was `r food` hours and clear fluids was `r fluids` hours.
 
@@
 No items required removal from the PTDS-7 based on the condition that scalability coefficients (*H~i~* and lower bound of 95% CI) should be over 0.3 (Appendix Table \@ref(tab:Hvaluesptds7)).
-We used a lower bound for *c*, starting from 0.05 and increasing to 0.80 in 0.05 increments, to explore the dimensionality of the PTDS.
+We used a lower bound for *c*, starting from 0.05 and increasing to 0.80 in 0.05 increments, to explore the dimensionality of the PTDS-7.
 From 0.05 to 0.45, all of the items formed a single scale.
@@
 The distribution of PTDS-5 scores was right-skewed (Figure \@ref(fig:distribution)).
-There was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (Figure \@ref(fig:globalthirst)), as well as thirst intensity (Figure \@ref(fig:intensity)).
-In contrast, a longer duration of fasting was not linearly associated with PTDS-5 scores.
+There was a positive correlation between PTDS-5 scores and thirst intensity (Figure \@ref(fig:intensity)), as well as the global thirst discomfort rating (Figure \@ref(fig:globalthirst)).
 The associations between PTDS-5 scores with food and fluid duration is displayed in Figure \@ref(fig:fasting).
+Longer duration of fasting was not linearly associated with PTDS-5 scores.
 Likewise, the correlation between pain and PTDS-5 scores was not significant (Spearman's rho = `r round(drake::readd(pain_corr)$rho, 2)`; 95% CI = `r round(drake::readd(pain_corr)$CI_low, 2)` to `r round(drake::readd(pain_corr)$CI_high, 2)`).
@@
 
-Result indicate that the PTDS-5 is a reliable and precise way to order patients according to their thirst discomfort.
+Results indicate that the PTDS-5 is a reliable and precise way to order patients according to their thirst discomfort in the *pre-procedure* period.
 In addition, an acceptable accuracy for invariant item ordering was identified, meaning patients are likely to rank the PTDS-5 symptoms  of thirst discomfort in the same way.
-As such, the item difficulty hierarchy, presented in Table \@ref(tab:tab2) can be used as a guide to easily determine which patients are expreriencing worse thirst discomfort.
+As such, the item difficulty hierarchy, presented in Table \@ref(tab:tab2) can be used as a guide to easily determine which patients are expriencing worse thirst discomfort.
 The hierarchical pattern of responses to items in the Mokken scale is interpretable in terms of respondents more easily endorsing items related to a general desire to drink water through to first endorsing more specific symptoms related to dryness of the mouth and lips prior to endorsing discomfort associated with the sensations of 'thick' saliva and 'thick' tongue.
@@
 Selection bias should be considered because a convenience sample was used for this study.
-Also, it is uncertain if similar results would be obtained for patients who are fasted prior to undergoing procedures in other procedural settings.
-Only a very small number of participants had to be excluded from analyses due to missing data, so it is unlikely that attrition bias would have exerted a major influence on the results.
+Also, it is uncertain if similar results would be obtained for patients who are fasted prior to undergoing procedures in other settings.
+Only one patient was excluded from the analysis due to missing data, so it is unlikely attrition bias would have exerted a major influence on the results.
 Further research is required to evaluate the minimal detectable difference for the PTDS-5 as well as the minimal clinically important difference prior to use of this scale as an outcome measure in research.
@@
 
-```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 5 to 15 with higher scores indicating worse thirst discomfort."}
+```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 0 to 10 with higher scores indicating worse thirst discomfort."}
 knitr::include_graphics("figures/thirst-intensity.png")

### 2020-07-23 Analysis/grant_plots.R

@@ -10,16 +10,16 @@ library(devtools)
 ### Install development version of ggplot and run code in developer mode
 # move to development mode
 dev_mode(TRUE)
-# # Run if not already installed 
+# # Run if not already installed
 # install_github("tidyverse/ggplot2")
 
 ### Load in data and prepare for manual coding
-rcon <- redcapConnection(url="https://redcap.thetacollaborative.ca/api/", token="CABFAFF8E4259517B23F344818EE682F")
+rcon <- redcapConnection(url = "https://redcap.thetacollaborative.ca/api/", token = "CABFAFF8E4259517B23F344818EE682F")
 myData <- exportRecords(rcon, labels = FALSE)
 
 # Export data to csv for qualitative coding
 fasting_descriptions <- myData %>% select(id, fasting_description)
-write.csv(fasting_descriptions, file="./Analysis/fasting_descriptions.csv")
+write.csv(fasting_descriptions, file = "./Analysis/fasting_descriptions.csv")
 
 ### Import latest version of manually coded data
 coded_descriptions <- read.csv("./Analysis/fasting_descriptions_coded20200130.csv")
@@ -30,30 +30,36 @@ coded_descriptions <- coded_descriptions %>% filter(is.na(ambiguous))
 
 # Add summary variables
 coded_descriptions <- coded_descriptions %>%
-  mutate(fasting_type =  case_when(npo_midnight == 1 ~ 1,
-                                   npo_morning == 1 ~ 2,
-                                   npo_hrsprior == 1 ~ 3,
-                                   npo_dinner == 1 ~ 1,
-                                   separate_fooddrink == 1 ~ 4,
-                                   unclear_patient == 1 ~ 6,
-                                   missing == 1 ~ 7)) %>%
-  mutate(fasting_type_text = case_when(fasting_type %in% 1 ~ "\"No food or drink after midnight\"",
-                                       fasting_type == 2 ~ "\"No food or drink after __ AM\"",
-                                       fasting_type == 3 ~ "\"No food or fluids __ hours prior\"",
-                                       fasting_type == 4 ~ "\"No food __ hours prior and no\ndrink after __ AM this morning\"",
[…]

### 2020-07-23 R/create_clarity_plot.R

@@ -1,0 +1,27 @@
+#' @title create_clarity_plot
+#' @return
+#' @author Aaron Conway
+#' @export
+#' @importFrom dplyr count
+#' @importFrom ggplot2 ggplot aes geom_col
+#' labs theme ggsave
+#'
+
+create_clarity_plot <- function(data_ptds) {
+  clarity_plot <- data_ptds %>%
+    count(instructions_factor) %>%
+    drop_na() %>%
+    ggplot(aes(x = n, y = instructions_factor)) +
+    geom_col(aes(fill = instructions_factor, color = instructions_factor)) +
+    labs(
+      x = "Number of participants",
+      y = element_blank()
+    ) +
+    theme(
+      panel.background = element_blank(),
+      legend.position = "none",
+      axis.ticks.x = element_blank(),
+      axis.ticks.y = element_blank(),
+      text = element_text(family = "BioRhyme")
+    )
+}

### 2020-07-23 R/create_deviation_plot.R

@@ -1,0 +1,236 @@
+#' @title create_deviation_plot
+#' @return
+#' @author Aaron Conway
+#' @export
+
+#' @importFrom dplyr mutate case_when inner_join select
+#' filter desc n group_by pull lag
+#' @importFrom ggplot2 ggplot aes facet_wrap labeller
+#' geom_point geom_polygon alpha geom_vline geom_segment
+#' scale_x_continuous scale_y_continuous scale_color_manual
+#' labs geom_curve arrow unit geom_label theme element_blank
+#' element_text element_line margin ggsave
+#' @importFrom ggtext element_markdown
+#' @importFrom here here
+#' @importFrom tidyr gather
+#'
+create_deviation_plot <- function(data_ptds, coded_descriptions) {
+    coded_descriptions <- coded_descriptions %>%
+        mutate(fasting_type = case_when(
+            npo_midnight == 1 ~ 1,
+            npo_morning == 1 ~ 2,
+            npo_hrsprior == 1 ~ 3,
+            npo_dinner == 1 ~ 1,
+            separate_fooddrink == 1 ~ 4,
+            unclear_patient == 1 ~ 6,
+            missing == 1 ~ 7
+        )) %>%
+        mutate(fasting_type_text = case_when(
+            fasting_type %in% 1 ~ "\"No food or drink after midnight\"",
+            fasting_type == 2 ~ "\"No food or drink after __ AM\"",
+            fasting_type == 3 ~ "\"No food or fluids __ hours prior\"",
+            fasting_type == 4 ~ "\"No food __ hours prior and no\ndrink after __ AM this morning\"",
+            fasting_type == 6 ~ "Unclear instructions",
+            fasting_type == 7 ~ "Missing"
+        ))
+
+    # Join coded data to original data
+    coded_descriptions$id <- as.character(coded_descriptions$id)
+    coded_data <- data_ptds %>%
+        inner_join(coded_descriptions %>%
+            select(id, fasting_type, fasting_type_text), by = c("id" = "id"))
+
+    # Compute food and fluids fasting deviations
+    coded_data <- coded_data %>% mutate(
+        food_dur = as.double(date_time - food),
+        fluids_dur = as.double(date_time - fluids)
+    )
+
+    # Filter out screening arm and select only relevant columns
+    coded_data <- coded_data %>%
+        #   filter(redcap_event_name == "t3_arm_1") %>%
[…]

### 2020-07-23 R/create_distribution_plot.r

@@ -4,20 +4,21 @@
 #' @importFrom ggplot2 ggplot aes theme_minimal scale_x_continuous
 #' labs element_blank theme ggsave geom_histogram
 create_distribution_plot <- function(data_ptds) {
-    data_ptds %>%
+    distribution_plot <- data_ptds %>%
         ggplot(aes(x = ptds)) +
-        geom_histogram(binwidth = 0.5) +
-        theme_minimal() +
+        geom_histogram(binwidth = 0.5, colour = "#2a6ebb", fill = "#2a6ebb") +
         labs(
             y = "Count",
-            x = "\nPTDS-5 (range is 5 to 15; higher scores indicate worse thirst discomfort)"
+            x = "\nPTDS-5 (range is 0 to 10; higher scores indicate worse thirst discomfort)"
         ) +
         scale_x_continuous(
-            breaks = c(seq(5, 15, by = 1))
+            breaks = seq(0, 10, by = 1)
+        ) +
+        theme(
+            panel.background = element_blank(),
+            legend.position = "none",
+            axis.ticks.x = element_blank(),
+            axis.ticks.y = element_blank(),
+            text = element_text(family = "BioRhyme")
         )
-
-    ggsave(
-        device = "png", filename = here("manuscript/figures/distribution.png"),
-        width = 174, units = "mm"
-    )
 }

### 2020-07-23 R/create_fasting_duration_ptds_plot.R

@@ -9,21 +9,22 @@
 #' @importFrom dplyr summarize mutate
 #' @importFrom glue glue
 #' @importFrom tibble tibble
-create_fasting_duration_ptds_plot <- function(data = data_ptds,
-  fluids_corr, food_corr) {
 
-rho <- round(fluids_corr$rho, 2)
-rho_low <- round(fluids_corr$CI_low, 2)
-rho_high <- round(fluids_corr$CI_high, 2)
+create_fasting_duration_ptds_plot <- function(data = data_ptds,
+                                              fluids_corr, food_corr) {
+
+  rho <- round(fluids_corr$rho, 2)
+  rho_low <- round(fluids_corr$CI_low, 2)
+  rho_high <- round(fluids_corr$CI_high, 2)
 
   ptds_fluids_cor <- tibble(
-      # location of each text label in data coordinates
-      ptds = 15,
-      fluids_duration = 20,
-      # text label containing rho value
-      label = glue("*rho* = {rho}
+  # location of each text label in data coordinates
+  ptds = 10,
+  fluids_duration = 20,
+  # text label containing rho value
+  label = glue("*rho* = {rho}
       (95% CI = {rho_low} to {rho_high})")
-    )
+  )
 
   fluids <- data %>%
     ggplot(aes(y = ptds, x = fluids_duration)) +
@@ -45,57 +46,49 @@ rho_high <- round(fluids_corr$CI_high, 2)
       data = ptds_fluids_cor,
       aes(label = label),
       hjust = 1, vjust = 2
+    ) +
+    scale_y_continuous(
+      breaks = seq(0, 10, by = 2)
     )
 
-  ptds_food_cor <- data %>%
-    drop_na(ptds, food_duration) %>%
-    summarize(r_square = cor(ptds, food_duration)^2) %>%
-    mutate(
-      # location of each text label in data coordinates
-      ptds = 14,
-      food_duration = 25,
-      # text label containing r^2 value
-      label = glue("*r*<sup>2</sup> = {round(r_square, 2)}")
-    )
-
-    rho <- round(food_corr$rho, 2)
-rho_low <- round(food_corr$CI_low, 2)
-rho_high <- round(food_corr$CI_high, 2)
+  rho <- round(food_corr$rho, 2)
+  rho_low <- round(food_corr$CI_low, 2)
+  rho_high <- round(food_corr$CI_high, 2)
 
   ptds_food_cor <- tibble(
-      # location of each text label in data coordinates
-      ptds = 15,
[…]

### 2020-07-23 R/create_global_thirst_plot.R

@@ -7,31 +7,31 @@
 #' @importFrom ggplot2 ggplot aes theme_minimal
 #' labs element_blank theme ggsave
 #' @importFrom ggtext geom_richtext
+
 create_global_thirst_plot <- function(data,
                                       global_thirst_corr) {
 
-tau <- round(global_thirst_corr$tau, 2)
-tau_low <- round(global_thirst_corr$CI_low, 2)
-tau_high <- round(global_thirst_corr$CI_high, 2)
+  tau <- round(global_thirst_corr$tau, 2)
+  tau_low <- round(global_thirst_corr$CI_low, 2)
+  tau_high <- round(global_thirst_corr$CI_high, 2)
 
   cor <- tibble(
-    ptds = 12,
+    ptds = 8,
     global_thirst_factor = "Extremely uncomfortable",
     # text label containing r^2 value
     label = glue("*tau* = {tau}; (95% CI = {tau_low} to {tau_high})")
   )
 
-data %>%
+  data %>%
     ggplot(aes(
       x = ptds, y = reorder(global_thirst_factor, desc(global_thirst_factor)
       ),
       colour = global_thirst_factor
     )) +
     geom_beeswarm(groupOnX = FALSE, alpha = 0.5, size = 2) +
-    theme_minimal() +
     labs(
       y = element_blank(),
-      x = "\nPTDS-5 (5-15; higher scores indicate worse thirst discomfort)"
+      x = "\nPTDS-5 (0-10; higher scores indicate worse thirst discomfort)"
     ) +
     theme(
       legend.position = "none",
@@ -41,10 +41,8 @@ data %>%
       data = cor,
       aes(label = label),
       hjust = 1, vjust = 2
+    ) +
+    scale_x_continuous(
+      breaks = seq(0, 10, by = 2)
     )
-
-  ggsave(
-    device = "png", filename = here("manuscript/figures/global-thirst.png"),
-    width = 174, units = "mm"
-  )
 }

### 2020-07-23 R/create_instructions_plot.R

@@ -1,0 +1,75 @@
+#' @title
+#' @export
+
+create_instruction_plot <- function(coded_descriptions,
+                                    data_ptds) {
+  coded_descriptions <- coded_descriptions %>%
+    mutate(fasting_type = case_when(
+      npo_midnight == 1 ~ 1,
+      npo_morning == 1 ~ 2,
+      npo_hrsprior == 1 ~ 3,
+      npo_dinner == 1 ~ 1,
+      separate_fooddrink == 1 ~ 4,
+      unclear_patient == 1 ~ 6,
+      missing == 1 ~ 7
+    )) %>%
+    mutate(fasting_type_text = case_when(
+      fasting_type %in% 1 ~ "\"No food or drink after midnight\"",
+      fasting_type == 2 ~ "\"No food or drink after __ AM\"",
+      fasting_type == 3 ~ "\"No food or fluids __ hours prior\"",
+      fasting_type == 4 ~ "\"No food __ hours prior and no\ndrink after __ AM this morning\"",
+      fasting_type == 6 ~ "Unclear instructions",
+      fasting_type == 7 ~ "Missing"
+    ))
+
+  # Join coded data to original data
+  coded_descriptions$id <- as.character(coded_descriptions$id)
+  coded_data <- data_ptds %>%
+    inner_join(coded_descriptions %>%
+      select(id, fasting_type, fasting_type_text), by = c("id" = "id"))
+
+  # Compute food and fluids fasting deviations
+  coded_data <- coded_data %>% mutate(
+    food_dur = as.double(date_time - food),
+    fluids_dur = as.double(date_time - fluids)
+  )
+
+  # Filter out screening arm and select only relevant columns
+  coded_data <- coded_data %>%
+    #   filter(redcap_event_name == "t3_arm_1") %>%
+    select(id, fasting_type, fasting_type_text, food_dur, fluids_dur, procedure, procedure_factor)
+
+  coded_data %>%
+    group_by(procedure_factor) %>%
+    mutate(department = ifelse(procedure < 5, "Cardiac Cath Lab", no = "Interventional Radiology")) %>%
+    group_by(department) %>%
+    count(fasting_type_text) %>%
+    drop_na() %>%
+    filter(n > 1) %>%
+    ggplot2::ggplot(aes(
+      y = fasting_type_text,
+      x = n,
+      fill = fasting_type_text
+    )) +
+    geom_col() +
+    facet_wrap(~department, nrow = 2) +
+    theme(
[…]

### 2020-07-23 R/create_thirst_intensity_plot.R

@@ -1,4 +1,3 @@
-
 ##' @title
 ##' @return
 ##' @author Aaron Conway
@@ -10,40 +9,39 @@
 #' @importFrom ggtext geom_richtext
 
 create_intensity_plot <- function(data,
-intensity_corr) {
-  
-tau <- round(intensity_corr$tau, 2)
-tau_low <- round(intensity_corr$CI_low, 2)
-tau_high <- round(intensity_corr$CI_high, 2)
+                                  intensity_corr) {
+
+  tau <- round(intensity_corr$tau, 2)
+  tau_low <- round(intensity_corr$CI_low, 2)
+  tau_high <- round(intensity_corr$CI_high, 2)
 
   cor <- tibble(
-    ptds = 14.5,
+    ptds = 9.5,
     thirst_intensity = 2,
     # text label containing r^2 value
     label = glue("*tau* = {tau}; (95% CI = {tau_low} to {tau_high})")
   )
-  
-  
+
+
   data %>%
     # mutate(thirst_intensity = as_factor(thirst_intensity)) %>%
     ggplot(aes(x = ptds, y = factor(thirst_intensity),
                colour = factor(thirst_intensity))) +
     geom_beeswarm(alpha = 0.5, size = 2, groupOnX = FALSE) +
-    theme_minimal() +
     labs(
       y = "\nThirst intensity (0-10; higher scores indicate greater thirst intensity)",
-      x = "PTDS-5 (5-15; higher scores indicate worse thirst discomfort)\n"
+      x = "PTDS-5 (0-10; higher scores indicate worse thirst discomfort)\n"
     ) +
     theme(
       legend.position = "none",
       plot.title.position = "plot"
-    )+
+    ) +
     geom_richtext(
       data = cor,
       aes(label = label),
       hjust = 1, vjust = 2
+    )+
+    scale_x_continuous(
+      breaks = seq(0, 10, by = 2)
     )
-
-  ggsave(device = "png", filename = here("manuscript/figures/thirst-intensity.png"),
-         width = 174, units = "mm")
 }

### 2020-07-23 R/mokken_variables.R

@@ -8,9 +8,9 @@ mokken_variables <- function(participation_df) {
   # variables for mokken scaling
   participation_df %>%
     select(starts_with("ptds")) %>%
-    select(ends_with("factor")) %>%
+    select(!ends_with("factor")) %>%
     rename_all(.funs = ~ str_remove_all(., "ptds_")) %>%
-    rename_all(.funs = ~ str_remove_all(., "_factor")) %>%
-    mutate_if(is.factor, ~ as.numeric(.x)) %>%
+    #rename_all(.funs = ~ str_remove_all(., "_factor")) %>%
+    #mutate_if(is.factor, ~ as.numeric(.x)) %>%
     as.data.frame()
 }

### 2020-07-23 R/plan.R

@@ -6,7 +6,9 @@
 #' @importFrom dplyr select
 #' @importFrom here here
 #' @importFrom drake drake_plan knitr_in
+#' @importFrom ggplot2 ggsave
 #'
+
 get_analysis_plan <- function() {
   drake_plan(
     original_data = read_csv(here("data/ThirstDiscomfortScal_DATA_2020-03-24_2016.csv")),
@@ -42,22 +44,98 @@ get_analysis_plan <- function() {
       data = data_ptds,
       global_thirst_corr = global_thirst_corr
     ),
-    thirst_intensity_plot = create_intensity_plot(data = data_ptds,
-    intensity_corr = intensity_corr),
-    fasting_duration_ptds_plot = create_fasting_duration_ptds_plot(data_ptds,
-    fluids_corr, food_corr),
+    global_thirst_plot_png = ggsave(
+      plot = global_thirst_plot,
+      device = "png", filename = here("manuscript/figures/global-thirst.png"),
+      width = 174, units = "mm"
+    ),
+    thirst_intensity_plot = create_intensity_plot(
+      data = data_ptds,
+      intensity_corr = intensity_corr
+    ),
+
+    thirst_intensity_plot_png = ggsave(
+      plot = thirst_intensity_plot,
+      device = "png",
+      filename = here("manuscript/figures/thirst-intensity.png"),
+      width = 174, units = "mm"
+    ),
+
+    fasting_duration_ptds_plot = create_fasting_duration_ptds_plot(
+      data_ptds,
+      fluids_corr, food_corr
+    ),
+
+    fasting_duration_ptds_plot_png = ggsave(
+      plot = fasting_duration_ptds_plot,
+      device = "png",
+      filename = here("manuscript/figures/fasting.png"),
+      width = 174, units = "mm"
+    ),
+
     age_plot = create_age_ptds_plot(data_ptds),
     DT = create_DT(data_ptds),
     distribution_plot = create_distribution_plot(data_ptds),
+    distribution_plot_png = ggsave(
+      plot = distribution_plot,
+      device = "png",
+      filename = here("manuscript/figures/distribution.png"),
+      width = 174,
+      units = "mm"
+    ),
 
     # correlations
 
     global_thirst_corr = create_global_thirst_corr(data_ptds),
-intensity_corr = create_intensity_corr(data_ptds),
-fluids_corr = create_fluids_corr(data_ptds),
-food_corr = create_food_corr(data_ptds),
[…]

### 2020-07-30 manuscript/index.Rmd
dockerfile added

@@
 The experience of pain may influence the participant's perception of thirst.[@Pierotti_2018] Participants were asked to rate their current level of pain on a scale of 0 to 10, where 0 represents "no pain" and 10 represents that their pain as "unimaginable/unspeakable".
-The use of oxygen therapy (e.g., via nasal prongs) may contribute to the experience of thirst symptoms (e.g., dry mouth, dry throat).[@conchon2015perioperative].
-The research assistant recorded "yes" or "no" as to whether the participant was currently receiving oxygen therapy at the time the questionnaire was administered.
+The use of oxygen therapy (e.g., via nasal prongs) may contribute to the experience of thirst symptoms (e.g., dry mouth, dry throat).[@conchon2015perioperative] The research assistant recorded "yes" or "no" as to whether the participant was currently receiving oxygen therapy at the time the questionnaire was administered.
 
@@
 It should be noted that an investigation of the dimensionality of the 7-item PTDS had not been performed prior to our study.
-A uni-dimensional solution was ientified in our analysis.
+A uni-dimensional solution was identified in our analysis.
 This should be expected considering the small number of items that were evaluated.

### 2020-07-30 Analysis/grant_plots.R

@@ -1,190 +1,0 @@
-library(dplyr)
-library(redcapAPI)
-library(lubridate)
-library(ggplot2)
-library(gridExtra)
-library(tidyr)
-library(ggtext)
-library(devtools)
-
-### Install development version of ggplot and run code in developer mode
-# move to development mode
-dev_mode(TRUE)
-# # Run if not already installed
-# install_github("tidyverse/ggplot2")
-
-### Load in data and prepare for manual coding
-rcon <- redcapConnection(url = "https://redcap.thetacollaborative.ca/api/", token = "CABFAFF8E4259517B23F344818EE682F")
-myData <- exportRecords(rcon, labels = FALSE)
-
-# Export data to csv for qualitative coding
-fasting_descriptions <- myData %>% select(id, fasting_description)
-write.csv(fasting_descriptions, file = "./Analysis/fasting_descriptions.csv")
-
-### Import latest version of manually coded data
-coded_descriptions <- read.csv("./Analysis/fasting_descriptions_coded20200130.csv")
-
-### Data cleaning
-# Remove ambiguous codings
-coded_descriptions <- coded_descriptions %>% filter(is.na(ambiguous))
-
-# Add summary variables
-coded_descriptions <- coded_descriptions %>%
-  mutate(fasting_type = case_when(
-    npo_midnight == 1 ~ 1,
-    npo_morning == 1 ~ 2,
-    npo_hrsprior == 1 ~ 3,
-    npo_dinner == 1 ~ 1,
-    separate_fooddrink == 1 ~ 4,
-    unclear_patient == 1 ~ 6,
-    missing == 1 ~ 7
-  )) %>%
-  mutate(fasting_type_text = case_when(
-    fasting_type %in% 1 ~ "\"No food or drink after midnight\"",
-    fasting_type == 2 ~ "\"No food or drink after __ AM\"",
-    fasting_type == 3 ~ "\"No food or fluids __ hours prior\"",
-    fasting_type == 4 ~ "\"No food __ hours prior and no\ndrink after __ AM this morning\"",
-    fasting_type == 6 ~ "Unclear instructions",
-    fasting_type == 7 ~ "Missing"
-  ))
-
-# Join coded data to original data
-coded_descriptions$id <- as.character(coded_descriptions$id)
-coded_data <- myData %>% inner_join(coded_descriptions %>% select(id, fasting_type, fasting_type_text), by = c("id" = "id"))
[…]

### 2020-07-30 R/moscales_lowerbounds.R

@@ -1,7 +1,7 @@
-##' @title Bounds
-##' @param x data frame
-##' @return
-##' @export
+#' @title Bounds
+#' @param x data frame
+#' @return
+#' @export
 #' @importFrom mokken aisp
 moscales_lowerbounds <- function(x, lowerbounds = seq(from = 0.05, to = 0.80, by = 0.05)) {
   ret.value <- NULL

### 2020-07-30 R/plan.R
targets not required for manuscript commented

@@ -94,48 +94,48 @@ get_analysis_plan <- function() {
 
     # deviations
 
-    coded_descriptions = read.csv(
-      here::here("Analysis/fasting_descriptions_complete.csv")
-    ),
+    #coded_descriptions = read.csv(
+    #  here::here("Analysis/fasting_descriptions_complete.csv")
+    #),
 
-    deviation_plot = create_deviation_plot(
-      data_ptds = data_ptds,
-      coded_descriptions = coded_descriptions
-    ),
-    deviation_plot_png = ggsave(
-      plot = deviation_plot,
-      filename = here("plots", "fasting-deviation-plot.png"),
-      height = 15,
-      width = 12,
-      units = "in",
-      device = "png"
-    ),
-    deviation_plot_svg = ggsave(
-      plot = deviation_plot,
-      filename = here("plots", "fasting-deviation-plot.svg"),
-      height = 15,
-      width = 12,
-      units = "in",
-      device = "svg"
-    ),
-
-    # instructions
-
-    clarity_plot = create_clarity_plot(data_ptds),
-    clarity_plot_png = ggsave(
-      plot = clarity_plot,
-      filename = here("plots", "clarity-plot.svg"),
-      device = "svg"
-    ),
-    instructions_plot = create_instruction_plot(
-      data_ptds = data_ptds,
-      coded_descriptions = coded_descriptions
-    ),
-    instructions_plot_svg = ggsave(
-      plot = instructions_plot,
-      filename = here("plots", "instructions_plot.svg"),
-      device = "svg"
-    ),
+    #deviation_plot = create_deviation_plot(
+    #  data_ptds = data_ptds,
+    #  coded_descriptions = coded_descriptions
+    #),
+    #deviation_plot_png = ggsave(
+    #  plot = deviation_plot,
+    #  filename = here("plots", "fasting-deviation-plot.png"),
+    #  height = 15,
+    #  width = 12,
+    #  units = "in",
+    #  device = "png"
+    #),
+    #deviation_plot_svg = ggsave(
+    #  plot = deviation_plot,
+    #  filename = here("plots", "fasting-deviation-plot.svg"),
+    #  height = 15,
+    #  width = 12,
+    #  units = "in",
+    #  device = "svg"
+    #),
+    #
+    ## instructions
+    #
+    #clarity_plot = create_clarity_plot(data_ptds),
+    #clarity_plot_png = ggsave(
+    #  plot = clarity_plot,
[…]

### 2020-08-12 manuscript/manuscript.Rmd
pdf version

--- 2020-07-30 manuscript/index.Rmd
@@
 ---
-title: "Validating the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures"
+title: A Template for Two or One-Column Vignettes
+
+# Use letters for affiliations
 author:
-  - Aaron Conway PhD:
-      email: aaron.conway@utoronto.ca
-      phone: +1 416 946 7112
-      address: 585 University Ave | Toronto, ON  M5G 2N2
-      institute: [PMCC, LBFON, QUT]
-      correspondence: true
-  - Kristina Chang MScN:
-      institute: [PMCC]
-      correspondence: false
-  - Megan Bittner BSc:
-      institute: [LBFON]
-      correspondence: false
-  - Dan Phan BSc:
-      institute: [LBFON]
-      correspondence: false
-  - Navpreet Kamboj BScN:
-      institute: [LBFON]
-      correspondence: false
-  - Matteo Parotto MD, PhD:
-      institute: [TGH, UoT]
-      correspondence: false
-  - Amanda Matthews:
-      institute: [PMCC]
-      correspondence: false
-  - Sherryl Alexander:
-      institute: [PMCC]
-      correspondence: false
-  - Pauline Glaves:
-      institute: [PMCC]
-      correspondence: false
-  - Shahvand Masihi:
-      institute: [IR]
-      correspondence: false
-  - Sebastian Mafeld:
-      institute: [IR]
-      correspondence: false
-institute:
-  - PMCC: Peter Munk Cardiac Centre, University Health Network, Toronto, Canada
-  - LBFON: Lawrence S. Bloomberg Faculty of Nursing, University of Toronto, Toronto, Canada
-  - QUT: School of Nursing, Queensland University of Technology (QUT), Brisbane, Australia
-  - TGH: Department of Anesthesia and Pain Management, Toronto General Hospital, UHN, Toronto, Canada
-  - UoT: Department of Anesthesia and Interdepartmental Division of Critical Care Medicine, University of Toronto, Toronto, Canada
-  - IR: Interventional Radiology, JDMI, Toronto General Hospital
-bibliography: Extras/references.bib
-csl: Extras/sp-basic.csl
-output:
-  officedown::rdocx_document:
-    reference_docx: Extras/manuscript_template.docx
-    pandoc_args:
-      - --lua-filter=Extras/scholarly-metadata.lua
-      - --lua-filter=Extras/author-info-blocks.lua
+  - name: ""
+#     affiliation: a
+#   - name: Second Author
+#     affiliation: a,b
+# address:
+#   - code: a
+#     address: Institute of Smoke and Magic, University of Sometown, Sometown, XY, 12345
+#   - code: b
+#     address: Department of Neat Tricks, Whereever State University, Someplace, MC, 67890
+
+# Optional: line of arbitrary text with additional information.
+# Could be used, for example, to mention the bibliographic info in a post-print.
+# If not specified, defaults to "This version was compiled on \today"
+#date_subtitle: Published in *Journal of Statistical Software*, 2018
+
+# For footer text  TODO(fold into template, allow free form two-authors)
+lead_author_surname: Author and Author
+
+# Place eg a DOI URL or CRAN Package URL here
+doi_footer: "https://cran.r-project.org/package=YourPackage"
+
+# Abstract
+abstract: |
+  Your abstract will be typeset here, and used by default a visually distinctive font.
+  An abstract should explain to the general reader the major contributions of the article.
+
+# Optional: Acknowledgements
+acknowledgements: |
+  This template package builds upon, and extends, the work of the excellent
+  [rticles](https://cran.r-project.org/package=rticles) package, and both packages rely on the
+  [PNAS LaTeX](http://www.pnas.org/site/authors/latex.xhtml) macros. Both these sources are
+  gratefully acknowledged as this work would not have been possible without them.  Our extensions
+  are under the same respective licensing term
+  ([GPL-3](https://www.gnu.org/licenses/gpl-3.0.en.html) and
+  [LPPL (>= 1.3)](https://www.latex-project.org/lppl/)).
+
+# Optional: One or more keywords
+keywords:
+  - one
+  - two
+  - optional
+  - keywords
+  - here
+
+# Paper size for the document, values of letter and a4
+papersize: letter
+
+# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
+fontsize: 9pt
+
+# Optional: Force one-column layout, default is two-column
+#one_column: true
+
+# Optional: Enables lineno mode, but only if one_column mode is also true
+#lineno: true
+
+# Optional: Enable one-sided layout, default is two-sided
+#one_sided: true
+
+# Optional: Enable section numbering, default is unnumbered
+#numbersections: true
+
+# Optional: Specify the depth of section number, default is 5
+#secnumdepth: 5
+
+# Optional: Skip inserting final break between acknowledgements, default is false
+skip_final_break: true
+
+# Optional: Bibliography 
+bibliography: references
+
+# Optional: Enable a 'Draft' watermark on the document
+#watermark: true
+
+# Customize footer, eg by referencing the vignette
+footer_contents: "Perioperative thirst discomfort scale"
+
+# Produce a pinp document
+output: pinp::pinp
+
 ---
@@
 ```{r global_options, include=FALSE}
-knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
-
+knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.retina = 5)
 library(drake)
 library(magrittr)
-library(flextable)
+# library(flextable)
 
@@
 ```
-
-\pagebreak
 
@@
 Mokken scaling analysis was conducted to investigate the unidimensionality and hierarchical nature of the scale.
-Correlations between perioperative thirst discomfort scale scores and fasting durations were calculated to evaluate construct validity and convergent validity was evaluated by comparing scores with global thirst discomfort and thirst intensity items.
+Correlations between perioperative thirst discomfort scale scores and fasting durations were calculated to evaluate construct validity and convergent validity was evaluated by comparing scores with global thirst discomfort and thirst intensity ratings.
 
@@
 
-\pagebreak
-
 ## Declarations
@@
 
-Author contributions: 
-AC: Study design, data analysis and wrote manuscript; KC: Study design, revised the manuscript for important intellectual content;  MB: Data collection, revised the manuscript for important intellectual content; DP: Data collection, revised the manuscript for important intellectual content; NK: Study design, revised the manuscript for important intellectual content; AM: Study design, revised the manuscript for important intellectual content; MP: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; PG: Study design, revised the manuscript for important intellectual content; SA: Study design, revised the manuscript for important intellectual content.
-
-\pagebreak
+Author contributions: AC: Study design, data analysis and wrote manuscript; KC: Study design, revised the manuscript for important intellectual content; MB: Data collection, revised the manuscript for important intellectual content; DP: Data collection, revised the manuscript for important intellectual content; NK: Study design, revised the manuscript for important intellectual content; AM: Study design, revised the manuscript for important intellectual content; MP: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; PG: Study design, revised the manuscript for important intellectual content; SA: Study design, revised the manuscript for important intellectual content.
 
@@
 As a direct result of prolonged pre-procedure fasting, symptoms of thirst discomfort have been reported as common and severe.
-In a qualitative study where 12 participants were interviewed from a tertiary hospital in Australia, surgical patients and patients who adhered to prolonged fasting instructions described the discomfort from thirst symptoms to be the worst physical effect of fasting.[@carey2015qualitative] Similarly, Madsen et al.
-interviewed a convenience sample of 50 adult surgical patients who reported that thirst symptoms caused more discomfort than hunger, sleep, or anxiety related to the procedure.[@madsen1998perioperative] However, thirst and its symptoms continue to be undervalued, under-reported and infrequently assessed by health care providers, including the nursing team.[@milani2016thirst]
+In a qualitative study where 12 participants were interviewed from a tertiary hospital in Australia, surgical patients and patients who adhered to prolonged fasting instructions described the discomfort from thirst symptoms to be the worst physical effect of fasting.[@carey2015qualitative] Similarly, Madsen et al. interviewed a convenience sample of 50 adult surgical patients who reported that thirst symptoms caused more discomfort than hunger, sleep, or anxiety related to the procedure.[@madsen1998perioperative] However, thirst and its symptoms continue to be undervalued, under-reported and infrequently assessed by health care providers, including the nursing team.[@milani2016thirst]
 
 Despite the relevance and value of assessing thirst-discomfort of patients, the subjective experience of thirst presents challenges in developing a valid and reliable tool to succinctly and accurately measure its symptoms and level of discomfort prior to procedures.
-A thirst-discomfort scale for perioperative use was developed at a surgical centre affiliated with an accredited public hospital in Brazil.[@Martins_2017] The seven-item perioperative thirst-discomfort scale (PTDS-7) was developed in three stages, including face and content validation and reliability, and based on the Consensus-based Standards for the selection of health status Measurement Instruments (COSMIN) checklist.[@mokkink2010cosmin] Inter-rater reliability was tested through inter-observer equivalence where a pair of nurses independently, but simultaneously administered the scale among 70 patients.
+A thirst-discomfort scale for perioperative use was developed at a surgical centre affiliated with an accredited public hospital in Brazil.[@Martins_2017] The seven-item perioperative thirst-discomfort scale (PTDS) was developed in three stages, including face and content validation and reliability, and based on the Consensus-based Standards for the selection of health status Measurement Instruments (COSMIN) checklist.[@mokkink2010cosmin] Inter-rater reliability was tested through inter-observer equivalence where a pair of nurses independently, but simultaneously administered the scale among 70 patients.
 Six items on the scale had a weighted kappa coefficient of 1, while the item "I feel like drinking water" was 0.97.[@Martins_2017] Cronbach's alpha was used to test the internal consistency of the scale with a value of 0.91.[@Martins_2017]
 
-Although the initial validation of the PTDS-7 was conducted in a different context (i.e.
-*post*-surgical patients), it demonstrates strong potential for accurately assessing thirst-discomfort more generally for peri-operative use.
-The aim of this study was to examine the validity and reliability of the PTDS-7 for measuring thirst discomfort for patients who are fasting *before* medical procedures.
+Although the initial validation of the PTDS was conducted in a different context (i.e. *post*-surgical patients), it demonstrates strong potential for accurately assessing thirst-discomfort more generally for peri-operative use.
+The aim of this study was to examine the validity and reliability of the PTDS for measuring thirst discomfort for patients who are fasting *before* medical procedures.
 
@@
 
-Mokken scaling is a method for establishing if items in a scale conform to a cumulative, hierarchical structure.[@perng2012construct] An intentionally simple example of a hierarchical scale is a set of items enquiring about height that increases in increments of 10cm from 100cm to 200cm.
+Mokken scaling is a method for establishing if items in a scale conform to a cumulative, hierarchical structure.[@perng2012construct] When an instrument contains hierarchically ordered items, the items can be ordered from those indicating mild to severe symptoms.
+An intentionally simple example of a hierarchical scale is a set of items enquiring about height that increases in increments of 10cm from 100cm to 200cm.
 In this example, taller people would endorse more items than those who were shorter.[@watson2008hierarchy]
@@
 *H* measures the extent to which items are arranged as expected by their mean values along the latent trait.
-*H* > 0.3 is the minimum acceptable value, with values > 0.4 indicating a moderate scale and > 0.5 indicates a strong scale.
+*H* \> 0.3 is the minimum acceptable value, with values \> 0.4 indicating a moderate scale and \> 0.5 indicates a strong scale.
 
@@
 Then, scale partitioning was carried out to explore the dimensions of the PTDS through increasing c (lower bound c defines the minimum value of coefficients **H~i~** in the Mokken scale by 0.05 increments).[@molenaar2000mps5] Monotone homogeneity and invariant item ordering were investigated at the whole scale level, as no sub-scales were identified.
-To assess for violations of invariant item ordering, crit values <40 were considered acceptable.[@molenaar2000mps5] Provided invariant item ordering holds for a set of items, the coefficient **H~T~** expresses the accuracy of the ordering, with values below 0.3 being unacceptable and above 0.4 indicating moderate invariant item ordering.[@Ligtvoet_2010] The coefficient score (rho), which is similar to Cronbach's alpha, was used to assess reliability of the scale.
+To assess for violations of invariant item ordering, crit values \<40 were considered acceptable.[@molenaar2000mps5] Provided invariant item ordering holds for a set of items, the coefficient **H~T~** expresses the accuracy of the ordering, with values below 0.3 being unacceptable and above 0.4 indicating moderate invariant item ordering.[@Ligtvoet_2010] The coefficient score (rho), which is similar to Cronbach's alpha, was used to assess reliability of the scale.
 
@@
 From 0.05 to 0.45, all of the items formed a single scale.
-The item, *'I have a bad taste in my mouth'*, was excluded by selecting the remaining items which showed uni-dimensionality at a threshold level of .50.
+The item, '*I have a bad taste in my mouth*', was excluded by selecting the remaining items which showed uni-dimensionality at a threshold level of .50.
 The 6 remaining items (PTDS-6) were further examined for the criteria of a Mokken scale and item invariant ordering.
@@
 The item difficult hierarchy for the PTDS-5 is presented in Table \@ref(tab:tab2).
-The most highly endorsed symptom (i.e.
-the least severe symptom of thirst discomfort in the PTDS-5) was being bothered by a desire to drink water.
-The least endorsed symptom (i.e.
-the most severe symptom of thirst discomfort in the PTDS-5) was being bothered by the perception of a 'thick tongue'.
+The most highly endorsed symptom (i.e. the least severe symptom of thirst discomfort in the PTDS-5) was being bothered by a desire to drink water.
+The least endorsed symptom (i.e. the most severe symptom of thirst discomfort in the PTDS-5) was being bothered by the perception of a 'thick tongue'.
 The items in the PTDS-5 form a strong Mokken scale (H~s~ = `r drake::readd(H_values_ptds5)$H[1]`).
+
+```{r }
+drake::readd(item_hierarchy_df_ptds5) %>% 
+  dplyr::select(item) 
+```
 
@@
 The hierarchical pattern of responses to items in the Mokken scale is interpretable in terms of respondents more easily endorsing items related to a general desire to drink water through to first endorsing more specific symptoms related to dryness of the mouth and lips prior to endorsing discomfort associated with the sensations of 'thick' saliva and 'thick' tongue.
-In more practical terms, a patient who reports feeling bothered by a sensation that saliva is thick, could be interpreted as experiencing worse thirst discomfort than a patient who did not endorse any of the items lower down on the item difficulty hierarchy (i.e.
-'my lips are dry' and 'my mouth is dry').
+In more practical terms, a patient who reports feeling bothered by a sensation that saliva is thick, could be interpreted as experiencing worse thirst discomfort than a patient who did not endorse any of the items lower down on the item difficulty hierarchy (i.e. 'my lips are dry' and 'my mouth is dry').
 
@@
 
-\pagebreak
+<!-- ## References -->
 
-## References
+<!-- ::: {#refs} -->
+<!-- ::: -->
 
-<div id="refs"></div>
+<!-- ## Figure legend -->
 
-\pagebreak
+<!-- Fig. \@ref(fig:distribution) Distribution of PTDS-5 scores -->
 
-## Figure legend
+<!-- Fig. \@ref(fig:intensity) Association between PTDS-5 and thirst intensity -->
 
-Fig. \@ref(fig:distribution) Distribution of PTDS-5 scores
+<!-- Fig. \@ref(fig:globalthirst) Association between PTDS-5 and global thirst rating -->
 
-Fig. \@ref(fig:intensity) Association between PTDS-5 and thirst intensity
-
-Fig. \@ref(fig:globalthirst) Association between PTDS-5 and global thirst rating
-
-Fig. \@ref(fig:fasting) Association between PTDS-5 and duration of fasting
-
-\pagebreak
+<!-- Fig. \@ref(fig:fasting) Association between PTDS-5 and duration of fasting -->
 
@@
 drake::loadd(summary_table)
-gtsummary::as_flextable(summary_table)
-```
-
-\pagebreak
-
-```{r tab.id = "tab2", tab.cap= "Item hierarchy of the PTDS-5"}
-flextable::flextable(drake::readd(item_hierarchy_df_ptds5)) %>% 
-  autofit()
+# gtsummary::as_flextable(summary_table)
+gtsummary::as_kable_extra(summary_table, booktabs = T)
 ```
@@
 ```{r distribution, fig.cap = "Distribution of PTDS-5 scores"}
-knitr::include_graphics("figures/distribution.png")
+# knitr::include_graphics("figures/distribution.png")
+drake::readd(distribution_plot)
 ```
 
-\pagebreak
-
 ```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 0 to 10 with higher scores indicating worse thirst discomfort."}
-knitr::include_graphics("figures/thirst-intensity.png")
+# knitr::include_graphics("figures/thirst-intensity.png")
+drake::readd(thirst_intensity_plot)
 ```
 
-\pagebreak
-
 ```{r globalthirst, fig.cap="Association between PTDS-5 and global thirst rating"}
-knitr::include_graphics("figures/global-thirst.png")
+# knitr::include_graphics("figures/global-thirst.png")
+library(dplyr)
+drake::readd(global_thirst_plot)
 ```
 
-\pagebreak
-
 ```{r fasting, fig.cap = "Association between PTDS-5 and duration of fasting"}
-knitr::include_graphics("figures/fasting.png")
+# knitr::include_graphics("figures/fasting.png")
+drake::readd(fasting_duration_ptds_plot)
 ```
-
-\pagebreak
-
-## Appendix
-
-```{r tab.id='Hvaluesptds7', tab.cap.pre="Appendix Table ", tab.cap = "Homogeneity values (and standard errors) for all items in the PTDS"}
-drake::readd(H_values_ptds7_table) %>% 
-  flextable::flextable() %>% 
-  autofit()
-```
-
-\pagebreak
-
-```{r tab.id='monotonicity', tab.cap.pre="Appendix Table ", tab.cap="monotonicity with default minsize"}
-
-flextable::flextable(drake::readd(monotonicity_df_ptds6_table)) %>%  
-  autofit()
-```
-
-\pagebreak
-
-```{r tab.id='aisp', tab.cap.pre="Appendix Table ", tab.cap = "aisp for increasing H thresholds (*c*)"}
-flextable::flextable(drake::readd(aisp)) %>% 
-  autofit()
-```
-
-\pagebreak
-
-```{r tab.id='iioptds6', tab.cap.pre="Appendix Table ", tab.cap = "Invariant item ordering for PTDS-6"}
-drake::readd(iio_ptds6_table) %>% 
-  flextable() %>% 
-  autofit()
-```
-
-\pagebreak
-
-```{r tab.id='iioptds5', tab.cap.pre="Appendix Table ", tab.cap = "Invariant item ordering for PTDS-5"}
-drake::readd(iio_ptds5_table) %>%
-    flextable() %>%
-    autofit()
-```
-
-<!---BLOCK_LANDSCAPE_STOP--->

### 2020-08-12 manuscript/manuscript.tex
pdf version

\documentclass[letterpaper,9pt,twocolumn,twoside,]{pinp}

%% Some pieces required from the pandoc template
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

% pinp change: the geometry package layout settings need to be set here, not in pinp.cls
\geometry{layoutsize={0.95588\paperwidth,0.98864\paperheight},%
  layouthoffset=0.02206\paperwidth, layoutvoffset=0.00568\paperheight}

\definecolor{pinpblue}{HTML}{185FAF}  % imagecolorpicker on blue for new R logo
\definecolor{pnasbluetext}{RGB}{101,0,0} %

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{A Template for Two or One-Column Vignettes}

\author[]{}

\setcounter{secnumdepth}{0}

% Please give the surname of the lead author for the running footer
\leadauthor{Author and Author}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
 \keywords{  one |  two |  optional |  keywords |  here  }

\begin{abstract}
Your abstract will be typeset here, and used by default a visually
distinctive font. An abstract should explain to the general reader the
major contributions of the article.
\end{abstract}

\dates{This version was compiled on \today}

% initially we use doi so keep for backwards compatibility
% new name is doi_footer
\doifooter{\url{https://cran.r-project.org/package=YourPackage}}

\pinpfootercontents{Perioperative thirst discomfort scale}

\begin{document}

% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalised the article contents.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.

\acknow{This template package builds upon, and extends, the work of the
excellent \href{https://cran.r-project.org/package=rticles}{rticles}
package, and both packages rely on the
\href{http://www.pnas.org/site/authors/latex.xhtml}{PNAS LaTeX} macros.
Both these sources are gratefully acknowledged as this work would not
have been possible without them. Our extensions are under the same
respective licensing term
(\href{https://www.gnu.org/licenses/gpl-3.0.en.html}{GPL-3} and
\href{https://www.latex-project.org/lppl/}{LPPL (\textgreater= 1.3)}).}

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

\hypertarget{introduction}{%
\subsubsection{Introduction}\label{introduction}}

Symptoms of thirst discomfort are common in patients who are required to
undergo long periods of fasting prior to medical and surgical
procedures. The aim of this study was to examine the validity and
reliability of the perioperative thirst discomfort scale for measuring
thirst discomfort prior to procedures.

\hypertarget{methods}{%
\subsubsection{Methods}\label{methods}}

A prospective observational study was conducted. Patients who were
fasted and scheduled for an elective cardiac or interventional radiology
procedure were included. Mokken scaling analysis was conducted to
investigate the unidimensionality and hierarchical nature of the scale.
Correlations between perioperative thirst discomfort scale scores and
fasting durations were calculated to evaluate construct validity and
convergent validity was evaluated by comparing scores with global thirst
discomfort and thirst intensity ratings.

\hypertarget{results}{%
\subsubsection{Results}\label{results}}

Responses from 193 participants were included. The mean duration of
fasting from food and non-clear fluids was 12.7 (SD 3.8) hours and clear
fluids was 9 (SD 4.5) hours. Five items from the perioperative thirst
discomfort scale (PTDS-5) formed a Mokken scale with evidence of
invariant item ordering. The median PTDS-5 score was 3 (IQR 1, 6) from a
possible range of 0 to 10. The hierarchical pattern of responses to
PTDS-5 items is interpretable in terms of respondents most easily
endorsing an item related to discomfort associated with the desire to
drink water through to first endorsing symptoms related to dryness of
the mouth and lips before those related to the abnormal sensations of
`thick' saliva and a `thick' tongue. Scale reliability was adequate with
a coefficient score (\emph{rho} =0.84). There was a positive correlation
between PTDS-5 scores and the global thirst discomfort rating
(\emph{tau} = 0.54; 95\% CI = 0.43 to 0.63), as well as thirst intensity
(\emph{tau} = 0.49; 95\% CI = 0.38 to 0.59). Duration of fasting was not
associated with PTDS-5 scores.

\hypertarget{conclusion}{%
\subsubsection{Conclusion}\label{conclusion}}

The items in the PTDS-5 form a strong Mokken scale, meaning it is a
reliable and precise way to order patients according to their thirst
discomfort. Nurses may use this simple scale to assess the severity of
thirst discomfort in patients who are fasted prior to procedures and
initiate interventions according to individual needs.

Keywords: fasting, aspiration, patient safety, thirst, anesthesia,
nursing, Mokken scale, scale development, item response theory

\hypertarget{declarations}{%
\subsection{Declarations}\label{declarations}}

Availability of data and material (data transparency): All data used is
available \href{https://github.com/awconway/ptds}{here}.

Code availability: Code to reproduce the analyses is available
\href{https://github.com/awconway/ptds}{here}.

Abbreviated title: Perioperative thirst discomfort scale

Funding statement: Support was provided solely from institutional and/or
departmental sources. MP is supported by an Early Investigator Merit
Award from the Department of Anesthesia, University of Toronto and
Toronto General Hospital.

Conflicts of interest: The authors declare no competing interests.

Author contributions: AC: Study design, data analysis and wrote
manuscript; KC: Study design, revised the manuscript for important
intellectual content; MB: Data collection, revised the manuscript for
important intellectual content; DP: Data collection, revised the
manuscript for important intellectual content; NK: Study design, revised
the manuscript for important intellectual content; AM: Study design,
revised the manuscript for important intellectual content; MP: Study
design, revised the manuscript for important intellectual content; SM:
Study design, revised the manuscript for important intellectual content;
SM: Study design, revised the manuscript for important intellectual
content; PG: Study design, revised the manuscript for important
intellectual content; SA: Study design, revised the manuscript for
important intellectual content.

\hypertarget{introduction-1}{%
\subsection{Introduction}\label{introduction-1}}

Pre-procedure fasting is used to reduce the risk of vomiting and
aspiration pneumonia during sedation and general anaesthesia, or in case
emergency intubation is required due to unexpected cardiac
arrest.\citep{hamid2014pre, osborne2002preoperative} However, prolonged
fluid restriction causes thirst symptoms to develop (e.g., dry mouth,
swollen tongue), which can lead to great
discomfort.\citep{madsen1998perioperative} Current guidelines related to
pre-procedure fasting for elective procedures recommend a minimum
fasting period of 2 hours nil-per-os (NPO) for clear
fluids.\citep{dobsonGuidelinesPracticeAnesthesia2018} Despite these
recommendations, current practice is for patients undergoing surgical
and other medical procedures that require sedation or anesthesia to
receive standardized `nil-by-mouth' fasting instructions at a
pre-specified time interval before procedures. For example, `no eating
or drinking after midnight' is most common. This `standardized'
instruction will not be changed regardless of whether or not there are
alterations in scheduling throughout the day that result in significant
delays in procedure start time. As a result, fasting durations far
exceed the recommended requirement for most patients undergoing medical
and surgical
procedures.\citep{de2014actual, sorita2015frequency, spitz2017impact}
For example, in a recent study of 3641 fasting orders at a large
academic institution in the USA, it was found that the median fasting
duration was 12.8 hours, averaging 2 missed
meals.\citep{sorita2015frequency}

As a direct result of prolonged pre-procedure fasting, symptoms of
thirst discomfort have been reported as common and severe. In a
qualitative study where 12 participants were interviewed from a tertiary
hospital in Australia, surgical patients and patients who adhered to
prolonged fasting instructions described the discomfort from thirst
symptoms to be the worst physical effect of
fasting.\citep{carey2015qualitative} Similarly, Madsen et
al.~interviewed a convenience sample of 50 adult surgical patients who
reported that thirst symptoms caused more discomfort than hunger, sleep,
or anxiety related to the procedure.\citep{madsen1998perioperative}
However, thirst and its symptoms continue to be undervalued,
under-reported and infrequently assessed by health care providers,
including the nursing team.\citep{milani2016thirst}

Despite the relevance and value of assessing thirst-discomfort of
patients, the subjective experience of thirst presents challenges in
developing a valid and reliable tool to succinctly and accurately
measure its symptoms and level of discomfort prior to procedures. A
thirst-discomfort scale for perioperative use was developed at a
surgical centre affiliated with an accredited public hospital in
Brazil.\citep{Martins_2017} The seven-item perioperative
thirst-discomfort scale (PTDS) was developed in three stages, including
face and content validation and reliability, and based on the
Consensus-based Standards for the selection of health status Measurement
Instruments (COSMIN) checklist.\citep{mokkink2010cosmin} Inter-rater
reliability was tested through inter-observer equivalence where a pair
of nurses independently, but simultaneously administered the scale among
70 patients. Six items on the scale had a weighted kappa coefficient of
1, while the item ``I feel like drinking water'' was
0.97.\citep{Martins_2017} Cronbach's alpha was used to test the internal
consistency of the scale with a value of 0.91.\citep{Martins_2017}

Although the initial validation of the PTDS was conducted in a different
context (i.e.~\emph{post}-surgical patients), it demonstrates strong
potential for accurately assessing thirst-discomfort more generally for
peri-operative use. The aim of this study was to examine the validity
and reliability of the PTDS for measuring thirst discomfort for patients
who are fasting \emph{before} medical procedures.

\hypertarget{methods-1}{%
\subsection{Methods}\label{methods-1}}

\hypertarget{study-design}{%
\subsubsection{Study Design}\label{study-design}}

This study used a prospective, observational design. No changes to usual
clinical practice were made for this study in regard to pre-procedure
fasting. Patients who were scheduled for a morning procedure were
typically asked to remain nil-per-os (NPO) from midnight. Patients with
afternoon procedures can have a small breakfast but must remain NPO from
6:00am on the day of the procedure. Ethical approval for this study
(Ethical Committee N° 19-5585) was provided by the University Health
Network Research Ethics Board, Toronto, Canada on October 31 2019.

\hypertarget{participants}{%
\subsubsection{Participants}\label{participants}}

\hypertarget{inclusion-criteria}{%
\paragraph{Inclusion criteria}\label{inclusion-criteria}}

Adult patients who were fasted and scheduled for an elective procedure
in the Cardiac Catheterization Laboratories or in Interventional
Radiology at an academic teaching hospital in Canada were included.

\hypertarget{exclusion-criteria}{%
\paragraph{Exclusion criteria}\label{exclusion-criteria}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Under 16 years of age
\item
  Undergoing an emergency procedure
\item
  Unable to understand or speak English and a translator is not
  immediately available
\item
  Nurse in pre-procedural bay considered that there was insufficient
  time prior to anticipated commencement of the procedure for
  participation in the study.
\end{enumerate}

\hypertarget{data-collection}{%
\subsubsection{Data collection}\label{data-collection}}

A Research Assistant administered a brief questionnaire prior to
procedures. It comprised the following components:

\begin{itemize}
\tightlist
\item
  the Perioperative Thirst Discomfort Scale (7 items);
\item
  a one-item global thirst discomfort rating;
\item
  a one-item thirst intensity rating;
\item
  an item to determine if the participant is currently on oxygen
  therapy;
\item
  an item to evaluate the presence of pain;
\item
  the time the patient last had any clear fluids;
\item
  the time the patient last had any food;
\item
  age, sex of patient and type of procedure to be performed.
\end{itemize}

\hypertarget{measures}{%
\paragraph{Measures}\label{measures}}

The seven-item Perioperative Thirst Discomfort Scale (PTDS-7) is a
7-item self-reported, composite score evaluating the severity of dry
mouth, lips, and throat, thick tongue and saliva, a bad taste in the
mouth, and a desire to drink water. The total score can range from 0 to
14, with 14 representing the ``most extreme'' intensity of discomfort
related to perioperative thirst. This instrument was developed and
validated for perioperative use in a surgical centre at a public
hospital in Brazil. Face validation was conducted by group consisting of
nurses with PhDs, new graduate nurses, and patients. Content validation
was conducted by a team consisting of specialists with perioperative
nursing experience related to thirst or instrument validation. The
reliability of the scale was assessed through inter-observer equivalence
by a research nurse and a staff nurse from the Post Anesthesia Care Unit
among 70 patients pre and postoperatively in an anesthetic recovery
room. The Cronbach alpha score was 0.91. The time taken to complete the
PTDS-7 was not evaluated.

Participants used a global thirst discomfort score to rate their overall
level of thirst discomfort on a rating scale with scores ranging from
``extreme thirst discomfort'' to ``extreme comfort''. Intensity of their
thirst was rated on a scale ranging from 0 to 10, where 0 represents
``No thirst'' and 10 represents ``Most intense thirst''. The experience
of pain may influence the participant's perception of
thirst.\citep{Pierotti_2018} Participants were asked to rate their
current level of pain on a scale of 0 to 10, where 0 represents ``no
pain'' and 10 represents that their pain as
``unimaginable/unspeakable''. The use of oxygen therapy (e.g., via nasal
prongs) may contribute to the experience of thirst symptoms (e.g., dry
mouth, dry throat).\citep{conchon2015perioperative} The research
assistant recorded ``yes'' or ``no'' as to whether the participant was
currently receiving oxygen therapy at the time the questionnaire was
administered.

\hypertarget{statistical-analysis}{%
\subsubsection{Statistical analysis}\label{statistical-analysis}}

Mokken scaling is a method for establishing if items in a scale conform
to a cumulative, hierarchical structure.\citep{perng2012construct} When
an instrument contains hierarchically ordered items, the items can be
ordered from those indicating mild to severe symptoms. An intentionally
simple example of a hierarchical scale is a set of items enquiring about
height that increases in increments of 10cm from 100cm to 200cm. In this
example, taller people would endorse more items than those who were
shorter.\citep{watson2008hierarchy}

To determine if items conform to a hierarchical structure, Mokken
scaling analyses properties of individual items as described by the item
characteristic curve. Item characteristic curves show the association
between the score on an item to the level of the latent trait being
measured. Mokken scaling makes only the assumption of monotone
homogeneity about the association between the score on an item and the
level of the latent trait. This means that as the trait increases, so
does the item score. ``Difficulty'', is a term used in Mokken Scale
Analysis to refer to the extent to which items are endorsed by
respondents. Items at the upper end of the range of the latent trait are
characterized as being more difficult. Items in a Mokken Scale are
arranged along the latent trait in terms of their difficulty. The
properties of items can be measured using the scalability coefficient,
\emph{H} (Loevinger's coefficient). \emph{H} measures the extent to
which items are arranged as expected by their mean values along the
latent trait. \emph{H} \textgreater{} 0.3 is the minimum acceptable
value, with values \textgreater{} 0.4 indicating a moderate scale and
\textgreater{} 0.5 indicates a strong scale.

Invariant item ordering, where the order of items along the latent trait
is the same for all respondents at all levels of the latent trait, is a
desirable property of a Mokken Scale. It can be assessed mathematically
to look for significant violations of this property. The accuracy of the
invariant item ordering can also be assessed by calculating the
\emph{H\textsubscript{TRANS}} (\textbf{H\textsubscript{T}}) coefficient.
Values of \textbf{H\textsubscript{T}} exceeding 0.3 indicates acceptable
accuracy of invariant item ordering.

Mokken scale analysis was conducted in this study to explore whether
there were hierarchical properties in participant ratings of thirst
discomfort and to explore the dimensions of the PTDS. Mokken scale
analysis proceeded by first checking the PTDS-7 scalability
coefficients. If \textbf{H\textsubscript{i}} was below 0.3 or if the
lower limit of the 95\% CI (confidence interval) for
\emph{H\textsubscript{i}} was below 0.3, the item was to be excluded.
Then, scale partitioning was carried out to explore the dimensions of
the PTDS through increasing c (lower bound c defines the minimum value
of coefficients \textbf{H\textsubscript{i}} in the Mokken scale by 0.05
increments).\citep{molenaar2000mps5} Monotone homogeneity and invariant
item ordering were investigated at the whole scale level, as no
sub-scales were identified. To assess for violations of invariant item
ordering, crit values \textless40 were considered
acceptable.\citep{molenaar2000mps5} Provided invariant item ordering
holds for a set of items, the coefficient \textbf{H\textsubscript{T}}
expresses the accuracy of the ordering, with values below 0.3 being
unacceptable and above 0.4 indicating moderate invariant item
ordering.\citep{Ligtvoet_2010} The coefficient score (rho), which is
similar to Cronbach's alpha, was used to assess reliability of the
scale.

Construct validity was evaluated by using correlations to identify
associations between scores on the PTDS and fasting duration. To
investigate construct validity we hypothesized that higher scores on the
PTDS would be associated with greater fasting duration. Convergent
validity was assessed by using Kendall's \emph{tau} to compare the PTDS
score with the global thirst discomfort score. Convergent validity was
assessed by using the Spearman's rank correlation to compare the PTDS
with the thirst intensity rating and pain scale rating.

\hypertarget{results-1}{%
\subsection{Results}\label{results-1}}

From November 2019 to March 2020, we screened 203 patients for inclusion
in the study. A total of 198 were eligible and 194 participated. One
participant did not complete 1 of the PTDS-7 items and was excluded from
the analysis. A summary of participant characteristics is displayed in
Table @ref(tab:tab1). The median age was 62 (IQR 48 - 72) and 58\%
(n=111) were female. 43\% of patients were scheduled to have their
procedure in the cardiac catheterization laboratory. A biopsy was the
most common interventional radiology procedure (n=62; 32\%), but a wide
range of other procedures were included. The mean duration of fasting
from food and non-clear fluids was 12.7 (SD 3.8) hours and clear fluids
was 9 (SD 4.5) hours.

\hypertarget{mokken-scale-analysis}{%
\subsubsection{Mokken scale analysis}\label{mokken-scale-analysis}}

No items required removal from the PTDS-7 based on the condition that
scalability coefficients (\emph{H\textsubscript{i}} and lower bound of
95\% CI) should be over 0.3 (Appendix Table @ref(tab:Hvaluesptds7)). We
used a lower bound for \emph{c}, starting from 0.05 and increasing to
0.80 in 0.05 increments, to explore the dimensionality of the PTDS-7.
From 0.05 to 0.45, all of the items formed a single scale. The item,
`\emph{I have a bad taste in my mouth}', was excluded by selecting the
remaining items which showed uni-dimensionality at a threshold level of
.50. The 6 remaining items (PTDS-6) were further examined for the
criteria of a Mokken scale and item invariant ordering. The PTDS-6 item
set has a homogeneity value \emph{H(se)} of 0.599, (0.034), which
indicated a strong Mokken scale. Results indicated that there were no
significant violations of the assumption of monotonicity (z-sig). As
such, no items were removed based on this criteria.

\hypertarget{invariant-item-ordering}{%
\paragraph{Invariant item ordering}\label{invariant-item-ordering}}

Analysis of invariant item ordering for the PTDS-6 revealed violations
for the dry throat/dry lips item pair (Appendix Table
@ref(tab:iioptds6)). Although there were no \emph{significant}
intersections between items, due to the intersection between the dry
throat and dry lips items, crit values were above 40. For this reason,
we excluded one of these items (dry throat), and re-assessed the
remaining 5 items for the assumptions of a Mokken scale. With the dry
throat item removed, none of the items violated the assumption of
invariant item ordering (Appendix Table @ref(tab:iioptds5)).

\hypertarget{accuracy-and-reliability}{%
\paragraph{Accuracy and reliability}\label{accuracy-and-reliability}}

The accuracy of the item ordering was found to be acceptable with a
\emph{H\textsubscript{T}} coefficient of 0.397. Scale reliability was
adequate with a coefficient score (Rho) of 0.84.

\hypertarget{item-difficulty-hierarchy}{%
\paragraph{Item difficulty hierarchy}\label{item-difficulty-hierarchy}}

The item difficult hierarchy for the PTDS-5 is presented in Table
@ref(tab:tab2). The most highly endorsed symptom (i.e.~the least severe
symptom of thirst discomfort in the PTDS-5) was being bothered by a
desire to drink water. The least endorsed symptom (i.e.~the most severe
symptom of thirst discomfort in the PTDS-5) was being bothered by the
perception of a `thick tongue'. The items in the PTDS-5 form a strong
Mokken scale (H\textsubscript{s} = 0.605).

\begin{ShadedResult}
\begin{verbatim}
#                     item
#  1 I want to drink water
#  2       My mouth is dry
#  3       My lips are dry
#  4    My saliva is thick
#  5    My tongue is thick
\end{verbatim}
\end{ShadedResult}

\hypertarget{construct-and-convergent-validity}{%
\subsubsection{Construct and convergent
validity}\label{construct-and-convergent-validity}}

The median PTDS-5 score was 3 (IQR 1, 6). The distribution of PTDS-5
scores was right-skewed (Figure @ref(fig:distribution)). There was a
positive correlation between PTDS-5 scores and thirst intensity (Figure
@ref(fig:intensity)), as well as the global thirst discomfort rating
(Figure @ref(fig:globalthirst)). The associations between PTDS-5 scores
with food and fluid duration is displayed in Figure @ref(fig:fasting).
Longer duration of fasting was not linearly associated with PTDS-5
scores. Likewise, the correlation between pain and PTDS-5 scores was not
significant (Spearman's rho = 0.07; 95\% CI = -0.07 to 0.21).

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

Results indicate that the PTDS-5 is a reliable and precise way to order
patients according to their thirst discomfort in the
\emph{pre-procedure} period. In addition, an acceptable accuracy for
invariant item ordering was identified, meaning patients are likely to
rank the PTDS-5 symptoms of thirst discomfort in the same way. As such,
the item difficulty hierarchy, presented in Table @ref(tab:tab2) can be
used as a guide to easily determine which patients are expriencing worse
thirst discomfort. The hierarchical pattern of responses to items in the
Mokken scale is interpretable in terms of respondents more easily
endorsing items related to a general desire to drink water through to
first endorsing more specific symptoms related to dryness of the mouth
and lips prior to endorsing discomfort associated with the sensations of
`thick' saliva and `thick' tongue. In more practical terms, a patient
who reports feeling bothered by a sensation that saliva is thick, could
be interpreted as experiencing worse thirst discomfort than a patient
who did not endorse any of the items lower down on the item difficulty
hierarchy (i.e.~`my lips are dry' and `my mouth is dry').

The original (7-item) perioperative thirst-discomfort scale has
previously been used to determine the relationship between thirst
intensity and thirst-discomfort in the immediate postoperative period
during anesthesia recovery.\citep{Pierotti_2018} In a non-probabilistic
sample of 203 adult participants in post-operative recovery following
elective, urgent and emergency surgeries, the investigators found no
association between fasting duration, or patient age, and the intensity
of thirst and degree of discomfort.\citep{Pierotti_2018} However, the
mean fasting duration among participants was uniformly high (mean 16.2
{[}SD 8.7{]} hours), which may have prevented the researchers from
discerning a correlation between fasting duration. Even in the subgroup
of participant who reported 8 hours of fasting, all reported
thirst.\citep{aroni2012assessment, Pierotti_2018}

It should be noted that an investigation of the dimensionality of the
7-item PTDS had not been performed prior to our study. A uni-dimensional
solution was identified in our analysis. This should be expected
considering the small number of items that were evaluated.

We did not observe an association between PTDS-5 and fasting duration
for either food or fluids. This would suggest that it is important to
assess thirst discomfort periodically, regardless of the duration of
fasting. If thirst discomfort is present, and fasting is still
indicated, interventions to alleviate symptoms should be implemented.
There is a lack of evidence regarding the efficacy of any such
interventions.

\hypertarget{limitations}{%
\subsubsection{Limitations}\label{limitations}}

Selection bias should be considered because a convenience sample was
used for this study. Also, it is uncertain if similar results would be
obtained for patients who are fasted prior to undergoing procedures in
other settings. Only one patient was excluded from the analysis due to
missing data, so it is unlikely attrition bias would have exerted a
major influence on the results. Further research is required to evaluate
the minimal detectable difference for the PTDS-5 as well as the minimal
clinically important difference prior to use of this scale as an outcome
measure in research.

\hypertarget{conclusion-1}{%
\subsubsection{Conclusion}\label{conclusion-1}}

The items in the PTDS-5 form a reliable and precise way to order
patients according to their thirst discomfort. Nurses may use this scale
to assess the severity of thirst discomfort in patients who are fasted
prior to procedures and initiate interventions according to individual
needs.

\begin{tabular}{ll}
\toprule
Characteristic & N = 193\\
\midrule
Age & 62 (48, 72)\\
Gender & \\
\hspace{1em}Male & 111 (58\%)\\
\hspace{1em}Female & 80 (42\%)\\
\hspace{1em}Prefer not to say & 1 (0.5\%)\\
\addlinespace
Procedure & \\
\hspace{1em}Angiogram or Percutaneous Coronary Intervention (PCI) & 58 (30\%)\\
\hspace{1em}Cardiac Implantable Electronic Device (CIED) & 15 (7.8\%)\\
\hspace{1em}Electrophysiology Study (EPS) & 5 (2.6\%)\\
\hspace{1em}Structural heart intervention & 5 (2.6\%)\\
\addlinespace
\hspace{1em}Vascular angiography or intervention & 17 (8.9\%)\\
\hspace{1em}Biopsy & 62 (32\%)\\
\hspace{1em}Port-a-cath & 4 (2.1\%)\\
\hspace{1em}Gall bladder stone removal & 0 (0\%)\\
\hspace{1em}Endovascular aneurysm repair (EVAR) & 0 (0\%)\\
\addlinespace
\hspace{1em}Radiofrequency ablation of a tumour (RFA) & 9 (4.7\%)\\
\hspace{1em}Fistula repair/de-clotting & 0 (0\%)\\
\hspace{1em}Vascular embolization (e.g., renal) & 1 (0.5\%)\\
\hspace{1em}Other procedure & 16 (8.3\%)\\
Time since last clear fluids (hours) & 9.5 (5.0, 13.0)\\
\addlinespace
Time since last food (hours) & 13.0 (11.0, 15.0)\\
\bottomrule
\multicolumn{2}{l}{\textsuperscript{1} Statistics presented: median (IQR); n (\%)}\\
\end{tabular}

\begin{figure}

{\centering \includegraphics[width=252px]{manuscript_files/figure-latex/distribution-1}

}

\caption{Distribution of PTDS-5 scores}\label{fig:distribution}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=252px]{manuscript_files/figure-latex/intensity-1}

}

\caption{Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 0 to 10 with higher scores indicating worse thirst discomfort.}\label{fig:intensity}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=252px]{manuscript_files/figure-latex/globalthirst-1}

}

\caption{Association between PTDS-5 and global thirst rating}\label{fig:globalthirst}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=252px]{manuscript_files/figure-latex/fasting-1}

}

\caption{Association between PTDS-5 and duration of fasting}\label{fig:fasting}
\end{figure}

%\showmatmethods
\showacknow

\bibliography{references}
\bibliographystyle{jss}

\end{document}

### 2020-08-12 R/create_distribution_plot.r

@@ -6,7 +6,7 @@
 create_distribution_plot <- function(data_ptds) {
     distribution_plot <- data_ptds %>%
         ggplot(aes(x = ptds)) +
-        geom_histogram(binwidth = 0.5, colour = "#2a6ebb", fill = "#2a6ebb") +
+        geom_histogram(binwidth = 0.5, fill = "#2a6ebb") +
         labs(
             y = "Count",
             x = "\nPTDS-5 (range is 0 to 10; higher scores indicate worse thirst discomfort)"
@@ -14,11 +14,12 @@ create_distribution_plot <- function(data_ptds) {
         scale_x_continuous(
             breaks = seq(0, 10, by = 1)
         ) +
+      theme_minimal()+
         theme(
-            panel.background = element_blank(),
+            #panel.background = element_blank(),
             legend.position = "none",
             axis.ticks.x = element_blank(),
-            axis.ticks.y = element_blank(),
-            text = element_text(family = "BioRhyme")
+            axis.ticks.y = element_blank()
+            #text = element_text(family = "BioRhyme")
         )
 }

### 2020-08-12 R/create_summary_table.R

@@ -15,6 +15,5 @@ create_summary_table <- function(data_ptds) {
         fluids_duration ~ "Time since last clear fluids (hours)",
         food_duration ~ "Time since last food (hours)"
       )
-    ) %>%
-    italicize_labels()
+    )
 }

### 2020-08-12 R/plan.R

@@ -83,6 +83,13 @@ get_analysis_plan <- function() {
       width = 174,
       units = "mm"
     ),
+    distribution_plot_pdf = ggsave(
+      plot = distribution_plot,
+      device = "pdf",
+      filename = here("manuscript/figures/distribution.pdf"),
+      width = 174,
+      units = "mm"
+    ),
 
     # correlations
 
@@ -137,11 +144,17 @@ get_analysis_plan <- function() {
     #  device = "svg"
     #),
 
-    manuscript = target(
+    # manuscript = target(
+    #   command = {
+    #     rmarkdown::render(knitr_in("manuscript/index.Rmd"))
+    #     file_out("manuscript/index.docx")
+    #   }
+    # ),
+    manuscript_pdf = target(
       command = {
         rmarkdown::render(knitr_in("manuscript/index.Rmd"))
-        file_out("manuscript/index.docx")
+        file_out("manuscript/index.pdf")
       }
-    ),
+    )
   )
 }

### 2020-08-12 manuscript/manuscript.Rmd
pdf

@@
 ---
-title: A Template for Two or One-Column Vignettes
+title: Mokken scale analysis of the perioperative thirst discomfort scale
 
@@
 
-# Optional: line of arbitrary text with additional information.
+date_subtitle: "Blinded for peer review"
 # Could be used, for example, to mention the bibliographic info in a post-print.
 # If not specified, defaults to "This version was compiled on \today"
-#date_subtitle: Published in *Journal of Statistical Software*, 2018
 
 # For footer text  TODO(fold into template, allow free form two-authors)
-lead_author_surname: Author and Author
+lead_author_surname: Conway et al.
 
 # Place eg a DOI URL or CRAN Package URL here
-doi_footer: "https://cran.r-project.org/package=YourPackage"
+doi_footer: "Data and statistical code is available [here](https://github.com/awconway/ptds)"
 
@@
 abstract: |
-  Your abstract will be typeset here, and used by default a visually distinctive font.
-  An abstract should explain to the general reader the major contributions of the article.
-
-# Optional: Acknowledgements
-acknowledgements: |
-  This template package builds upon, and extends, the work of the excellent
-  [rticles](https://cran.r-project.org/package=rticles) package, and both packages rely on the
-  [PNAS LaTeX](http://www.pnas.org/site/authors/latex.xhtml) macros. Both these sources are
-  gratefully acknowledged as this work would not have been possible without them.  Our extensions
-  are under the same respective licensing term
-  ([GPL-3](https://www.gnu.org/licenses/gpl-3.0.en.html) and
-  [LPPL (>= 1.3)](https://www.latex-project.org/lppl/)).
+   *Introduction:* Symptoms of thirst discomfort are common in patients who are required to undergo long periods of fasting prior to medical and surgical procedures.
+The aim of this study was to examine the validity and reliability of the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures.
+   *Methods:* A prospective observational study was conducted.
+Patients who were fasted and scheduled for an elective cardiac or interventional radiology procedure were included.
+Mokken scaling analysis was conducted to investigate the unidimensionality and hierarchical nature of the scale.
+Correlations between perioperative thirst discomfort scale scores and fasting durations were calculated to evaluate construct validity and convergent validity was evaluated by comparing scores with global thirst discomfort and thirst intensity ratings.
+   *Results:* 
+   *Conclusion:* The items in the PTDS-5 form a strong Mokken scale, meaning it is a reliable and precise way to order patients according to their thirst discomfort.
+Nurses may use this simple scale to assess the severity of thirst discomfort in patients who are fasted prior to procedures and initiate interventions according to individual needs.
 
@@
 keywords:
-  - one
-  - two
-  - optional
-  - keywords
-  - here
-
-# Paper size for the document, values of letter and a4
-papersize: letter
-
-# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
-fontsize: 9pt
-
-# Optional: Force one-column layout, default is two-column
-#one_column: true
-
-# Optional: Enables lineno mode, but only if one_column mode is also true
-#lineno: true
-
-# Optional: Enable one-sided layout, default is two-sided
-#one_sided: true
-
-# Optional: Enable section numbering, default is unnumbered
-#numbersections: true
-
-# Optional: Specify the depth of section number, default is 5
-#secnumdepth: 5
-
-# Optional: Skip inserting final break between acknowledgements, default is false
-skip_final_break: true
+  - fasting
+  - anesthesia
+  - thirst discomfort
+  - patient-reported outcomes
+  - scale validation
+  - mokken scale analysis
 
@@
 bibliography: references
-
-# Optional: Enable a 'Draft' watermark on the document
-#watermark: true
 
@@
 output: pinp::pinp
-
+params: 
+  word: false
 ---
@@
 
+```{r eval=params$word, results='asis'}
+cat(
+"
 ## Abstract
@@
 
-## Introduction
+"
+)
+```
+
+# Introduction
 
@@
 
-## Methods
+# Methods
 
-### Study Design
+## Study Design
 
@@
 
-### Participants
+## Participants
 
-#### Inclusion criteria
+### Inclusion criteria
 
@@
 
-#### Exclusion criteria
+### Exclusion criteria
 
@@
 
-### Data collection
+## Data collection
 
@@
 
-#### Measures
+### Measures
 
@@
 
-### Statistical analysis
+## Statistical analysis
 
@@
 
-## Results
+# Results
 
@@
 
-### Mokken scale analysis
+```{r tab.cap="Participant characteristics", tab.id = 'tab1'}
+drake::loadd(summary_table)
+# gtsummary::as_flextable(summary_table)
+gtsummary::as_kable_extra(summary_table, booktabs = T)
+```
+
+## Mokken scale analysis
 
@@
 
-#### Invariant item ordering
+### Invariant item ordering
 
@@
 
-#### Accuracy and reliability
+### Accuracy and reliability
 
@@
 
-#### Item difficulty hierarchy
+### Item difficulty hierarchy
 
@@
 
-```{r }
+```{r tab.id = "tab2", tab.cap= "Item hierarchy of the PTDS-5"}
 drake::readd(item_hierarchy_df_ptds5) %>% 
-  dplyr::select(item) 
+  kableExtra::kable(booktabs=T)
 ```
 
-### Construct and convergent validity
+## Construct and convergent validity
 
-The median PTDS-5 score was `r ptds5`.
-The distribution of PTDS-5 scores was right-skewed (Figure \@ref(fig:distribution)).
+The median PTDS-5 score was `r ptds5`. The distribution of PTDS-5 scores was right-skewed (Figure \@ref(fig:distribution)).
+
+```{r distribution, fig.cap = "Distribution of PTDS-5 scores (higher scores indicate worse thirst discomfort)"}
+# knitr::include_graphics("figures/distribution.png")
+drake::readd(distribution_plot)
+```
+
 There was a positive correlation between PTDS-5 scores and thirst intensity (Figure \@ref(fig:intensity)), as well as the global thirst discomfort rating (Figure \@ref(fig:globalthirst)).
 The associations between PTDS-5 scores with food and fluid duration is displayed in Figure \@ref(fig:fasting).
+
+```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 0 to 10 with higher scores indicating worse thirst discomfort."}
+# knitr::include_graphics("figures/thirst-intensity.png")
+drake::readd(thirst_intensity_plot)
+```
+
+```{r globalthirst, fig.cap="Association between PTDS-5 and global thirst rating", eval=params}
+# knitr::include_graphics("figures/global-thirst.png")
+library(dplyr)
+drake::readd(global_thirst_plot)
+```
+
 Longer duration of fasting was not linearly associated with PTDS-5 scores.
@@
 
-## Discussion
+```{r fasting, fig.cap = "Association between PTDS-5 and duration of fasting"}
+# knitr::include_graphics("figures/fasting.png")
+drake::readd(fasting_duration_ptds_plot)
+```
+
+# Discussion
 
@@
 In addition, an acceptable accuracy for invariant item ordering was identified, meaning patients are likely to rank the PTDS-5 symptoms of thirst discomfort in the same way.
-As such, the item difficulty hierarchy, presented in Table \@ref(tab:tab2) can be used as a guide to easily determine which patients are expriencing worse thirst discomfort.
+As such, the item difficulty hierarchy, presented in Table \@ref(tab:tab2) can be used as a guide to easily determine which patients are experiencing worse thirst discomfort.
 The hierarchical pattern of responses to items in the Mokken scale is interpretable in terms of respondents more easily endorsing items related to a general desire to drink water through to first endorsing more specific symptoms related to dryness of the mouth and lips prior to endorsing discomfort associated with the sensations of 'thick' saliva and 'thick' tongue.
@@
 
-### Limitations
+## Limitations
 
@@
 
-### Conclusion
+## Conclusion
 
@@
 
-```{r tab.cap="Participant characteristics", tab.id = 'tab1'}
-drake::loadd(summary_table)
-# gtsummary::as_flextable(summary_table)
-gtsummary::as_kable_extra(summary_table, booktabs = T)
-```
-
 <!---BLOCK_MULTICOL_START--->
@@
 <!--BLOCK_LANDSCAPE_START--->
-
-```{r distribution, fig.cap = "Distribution of PTDS-5 scores"}
-# knitr::include_graphics("figures/distribution.png")
-drake::readd(distribution_plot)
-```
-
-```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 0 to 10 with higher scores indicating worse thirst discomfort."}
-# knitr::include_graphics("figures/thirst-intensity.png")
-drake::readd(thirst_intensity_plot)
-```
-
-```{r globalthirst, fig.cap="Association between PTDS-5 and global thirst rating"}
-# knitr::include_graphics("figures/global-thirst.png")
-library(dplyr)
-drake::readd(global_thirst_plot)
-```
-
-```{r fasting, fig.cap = "Association between PTDS-5 and duration of fasting"}
-# knitr::include_graphics("figures/fasting.png")
-drake::readd(fasting_duration_ptds_plot)
-```

### 2020-08-12 R/create_distribution_plot.r

@@ -9,7 +9,7 @@ create_distribution_plot <- function(data_ptds) {
         geom_histogram(binwidth = 0.5, fill = "#2a6ebb") +
         labs(
             y = "Count",
-            x = "\nPTDS-5 (range is 0 to 10; higher scores indicate worse thirst discomfort)"
+            x = "\nPTDS-5"
         ) +
         scale_x_continuous(
             breaks = seq(0, 10, by = 1)
@@ -19,7 +19,7 @@ create_distribution_plot <- function(data_ptds) {
             #panel.background = element_blank(),
             legend.position = "none",
             axis.ticks.x = element_blank(),
-            axis.ticks.y = element_blank()
-            #text = element_text(family = "BioRhyme")
+            axis.ticks.y = element_blank(),
+            text = element_text(size=9)
         )
 }

### 2020-08-12 R/create_fasting_duration_ptds_plot.R

@@ -13,18 +13,18 @@
 create_fasting_duration_ptds_plot <- function(data = data_ptds,
                                               fluids_corr, food_corr) {
 
-  rho <- round(fluids_corr$rho, 2)
-  rho_low <- round(fluids_corr$CI_low, 2)
-  rho_high <- round(fluids_corr$CI_high, 2)
-
-  ptds_fluids_cor <- tibble(
-  # location of each text label in data coordinates
-  ptds = 10,
-  fluids_duration = 20,
-  # text label containing rho value
-  label = glue("*rho* = {rho}
-      (95% CI = {rho_low} to {rho_high})")
-  )
+  # rho <- round(fluids_corr$rho, 2)
+  # rho_low <- round(fluids_corr$CI_low, 2)
+  # rho_high <- round(fluids_corr$CI_high, 2)
+  #
+  # ptds_fluids_cor <- tibble(
+  # # location of each text label in data coordinates
+  # ptds = 10,
+  # fluids_duration = 20,
+  # # text label containing rho value
+  # label = glue("*rho* = {rho}
+  #     (95% CI = {rho_low} to {rho_high})")
+  # )
 
   fluids <- data %>%
     ggplot(aes(y = ptds, x = fluids_duration)) +
@@ -39,30 +39,30 @@ create_fasting_duration_ptds_plot <- function(data = data_ptds,
     theme(
       legend.position = "none",
       plot.title.position = "plot",
-      axis.title.x = element_markdown()
-
+      axis.title.x = element_markdown(),
+      text = element_text(size=9)
     ) +
-    ggtext::geom_richtext(
-      data = ptds_fluids_cor,
-      aes(label = label),
-      hjust = 1, vjust = 2
-    ) +
+    # ggtext::geom_richtext(
+    #   data = ptds_fluids_cor,
+    #   aes(label = label),
+    #   hjust = 1, vjust = 2
+    # ) +
     scale_y_continuous(
       breaks = seq(0, 10, by = 2)
     )
 
-  rho <- round(food_corr$rho, 2)
-  rho_low <- round(food_corr$CI_low, 2)
-  rho_high <- round(food_corr$CI_high, 2)
-
-  ptds_food_cor <- tibble(
-  # location of each text label in data coordinates
-  ptds = 10,
-  food_duration = 25,
-  # text label containing rho value
-  label = glue("*rho* = {rho}
-      (95% CI = {rho_low} to {rho_high})")
-  )
[…]

### 2020-08-12 R/create_global_thirst_plot.R

@@ -31,11 +31,12 @@ create_global_thirst_plot <- function(data,
     geom_beeswarm(groupOnX = FALSE, alpha = 0.5, size = 2) +
     labs(
       y = element_blank(),
-      x = "\nPTDS-5 (0-10; higher scores indicate worse thirst discomfort)"
+      x = "\nPTDS-5"
     ) +
     theme(
       legend.position = "none",
-      plot.title.position = "plot"
+      plot.title.position = "plot",
+      text = element_text(size=9)
     ) +
     geom_richtext(
       data = cor,

### 2020-08-12 R/plan.R

@@ -49,6 +49,11 @@ get_analysis_plan <- function() {
       device = "png", filename = here("manuscript/figures/global-thirst.png"),
       width = 174, units = "mm"
     ),
+    global_thirst_plot_pdf = ggsave(
+      plot = global_thirst_plot,
+      device = "pdf", filename = here("manuscript/figures/global-thirst.pdf"),
+      width = 174, units = "mm"
+    ),
     thirst_intensity_plot = create_intensity_plot(
       data = data_ptds,
       intensity_corr = intensity_corr
@@ -152,8 +157,8 @@ get_analysis_plan <- function() {
     # ),
     manuscript_pdf = target(
       command = {
-        rmarkdown::render(knitr_in("manuscript/index.Rmd"))
-        file_out("manuscript/index.pdf")
+        rmarkdown::render(knitr_in("manuscript/manuscript.Rmd"))
+        file_out("manuscript/manuscript.pdf")
       }
     )
   )

### 2020-08-13 manuscript/cover-letter/cover-letter.Rmd
pdf layout with numbers and 1.5 spacing

---
toname: Applied Nursing Research
toaddress: Editorial Board
opening: To the editor, 
closing: Regards,
spacing: 1
output: ConwayLetterTemplate::letter
---

On behalf of my co-authors, I am pleased to submit our manuscript entitled “Mokken scale analysis of the perioperative thirst discomfort scale".

Prior research indicates that most patients do not spontaneously verbalize thirst discomfort even when experiencing intense distress. As such, providing perioperative nurses with a validated tool that accurately measures thirst discomfort is vital. Results from our study provide additional insight regarding the validity of perioperative thirst discomfort scale in patients who were fasted before diagnostic or interventional procedures. We have not published this data previously.

### 2020-08-13 manuscript/index.Rmd
pdf layout with numbers and 1.5 spacing

@@
       correspondence: false
-  - Pauline Glaves:
-      institute: [PMCC]
-      correspondence: false
   - Shahvand Masihi:
@@
   - IR: Interventional Radiology, JDMI, Toronto General Hospital
-bibliography: Extras/references.bib
+bibliography: references.bib
 csl: Extras/sp-basic.csl
+always_allow_html: true
+
 output:
@@
 ```{r global_options, include=FALSE}
-knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
-
+knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.retina = 5)
+# tint options
+library(tint)
+knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
+options(htmltools.dir.version = FALSE)
 library(drake)
@@
 
-Symptoms of thirst discomfort are common in patients who are required to undergo long periods of fasting prior to medical and surgical procedures.
-The aim of this study was to examine the validity and reliability of the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures.
+Thirst discomfort is common because patients are required to undergo long periods of fasting before medical and surgical procedures.
+The aim of this study was to examine the validity and reliability of the perioperative thirst discomfort scale (PTDS) for measuring thirst discomfort before procedures.
 
@@
 
-A prospective observational study was conducted.
-Patients who were fasted and scheduled for an elective cardiac or interventional radiology procedure were included.
-Mokken scaling analysis was conducted to investigate the unidimensionality and hierarchical nature of the scale.
-Correlations between perioperative thirst discomfort scale scores and fasting durations were calculated to evaluate construct validity and convergent validity was evaluated by comparing scores with global thirst discomfort and thirst intensity items.
+Fasting patients who were scheduled for an elective cardiac or interventional radiology procedure were included in a prospective observational study.
+Mokken scaling analysis was used to investigate the dimensionality and hierarchical nature of the PTDS.
+PTDS score were compated with fasting duration to evaluate construct validity and convergent validity was evaluated by comparing scpre with global thirst discomfort and thirst intensity ratings.
 
@@
 
-Responses from 193 participants were included.
-The mean duration of fasting from food and non-clear fluids was `r food` hours and clear fluids was `r fluids` hours.
 Five items from the perioperative thirst discomfort scale (PTDS-5) formed a Mokken scale with evidence of invariant item ordering.
-The median  PTDS-5 score was `r ptds5`  from a possible range of 0 to 10.
-The hierarchical pattern of responses to PTDS-5 items is interpretable in terms of respondents most easily endorsing an item related to discomfort associated with the desire to drink water through to first endorsing symptoms related to dryness of the mouth and lips before those related to the abnormal sensations of 'thick' saliva and a 'thick' tongue.
-Scale reliability was adequate with a coefficient score (*rho* =`r drake::readd(reliability_ptds5)`).
+Participants most easily endorsed the item related to the desire to drink water through to first endorsing symptoms related to dryness of the mouth and lips before those related to the abnormal sensations of 'thick' saliva and a 'thick' tongue.
+Scale reliability was adequate (*rho* =`r drake::readd(reliability_ptds5)`).
 There was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (*tau* = `r round(global_thirst_corr$tau, 2)`; 95% CI = `r round(global_thirst_corr$CI_low, 2)` to `r round(global_thirst_corr$CI_high, 2)`), as well as thirst intensity (*tau* = `r round(intensity_corr$tau, 2)`; 95% CI = `r round(intensity_corr$CI_low, 2)` to `r round(intensity_corr$CI_high, 2)`).
@@
 The items in the PTDS-5 form a strong Mokken scale, meaning it is a reliable and precise way to order patients according to their thirst discomfort.
-Nurses may use this simple scale to assess the severity of thirst discomfort in patients who are fasted prior to procedures and initiate interventions according to individual needs.
 
@@
 
-Author contributions: 
-AC: Study design, data analysis and wrote manuscript; KC: Study design, revised the manuscript for important intellectual content;  MB: Data collection, revised the manuscript for important intellectual content; DP: Data collection, revised the manuscript for important intellectual content; NK: Study design, revised the manuscript for important intellectual content; AM: Study design, revised the manuscript for important intellectual content; MP: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; PG: Study design, revised the manuscript for important intellectual content; SA: Study design, revised the manuscript for important intellectual content.
+Author contributions: AC: Study design, data analysis and wrote manuscript; KC: Study design, revised the manuscript for important intellectual content; MB: Data collection, revised the manuscript for important intellectual content; DP: Data collection, revised the manuscript for important intellectual content; NK: Study design, revised the manuscript for important intellectual content; AM: Study design, revised the manuscript for important intellectual content; MP: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; SM: Study design, revised the manuscript for important intellectual content; PG: Study design, revised the manuscript for important intellectual content; SA: Study design, revised the manuscript for important intellectual content.
 
@@
 The item difficult hierarchy for the PTDS-5 is presented in Table \@ref(tab:tab2).
-The most highly endorsed symptom (i.e.
-the least severe symptom of thirst discomfort in the PTDS-5) was being bothered by a desire to drink water.
-The least endorsed symptom (i.e.
-the most severe symptom of thirst discomfort in the PTDS-5) was being bothered by the perception of a 'thick tongue'.
+The most highly endorsed symptom (i.e. the least severe symptom of thirst discomfort in the PTDS-5) was being bothered by a desire to drink water.
+The least endorsed symptom (i.e. the most severe symptom of thirst discomfort in the PTDS-5) was being bothered by the perception of a 'thick tongue'.
 The items in the PTDS-5 form a strong Mokken scale (H~s~ = `r drake::readd(H_values_ptds5)$H[1]`).
@@
 
-<div id="refs"></div>
+::: {#refs}
+:::
 
@@
 drake::loadd(summary_table)
-gtsummary::as_flextable(summary_table)
+# gtsummary::as_flextable(summary_table)
+gtsummary::as_kable_extra(summary_table, booktabs = T)
 ```
@@
 ```{r tab.id = "tab2", tab.cap= "Item hierarchy of the PTDS-5"}
-flextable::flextable(drake::readd(item_hierarchy_df_ptds5)) %>% 
-  autofit()
+# flextable::flextable(drake::readd(item_hierarchy_df_ptds5)) %>% 
+#   autofit()
+drake::readd(item_hierarchy_df_ptds5) %>% 
+  kableExtra::kable(booktabs = T)
 ```
@@
 ```{r distribution, fig.cap = "Distribution of PTDS-5 scores"}
-knitr::include_graphics("figures/distribution.png")
+# knitr::include_graphics("figures/distribution.png")
+drake::readd(distribution_plot)
 ```
@@
 ```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 0 to 10 with higher scores indicating worse thirst discomfort."}
-knitr::include_graphics("figures/thirst-intensity.png")
+# knitr::include_graphics("figures/thirst-intensity.png")
+drake::readd(thirst_intensity_plot)
 ```
@@
 ```{r globalthirst, fig.cap="Association between PTDS-5 and global thirst rating"}
-knitr::include_graphics("figures/global-thirst.png")
+# knitr::include_graphics("figures/global-thirst.png")
+library(dplyr)
+drake::readd(global_thirst_plot)
 ```
@@
 ```{r fasting, fig.cap = "Association between PTDS-5 and duration of fasting"}
-knitr::include_graphics("figures/fasting.png")
+# knitr::include_graphics("figures/fasting.png")
+drake::readd(fasting_duration_ptds_plot)
 ```
@@
 drake::readd(H_values_ptds7_table) %>% 
-  flextable::flextable() %>% 
-  autofit()
+  # flextable::flextable() %>% 
+  # autofit()
+  kableExtra::kable(booktabs = T)
 ```
@@
 
-flextable::flextable(drake::readd(monotonicity_df_ptds6_table)) %>%  
-  autofit()
+# flextable::flextable(drake::readd(monotonicity_df_ptds6_table)) %>%  
+#   autofit()
+ drake::readd(monotonicity_df_ptds6_table) %>% 
+kableExtra::kable(booktabs = T)
 ```
@@
 ```{r tab.id='aisp', tab.cap.pre="Appendix Table ", tab.cap = "aisp for increasing H thresholds (*c*)"}
-flextable::flextable(drake::readd(aisp)) %>% 
-  autofit()
+# flextable::flextable(drake::readd(aisp)) %>% 
+  # autofit()
+drake::readd(aisp) %>% 
+  kableExtra::kable(booktabs = T)
 ```
@@
 drake::readd(iio_ptds6_table) %>% 
-  flextable() %>% 
-  autofit()
+  # flextable() %>% 
+  # autofit()
+  kableExtra::kable(booktabs = T)
 ```
@@
 drake::readd(iio_ptds5_table) %>%
-    flextable() %>%
-    autofit()
+    # flextable() %>%
+    # autofit()
+  kableExtra::kable(booktabs = T)
 ```

### 2020-08-13 manuscript/manuscript.Rmd
pdf layout with numbers and 1.5 spacing

@@
 title: Mokken scale analysis of the perioperative thirst discomfort scale
-
-# Use letters for affiliations
 author:
   - name: ""
-#     affiliation: a
-#   - name: Second Author
-#     affiliation: a,b
-# address:
-#   - code: a
-#     address: Institute of Smoke and Magic, University of Sometown, Sometown, XY, 12345
-#   - code: b
-#     address: Department of Neat Tricks, Whereever State University, Someplace, MC, 67890
-
 date_subtitle: "Blinded for peer review"
-# Could be used, for example, to mention the bibliographic info in a post-print.
-# If not specified, defaults to "This version was compiled on \today"
-
-# For footer text  TODO(fold into template, allow free form two-authors)
-lead_author_surname: Conway et al.
-
-# Place eg a DOI URL or CRAN Package URL here
-doi_footer: "Data and statistical code is available [here](https://github.com/awconway/ptds)"
-
-# Abstract
+lead_author_surname: Blinded for peer review
+doi_footer: "Pre-print submitted to *Applied Nursing Research*"
 abstract: |
-   *Introduction:* Symptoms of thirst discomfort are common in patients who are required to undergo long periods of fasting prior to medical and surgical procedures.
-The aim of this study was to examine the validity and reliability of the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures.
-   *Methods:* A prospective observational study was conducted.
-Patients who were fasted and scheduled for an elective cardiac or interventional radiology procedure were included.
-Mokken scaling analysis was conducted to investigate the unidimensionality and hierarchical nature of the scale.
-Correlations between perioperative thirst discomfort scale scores and fasting durations were calculated to evaluate construct validity and convergent validity was evaluated by comparing scores with global thirst discomfort and thirst intensity ratings.
-   *Results:* 
+  *Introduction:* Thirst discomfort is common because patients are required to undergo long periods of fasting before medical and surgical procedures.
+The aim of this study was to examine the validity and reliability of the perioperative thirst discomfort scale (PTDS) for measuring thirst discomfort before procedures.
+  *Methods:* Fasting patients who were scheduled for an elective cardiac or interventional radiology procedure were included in a prospective observational study.
+Mokken scaling analysis was used to investigate the dimensionality and hierarchical nature of the PTDS.
+PTDS scores were compared with fasting duration to evaluate construct validity.
+Convergent validity was evaluated by comparing the PTDS with global thirst discomfort and thirst intensity ratings.
+  *Results:* Five items from the perioperative thirst discomfort scale (PTDS-5) formed a Mokken scale with evidence of invariant item ordering.
+Participants most easily endorsed the item related to the desire to drink water through to first endorsing symptoms related to dryness of the mouth and lips before those related to the abnormal sensations of ‘thick’ saliva and a ‘thick’ tongue.
+Scale reliability was adequate (rho = 0.84).
+There was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (tau = 0.54; 95% CI = 0.43 to 0.63), as well as thirst intensity (tau = 0.49; 95% CI = 0.38 to 0.59).
+Duration of fasting was not associated with PTDS-5 scores.
   *Conclusion:* The items in the PTDS-5 form a strong Mokken scale, meaning it is a reliable and precise way to order patients according to their thirst discomfort.
-Nurses may use this simple scale to assess the severity of thirst discomfort in patients who are fasted prior to procedures and initiate interventions according to individual needs.
-
-# Optional: One or more keywords
 keywords:
@@
   - mokken scale analysis
-
-# Optional: Bibliography 
-bibliography: references
-
-# Customize footer, eg by referencing the vignette
+bibliography: references.bib
 footer_contents: "Perioperative thirst discomfort scale"
-
-# Produce a pinp document
+#one_column: true
+fontsize: 12pt
 output: pinp::pinp
@@
   word: false
+  pdf: true
+header-includes: >
+  \usepackage[switch]{lineno}
+  \linenumbers
+  \usepackage{setspace}
+  \onehalfspacing
+
 ---
@@
 
-A prospective observational study was conducted.
-Patients who were fasted and scheduled for an elective cardiac or interventional radiology procedure were included.
-Mokken scaling analysis was conducted to investigate the unidimensionality and hierarchical nature of the scale.
-Correlations between perioperative thirst discomfort scale scores and fasting durations were calculated to evaluate construct validity and convergent validity was evaluated by comparing scores with global thirst discomfort and thirst intensity ratings.
+Fasting patients who were scheduled for an elective cardiac or interventional radiology procedure were included in a prospective observational study.
+Mokken scaling analysis was used to investigate the dimensionality and hierarchical nature of the PTDS.
+PTDS scores were compared with fasting duration to evaluate construct validity.
+Convergent validity was evaluated by comparing the PTDS with global thirst discomfort and thirst intensity ratings.
 
@@
 
-Responses from 193 participants were included.
-The mean duration of fasting from food and non-clear fluids was `r food` hours and clear fluids was `r fluids` hours.
 Five items from the perioperative thirst discomfort scale (PTDS-5) formed a Mokken scale with evidence of invariant item ordering.
-The median PTDS-5 score was `r ptds5` from a possible range of 0 to 10.
-The hierarchical pattern of responses to PTDS-5 items is interpretable in terms of respondents most easily endorsing an item related to discomfort associated with the desire to drink water through to first endorsing symptoms related to dryness of the mouth and lips before those related to the abnormal sensations of 'thick' saliva and a 'thick' tongue.
-Scale reliability was adequate with a coefficient score (*rho* =`r drake::readd(reliability_ptds5)`).
+Participants most easily endorsed the item related to the desire to drink water through to first endorsing symptoms related to dryness of the mouth and lips before those related to the abnormal sensations of 'thick' saliva and a 'thick' tongue.
+Scale reliability was adequate (*rho* = `r drake::readd(reliability_ptds5)`).
 There was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (*tau* = `r round(global_thirst_corr$tau, 2)`; 95% CI = `r round(global_thirst_corr$CI_low, 2)` to `r round(global_thirst_corr$CI_high, 2)`), as well as thirst intensity (*tau* = `r round(intensity_corr$tau, 2)`; 95% CI = `r round(intensity_corr$CI_low, 2)` to `r round(intensity_corr$CI_high, 2)`).
@@
 
-Pre-procedure fasting is used to reduce the risk of vomiting and aspiration pneumonia during sedation and general anaesthesia, or in case emergency intubation is required due to unexpected cardiac arrest.[@hamid2014pre; @osborne2002preoperative] However, prolonged fluid restriction causes thirst symptoms to develop (e.g., dry mouth, swollen tongue), which can lead to great discomfort.[@madsen1998perioperative] Current guidelines related to pre-procedure fasting for elective procedures recommend a minimum fasting period of 2 hours nil-per-os (NPO) for clear fluids.[@dobsonGuidelinesPracticeAnesthesia2018] Despite these recommendations, current practice is for patients undergoing surgical and other medical procedures that require sedation or anesthesia to receive standardized 'nil-by-mouth' fasting instructions at a pre-specified time interval before procedures.
+Pre-procedure fasting is used to reduce the risk of vomiting and aspiration pneumonia during sedation and general anaesthesia, or in case emergency intubation is required due to unexpected cardiac arrest [@hamid2014pre; @osborne2002preoperative].
+However, prolonged fluid restriction causes thirst symptoms to develop (e.g., dry mouth, swollen tongue), which can lead to great discomfort [@madsen1998perioperative].
+Current guidelines related to pre-procedure fasting for elective procedures recommend a minimum fasting period of 2 hours nil-per-os (NPO) for clear fluids [@dobsonGuidelinesPracticeAnesthesia2018].
+Despite these recommendations, current practice is for patients undergoing surgical and other medical procedures that require sedation or anesthesia to receive standardized 'nil-by-mouth' fasting instructions at a pre-specified time interval before procedures.
 For example, 'no eating or drinking after midnight' is most common.
 This 'standardized' instruction will not be changed regardless of whether or not there are alterations in scheduling throughout the day that result in significant delays in procedure start time.
-As a result, fasting durations far exceed the recommended requirement for most patients undergoing medical and surgical procedures.[@de2014actual; @sorita2015frequency; @spitz2017impact] For example, in a recent study of 3641 fasting orders at a large academic institution in the USA, it was found that the median fasting duration was 12.8 hours, averaging 2 missed meals.[@sorita2015frequency]
+As a result, fasting durations far exceed the recommended requirement for most patients undergoing medical and surgical procedures [@de2014actual; @sorita2015frequency; @spitz2017impact].
+For example, in a recent study of 3641 fasting orders at a large academic institution in the USA, it was found that the median fasting duration was 12.8 hours, averaging 2 missed meals [@sorita2015frequency].
 
 As a direct result of prolonged pre-procedure fasting, symptoms of thirst discomfort have been reported as common and severe.
-In a qualitative study where 12 participants were interviewed from a tertiary hospital in Australia, surgical patients and patients who adhered to prolonged fasting instructions described the discomfort from thirst symptoms to be the worst physical effect of fasting.[@carey2015qualitative] Similarly, Madsen et al. interviewed a convenience sample of 50 adult surgical patients who reported that thirst symptoms caused more discomfort than hunger, sleep, or anxiety related to the procedure.[@madsen1998perioperative] However, thirst and its symptoms continue to be undervalued, under-reported and infrequently assessed by health care providers, including the nursing team.[@milani2016thirst]
+In a qualitative study where 12 participants were interviewed from a tertiary hospital in Australia, surgical patients and patients who adhered to prolonged fasting instructions described the discomfort from thirst symptoms to be the worst physical effect of fasting [@carey2015qualitative].
+Similarly, [@madsen1998perioperative] interviewed a convenience sample of 50 adult surgical patients who reported that thirst symptoms caused more discomfort than hunger, sleep, or anxiety related to the procedure.
+However, thirst and its symptoms continue to be undervalued, under-reported and infrequently assessed by health care providers, including the nursing team [@milani2016thirst].
 
 Despite the relevance and value of assessing thirst-discomfort of patients, the subjective experience of thirst presents challenges in developing a valid and reliable tool to succinctly and accurately measure its symptoms and level of discomfort prior to procedures.
-A thirst-discomfort scale for perioperative use was developed at a surgical centre affiliated with an accredited public hospital in Brazil.[@Martins_2017] The seven-item perioperative thirst-discomfort scale (PTDS) was developed in three stages, including face and content validation and reliability, and based on the Consensus-based Standards for the selection of health status Measurement Instruments (COSMIN) checklist.[@mokkink2010cosmin] Inter-rater reliability was tested through inter-observer equivalence where a pair of nurses independently, but simultaneously administered the scale among 70 patients.
-Six items on the scale had a weighted kappa coefficient of 1, while the item "I feel like drinking water" was 0.97.[@Martins_2017] Cronbach's alpha was used to test the internal consistency of the scale with a value of 0.91.[@Martins_2017]
+A thirst-discomfort scale for perioperative use was developed at a surgical centre affiliated with an accredited public hospital in Brazil [@Martins_2017].
+The seven-item perioperative thirst-discomfort scale (PTDS) was developed in three stages, including face and content validation and reliability, and based on the Consensus-based Standards for the selection of health status Measurement Instruments (COSMIN) checklist [@mokkink2010cosmin].
+Inter-rater reliability was tested through inter-observer equivalence where a pair of nurses independently, but simultaneously administered the scale among 70 patients.
+Six items on the scale had a weighted kappa coefficient of 1, while the item "I feel like drinking water" was 0.97 [@Martins_2017].
+Cronbach's alpha was used to test the internal consistency of the scale with a value of 0.91 [@Martins_2017].
 
@@
 
-### Inclusion criteria
-
 Adult patients who were fasted and scheduled for an elective procedure in the Cardiac Catheterization Laboratories or in Interventional Radiology at an academic teaching hospital in Canada were included.
@@
 1.  Under 16 years of age
-2.  Undergoing an emergency procedure
-3.  Unable to understand or speak English and a translator is not immediately available
-4.  Nurse in pre-procedural bay considered that there was insufficient time prior to anticipated commencement of the procedure for participation in the study.
+2.  Scheduled for an emergency procedure
+3.  Unable to understand or speak English and a translator was not immediately available
+4.  Nurse in pre-procedure bay considered that there was insufficient time prior to anticipated commencement of the procedure for participation in the study.
 
@@
 
-Mokken scaling is a method for establishing if items in a scale conform to a cumulative, hierarchical structure.[@perng2012construct] When an instrument contains hierarchically ordered items, the items can be ordered from those indicating mild to severe symptoms.
+Mokken scaling is a method for establishing if items in a scale conform to a cumulative, hierarchical structure [@perng2012construct].
+When an instrument contains hierarchically ordered items, the items can be ordered from those indicating mild to severe symptoms.
 An intentionally simple example of a hierarchical scale is a set of items enquiring about height that increases in increments of 10cm from 100cm to 200cm.
-In this example, taller people would endorse more items than those who were shorter.[@watson2008hierarchy]
+In this example, taller people would endorse more items than those who were shorter [@watson2008hierarchy].
 
@@
 If **H~i~** was below 0.3 or if the lower limit of the 95% CI (confidence interval) for *H~i~* was below 0.3, the item was to be excluded.
-Then, scale partitioning was carried out to explore the dimensions of the PTDS through increasing c (lower bound c defines the minimum value of coefficients **H~i~** in the Mokken scale by 0.05 increments).[@molenaar2000mps5] Monotone homogeneity and invariant item ordering were investigated at the whole scale level, as no sub-scales were identified.
+Then, scale partitioning was carried out to explore the dimensions of the PTDS through increasing c (lower bound c defines the minimum value of coefficients **H~i~** in the Mokken scale by 0.05 increments) [@molenaar2000mps5].
+Monotone homogeneity and invariant item ordering were investigated at the whole scale level, as no sub-scales were identified.
 To assess for violations of invariant item ordering, crit values \<40 were considered acceptable.[@molenaar2000mps5] Provided invariant item ordering holds for a set of items, the coefficient **H~T~** expresses the accuracy of the ordering, with values below 0.3 being unacceptable and above 0.4 indicating moderate invariant item ordering.[@Ligtvoet_2010] The coefficient score (rho), which is similar to Cronbach's alpha, was used to assess reliability of the scale.
@@
 
+The R statistical program was used to conduct all analyses [@rcore2020].
+All data and R code, along with instructions for how to completely reproduce results of the analyses, is available [here](https://github.com/awconway/ptds).
+
 # Results
@@
 One participant did not complete 1 of the PTDS-7 items and was excluded from the analysis.
-A summary of participant characteristics is displayed in Table \@ref(tab:tab1).
+A summary of participant characteristics is displayed in Table <!--\@ref(tab:tab1)-->\ref{tab:tab1pdf}.
 The median age was 62 (IQR 48 - 72) and 58% (n=111) were female.
@@
 
-```{r tab.cap="Participant characteristics", tab.id = 'tab1'}
+```{r tab1pdf, eval=params$pdf}
 drake::loadd(summary_table)
-# gtsummary::as_flextable(summary_table)
-gtsummary::as_kable_extra(summary_table, booktabs = T)
+gtsummary::as_kable_extra(summary_table, booktabs = T,
+                          caption = "Participant characteristics")
 ```
@@
 
-No items required removal from the PTDS-7 based on the condition that scalability coefficients (*H~i~* and lower bound of 95% CI) should be over 0.3 (Appendix Table \@ref(tab:Hvaluesptds7)).
+No items required removal from the PTDS-7 based on the condition that scalability coefficients (*H~i~* and lower bound of 95% CI) should be over 0.3.
 We used a lower bound for *c*, starting from 0.05 and increasing to 0.80 in 0.05 increments, to explore the dimensionality of the PTDS-7.
@@
 The 6 remaining items (PTDS-6) were further examined for the criteria of a Mokken scale and item invariant ordering.
-The PTDS-6 item set has a homogeneity value *H(se)* of `r readd(H_values_ptds6)$H`, which indicated a strong Mokken scale.
-Results indicated that there were no significant violations of the assumption of monotonicity (z-sig).
-As such, no items were removed based on this criteria.
+The PTDS-6 item set had a homogeneity value *H(se)* of `r drake::readd(H_values_ptds6)$H[1]`, which indicated a strong Mokken scale.
+No items were removed based on the criteria that there should be no significant violations of the assumption of monotonicity.
+
+```{r hierarchytabpdf, eval=params$pdf}
+drake::readd(item_hierarchy_df_ptds5) %>% 
+  kableExtra::kable(booktabs=T, caption="Item hierarchy of the PTDS-5")
+```
 
@@
 
-Analysis of invariant item ordering for the PTDS-6 revealed violations for the dry throat/dry lips item pair (Appendix Table \@ref(tab:iioptds6)).
+Analysis of invariant item ordering for the PTDS-6 revealed violations for the dry throat/dry lips item pair.
 Although there were no *significant* intersections between items, due to the intersection between the dry throat and dry lips items, crit values were above 40.
 For this reason, we excluded one of these items (dry throat), and re-assessed the remaining 5 items for the assumptions of a Mokken scale.
-With the dry throat item removed, none of the items violated the assumption of invariant item ordering (Appendix Table \@ref(tab:iioptds5)).
+With the dry throat item removed, none of the items violated the assumption of invariant item ordering.
 
@@
 
-The item difficult hierarchy for the PTDS-5 is presented in Table \@ref(tab:tab2).
+The item difficult hierarchy for the PTDS-5 is presented in Table <!--\@ref(tab:hierarchytabpdf)-->\ref{tab:hierarchytabpdf}.
 The most highly endorsed symptom (i.e. the least severe symptom of thirst discomfort in the PTDS-5) was being bothered by a desire to drink water.
@@
 
-```{r tab.id = "tab2", tab.cap= "Item hierarchy of the PTDS-5"}
-drake::readd(item_hierarchy_df_ptds5) %>% 
-  kableExtra::kable(booktabs=T)
-```
-
 ## Construct and convergent validity
 
-The median PTDS-5 score was `r ptds5`. The distribution of PTDS-5 scores was right-skewed (Figure \@ref(fig:distribution)).
+The distribution of PTDS-5 scores in the sample was right-skewed, with a median score of `r ptds5`.
+(Figure <!--\@ref(fig:distribution)-->\ref{fig:distributionpdf}).
+As shown in Figure <!--\@ref(fig:globalthirst)-->\ref{fig:globalthirst}, there was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (*tau* = `r round(global_thirst_corr$tau, 2)`; 95% CI = `r round(global_thirst_corr$CI_low, 2)` to `r round(global_thirst_corr$CI_high, 2)`).
+The correlation between PTDS-5 scores and thirst intensity was also high (*tau* = `r round(intensity_corr$tau, 2)`; 95% CI = `r round(intensity_corr$CI_low, 2)` to `r round(intensity_corr$CI_high, 2)`).
+Likewise, the correlation between pain and PTDS-5 scores was not significant (Spearman's rho = `r round(drake::readd(pain_corr)$rho, 2)`; 95% CI = `r round(drake::readd(pain_corr)$CI_low, 2)` to `r round(drake::readd(pain_corr)$CI_high, 2)`).
 
-```{r distribution, fig.cap = "Distribution of PTDS-5 scores (higher scores indicate worse thirst discomfort)"}
+```{r distributionpdf, fig.cap = "Distribution of PTDS-5 scores (higher scores indicate worse thirst discomfort)", eval=params$pdf}
 # knitr::include_graphics("figures/distribution.png")
@@
 
-There was a positive correlation between PTDS-5 scores and thirst intensity (Figure \@ref(fig:intensity)), as well as the global thirst discomfort rating (Figure \@ref(fig:globalthirst)).
-The associations between PTDS-5 scores with food and fluid duration is displayed in Figure \@ref(fig:fasting).
-
-```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 0 to 10 with higher scores indicating worse thirst discomfort."}
-# knitr::include_graphics("figures/thirst-intensity.png")
-drake::readd(thirst_intensity_plot)
-```
-
-```{r globalthirst, fig.cap="Association between PTDS-5 and global thirst rating", eval=params}
+```{r globalthirst, fig.cap="Association between PTDS-5 and global thirst rating", eval=params$pdf}
 # knitr::include_graphics("figures/global-thirst.png")
@@
 
-Longer duration of fasting was not linearly associated with PTDS-5 scores.
-Likewise, the correlation between pain and PTDS-5 scores was not significant (Spearman's rho = `r round(drake::readd(pain_corr)$rho, 2)`; 95% CI = `r round(drake::readd(pain_corr)$CI_low, 2)` to `r round(drake::readd(pain_corr)$CI_high, 2)`).
+The associations between PTDS-5 scores with food and fluid duration are displayed in Figure <!--\@ref(fig:fasting)-->\ref{fig:fasting}.
+Longer duration of fasting was not linearly associated with PTDS-5 scores for fluids (Spearman's rho = `r round(drake::readd(fluids_corr)$rho, 2)`; 95% CI = `r round(drake::readd(fluids_corr)$CI_low, 2)` to `r round(drake::readd(fluids_corr)$CI_high, 2)`) or food (Spearman's rho = `r round(drake::readd(food_corr)$rho, 2)`; 95% CI = `r round(drake::readd(food_corr)$CI_low, 2)` to `r round(drake::readd(food_corr)$CI_high, 2)`).
 
@@
 In addition, an acceptable accuracy for invariant item ordering was identified, meaning patients are likely to rank the PTDS-5 symptoms of thirst discomfort in the same way.
-As such, the item difficulty hierarchy, presented in Table \@ref(tab:tab2) can be used as a guide to easily determine which patients are experiencing worse thirst discomfort.
+As such, the item difficulty hierarchy, presented in Table <!--\@ref(tab:hierarchytabpdf)-->\ref{tab:hierarchytabpdf}, can be used as a guide to easily determine which patients are experiencing worse thirst discomfort.
 The hierarchical pattern of responses to items in the Mokken scale is interpretable in terms of respondents more easily endorsing items related to a general desire to drink water through to first endorsing more specific symptoms related to dryness of the mouth and lips prior to endorsing discomfort associated with the sensations of 'thick' saliva and 'thick' tongue.
@@
 
-The original (7-item) perioperative thirst-discomfort scale has previously been used to determine the relationship between thirst intensity and thirst-discomfort in the immediate postoperative period during anesthesia recovery.[@Pierotti_2018] In a non-probabilistic sample of 203 adult participants in post-operative recovery following elective, urgent and emergency surgeries, the investigators found no association between fasting duration, or patient age, and the intensity of thirst and degree of discomfort.[@Pierotti_2018] However, the mean fasting duration among participants was uniformly high (mean 16.2 [SD 8.7] hours), which may have prevented the researchers from discerning a correlation between fasting duration.
-Even in the subgroup of participant who reported 8 hours of fasting, all reported thirst.[@aroni2012assessment; @Pierotti_2018]
+The original (7-item) perioperative thirst-discomfort scale has previously been used to determine the relationship between thirst intensity and thirst-discomfort in the immediate postoperative period during anesthesia recovery.[@Pierotti_2018] In a non-probabilistic sample of 203 adult participants in post-operative recovery following elective, urgent and emergency surgeries, the investigators found no association between fasting duration, or patient age, and the intensity of thirst and degree of discomfort [@Pierotti_2018].
+However, the mean fasting duration among participants was uniformly high (mean 16.2 [SD 8.7] hours), which may have prevented the researchers from discerning a correlation between fasting duration.
+Even in the subgroup of participant who reported 8 hours of fasting, all reported thirst [@aroni2012assessment; @Pierotti_2018].
 
@@
 If thirst discomfort is present, and fasting is still indicated, interventions to alleviate symptoms should be implemented.
-There is a lack of evidence regarding the efficacy of any such interventions.
+There is a lack of evidence regarding the efficacy of interventions to alleviate perioperative thirst discomfort.
+However, recent randomized controlled trials have identified clinically relevant redictions in thirst discomfort with menthol chewing gum and 30mL menthol popsicles [@garcia2019; @nascimento2020].
 
@@
 <!-- ::: {#refs} -->
+
 <!-- ::: -->
@@
 
+```{r tab1, tab.cap="Participant characteristics", tab.id = 'tab1', eval=params$word}
+drake::loadd(summary_table)
+gtsummary::as_flextable(summary_table)
+```
+
+```{r hierarchytab, tab.id = "tab2", tab.cap= "Item hierarchy of the PTDS-5", eval=params$word}
+drake::readd(item_hierarchy_df_ptds5) %>% 
+  flextable::flextable() %>% 
+  flextable::autofit()
+```
+
+```{r distribution, fig.cap = "Distribution of PTDS-5 scores (higher scores indicate worse thirst discomfort)", eval=params$word}
+# knitr::include_graphics("figures/distribution.png")
+drake::readd(distribution_plot)
+```
+
+```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 0 to 10 with higher scores indicating worse thirst discomfort.", eval=params$word}
+# knitr::include_graphics("figures/thirst-intensity.png")
+drake::readd(thirst_intensity_plot)
+```
+
 <!---BLOCK_MULTICOL_START--->

### 2020-08-13 manuscript/manuscript.tex
pdf layout with numbers and 1.5 spacing

@@
-\documentclass[letterpaper,9pt,twocolumn,twoside,]{pinp}
+\documentclass[12pt,twocolumn,twoside,]{pinp}
 
@@
 
+\usepackage[switch]{lineno} \linenumbers \usepackage{setspace} \onehalfspacing
+
 \usepackage{booktabs}
@@
 
-\title{A Template for Two or One-Column Vignettes}
+\title{Mokken scale analysis of the perioperative thirst discomfort
+scale}
 
@@
 % Please give the surname of the lead author for the running footer
-\leadauthor{Author and Author}
+\leadauthor{Blinded for peer review}
 
 % Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
- \keywords{  one |  two |  optional |  keywords |  here  }
+ \keywords{  fasting |  anesthesia |  thirst
+discomfort |  patient-reported outcomes |  scale validation |  mokken
+scale analysis  }
 
 \begin{abstract}
-Your abstract will be typeset here, and used by default a visually
-distinctive font. An abstract should explain to the general reader the
-major contributions of the article.
+\emph{Introduction:} Thirst discomfort is common because patients are
+required to undergo long periods of fasting before medical and surgical
+procedures. The aim of this study was to examine the validity and
+reliability of the perioperative thirst discomfort scale (PTDS) for
+measuring thirst discomfort before procedures. \emph{Methods:} Fasting
+patients who were scheduled for an elective cardiac or interventional
+radiology procedure were included in a prospective observational study.
+Mokken scaling analysis was used to investigate the dimensionality and
+hierarchical nature of the PTDS. PTDS scores were compared with fasting
+duration to evaluate construct validity. Convergent validity was
+evaluated by comparing the PTDS with global thirst discomfort and thirst
+intensity ratings. \emph{Results:} Five items from the perioperative
+thirst discomfort scale (PTDS-5) formed a Mokken scale with evidence of
+invariant item ordering. Participants most easily endorsed the item
+related to the desire to drink water through to first endorsing symptoms
+related to dryness of the mouth and lips before those related to the
+abnormal sensations of `thick' saliva and a `thick' tongue. Scale
+reliability was adequate (rho = 0.84). There was a positive correlation
+between PTDS-5 scores and the global thirst discomfort rating (tau =
+0.54; 95\% CI = 0.43 to 0.63), as well as thirst intensity (tau = 0.49;
+95\% CI = 0.38 to 0.59). Duration of fasting was not associated with
+PTDS-5 scores. \emph{Conclusion:} The items in the PTDS-5 form a strong
+Mokken scale, meaning it is a reliable and precise way to order patients
+according to their thirst discomfort.
 \end{abstract}
 
-\dates{This version was compiled on \today}
+\dates{Blinded for peer review}
 
@@
 % new name is doi_footer
-\doifooter{\url{https://cran.r-project.org/package=YourPackage}}
+\doifooter{Pre-print submitted to \emph{Applied Nursing Research}}
 
@@
 
-\acknow{This template package builds upon, and extends, the work of the
-excellent \href{https://cran.r-project.org/package=rticles}{rticles}
-package, and both packages rely on the
-\href{http://www.pnas.org/site/authors/latex.xhtml}{PNAS LaTeX} macros.
-Both these sources are gratefully acknowledged as this work would not
-have been possible without them. Our extensions are under the same
-respective licensing term
-(\href{https://www.gnu.org/licenses/gpl-3.0.en.html}{GPL-3} and
-\href{https://www.latex-project.org/lppl/}{LPPL (\textgreater= 1.3)}).}
-
-\hypertarget{abstract}{%
-\subsection{Abstract}\label{abstract}}
-
 \hypertarget{introduction}{%
-\subsubsection{Introduction}\label{introduction}}
-
-Symptoms of thirst discomfort are common in patients who are required to
-undergo long periods of fasting prior to medical and surgical
-procedures. The aim of this study was to examine the validity and
-reliability of the perioperative thirst discomfort scale for measuring
-thirst discomfort prior to procedures.
-
-\hypertarget{methods}{%
-\subsubsection{Methods}\label{methods}}
-
-A prospective observational study was conducted. Patients who were
-fasted and scheduled for an elective cardiac or interventional radiology
-procedure were included. Mokken scaling analysis was conducted to
-investigate the unidimensionality and hierarchical nature of the scale.
-Correlations between perioperative thirst discomfort scale scores and
-fasting durations were calculated to evaluate construct validity and
-convergent validity was evaluated by comparing scores with global thirst
-discomfort and thirst intensity ratings.
-
-\hypertarget{results}{%
-\subsubsection{Results}\label{results}}
-
-Responses from 193 participants were included. The mean duration of
-fasting from food and non-clear fluids was 12.7 (SD 3.8) hours and clear
-fluids was 9 (SD 4.5) hours. Five items from the perioperative thirst
-discomfort scale (PTDS-5) formed a Mokken scale with evidence of
-invariant item ordering. The median PTDS-5 score was 3 (IQR 1, 6) from a
-possible range of 0 to 10. The hierarchical pattern of responses to
-PTDS-5 items is interpretable in terms of respondents most easily
-endorsing an item related to discomfort associated with the desire to
-drink water through to first endorsing symptoms related to dryness of
-the mouth and lips before those related to the abnormal sensations of
-`thick' saliva and a `thick' tongue. Scale reliability was adequate with
-a coefficient score (\emph{rho} =0.84). There was a positive correlation
-between PTDS-5 scores and the global thirst discomfort rating
-(\emph{tau} = 0.54; 95\% CI = 0.43 to 0.63), as well as thirst intensity
-(\emph{tau} = 0.49; 95\% CI = 0.38 to 0.59). Duration of fasting was not
-associated with PTDS-5 scores.
-
-\hypertarget{conclusion}{%
-\subsubsection{Conclusion}\label{conclusion}}
-
-The items in the PTDS-5 form a strong Mokken scale, meaning it is a
-reliable and precise way to order patients according to their thirst
-discomfort. Nurses may use this simple scale to assess the severity of
-thirst discomfort in patients who are fasted prior to procedures and
-initiate interventions according to individual needs.
-
-Keywords: fasting, aspiration, patient safety, thirst, anesthesia,
-nursing, Mokken scale, scale development, item response theory
-
-\hypertarget{declarations}{%
-\subsection{Declarations}\label{declarations}}
-
-Availability of data and material (data transparency): All data used is
-available \href{https://github.com/awconway/ptds}{here}.
-
-Code availability: Code to reproduce the analyses is available
-\href{https://github.com/awconway/ptds}{here}.
-
-Abbreviated title: Perioperative thirst discomfort scale
-
-Funding statement: Support was provided solely from institutional and/or
-departmental sources. MP is supported by an Early Investigator Merit
-Award from the Department of Anesthesia, University of Toronto and
-Toronto General Hospital.
-
-Conflicts of interest: The authors declare no competing interests.
-
-Author contributions: AC: Study design, data analysis and wrote
-manuscript; KC: Study design, revised the manuscript for important
-intellectual content; MB: Data collection, revised the manuscript for
-important intellectual content; DP: Data collection, revised the
-manuscript for important intellectual content; NK: Study design, revised
-the manuscript for important intellectual content; AM: Study design,
-revised the manuscript for important intellectual content; MP: Study
-design, revised the manuscript for important intellectual content; SM:
-Study design, revised the manuscript for important intellectual content;
-SM: Study design, revised the manuscript for important intellectual
-content; PG: Study design, revised the manuscript for important
-intellectual content; SA: Study design, revised the manuscript for
-important intellectual content.
-
-\hypertarget{introduction-1}{%
-\subsection{Introduction}\label{introduction-1}}
+\section{Introduction}\label{introduction}}
 
@@
 aspiration pneumonia during sedation and general anaesthesia, or in case
-emergency intubation is required due to unexpected cardiac
-arrest.\citep{hamid2014pre, osborne2002preoperative} However, prolonged
-fluid restriction causes thirst symptoms to develop (e.g., dry mouth,
-swollen tongue), which can lead to great
-discomfort.\citep{madsen1998perioperative} Current guidelines related to
+emergency intubation is required due to unexpected cardiac arrest
+\citep{hamid2014pre, osborne2002preoperative}. However, prolonged fluid
+restriction causes thirst symptoms to develop (e.g., dry mouth, swollen
+tongue), which can lead to great discomfort
+\citep{madsen1998perioperative}. Current guidelines related to
 pre-procedure fasting for elective procedures recommend a minimum
-fasting period of 2 hours nil-per-os (NPO) for clear
-fluids.\citep{dobsonGuidelinesPracticeAnesthesia2018} Despite these
+fasting period of 2 hours nil-per-os (NPO) for clear fluids
+\citep{dobsonGuidelinesPracticeAnesthesia2018}. Despite these
 recommendations, current practice is for patients undergoing surgical
@@
 exceed the recommended requirement for most patients undergoing medical
-and surgical
-procedures.\citep{de2014actual, sorita2015frequency, spitz2017impact}
-For example, in a recent study of 3641 fasting orders at a large
-academic institution in the USA, it was found that the median fasting
-duration was 12.8 hours, averaging 2 missed
-meals.\citep{sorita2015frequency}
+and surgical procedures
+\citep{de2014actual, sorita2015frequency, spitz2017impact}. For example,
+in a recent study of 3641 fasting orders at a large academic institution
+in the USA, it was found that the median fasting duration was 12.8
+hours, averaging 2 missed meals \citep{sorita2015frequency}.
 
@@
 prolonged fasting instructions described the discomfort from thirst
-symptoms to be the worst physical effect of
-fasting.\citep{carey2015qualitative} Similarly, Madsen et
-al.~interviewed a convenience sample of 50 adult surgical patients who
+symptoms to be the worst physical effect of fasting
+\citep{carey2015qualitative}. Similarly, \citep{madsen1998perioperative}
+interviewed a convenience sample of 50 adult surgical patients who
 reported that thirst symptoms caused more discomfort than hunger, sleep,
-or anxiety related to the procedure.\citep{madsen1998perioperative}
-However, thirst and its symptoms continue to be undervalued,
-under-reported and infrequently assessed by health care providers,
-including the nursing team.\citep{milani2016thirst}
+or anxiety related to the procedure. However, thirst and its symptoms
+continue to be undervalued, under-reported and infrequently assessed by
+health care providers, including the nursing team
+\citep{milani2016thirst}.
 
@@
 thirst-discomfort scale for perioperative use was developed at a
-surgical centre affiliated with an accredited public hospital in
-Brazil.\citep{Martins_2017} The seven-item perioperative
-thirst-discomfort scale (PTDS) was developed in three stages, including
-face and content validation and reliability, and based on the
-Consensus-based Standards for the selection of health status Measurement
-Instruments (COSMIN) checklist.\citep{mokkink2010cosmin} Inter-rater
-reliability was tested through inter-observer equivalence where a pair
-of nurses independently, but simultaneously administered the scale among
-70 patients. Six items on the scale had a weighted kappa coefficient of
-1, while the item ``I feel like drinking water'' was
-0.97.\citep{Martins_2017} Cronbach's alpha was used to test the internal
-consistency of the scale with a value of 0.91.\citep{Martins_2017}
+surgical centre affiliated with an accredited public hospital in Brazil
+\citep{Martins_2017}. The seven-item perioperative thirst-discomfort
+scale (PTDS) was developed in three stages, including face and content
+validation and reliability, and based on the Consensus-based Standards
+for the selection of health status Measurement Instruments (COSMIN)
+checklist \citep{mokkink2010cosmin}. Inter-rater reliability was tested
+through inter-observer equivalence where a pair of nurses independently,
+but simultaneously administered the scale among 70 patients. Six items
+on the scale had a weighted kappa coefficient of 1, while the item ``I
+feel like drinking water'' was 0.97 \citep{Martins_2017}. Cronbach's
+alpha was used to test the internal consistency of the scale with a
+value of 0.91 \citep{Martins_2017}.
 
@@
 
-\hypertarget{methods-1}{%
-\subsection{Methods}\label{methods-1}}
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
 \hypertarget{study-design}{%
-\subsubsection{Study Design}\label{study-design}}
+\subsection{Study Design}\label{study-design}}
 
@@
 \hypertarget{participants}{%
-\subsubsection{Participants}\label{participants}}
-
-\hypertarget{inclusion-criteria}{%
-\paragraph{Inclusion criteria}\label{inclusion-criteria}}
+\subsection{Participants}\label{participants}}
 
@@
 \hypertarget{exclusion-criteria}{%
-\paragraph{Exclusion criteria}\label{exclusion-criteria}}
+\subsubsection{Exclusion criteria}\label{exclusion-criteria}}
 
@@
 \item
-  Undergoing an emergency procedure
+  Scheduled for an emergency procedure
 \item
-  Unable to understand or speak English and a translator is not
+  Unable to understand or speak English and a translator was not
   immediately available
 \item
-  Nurse in pre-procedural bay considered that there was insufficient
-  time prior to anticipated commencement of the procedure for
-  participation in the study.
+  Nurse in pre-procedure bay considered that there was insufficient time
+  prior to anticipated commencement of the procedure for participation
+  in the study.
 \end{enumerate}
@@
 \hypertarget{data-collection}{%
-\subsubsection{Data collection}\label{data-collection}}
+\subsection{Data collection}\label{data-collection}}
 
@@
 \hypertarget{measures}{%
-\paragraph{Measures}\label{measures}}
+\subsubsection{Measures}\label{measures}}
 
@@
 \hypertarget{statistical-analysis}{%
-\subsubsection{Statistical analysis}\label{statistical-analysis}}
+\subsection{Statistical analysis}\label{statistical-analysis}}
 
 Mokken scaling is a method for establishing if items in a scale conform
-to a cumulative, hierarchical structure.\citep{perng2012construct} When
+to a cumulative, hierarchical structure \citep{perng2012construct}. When
 an instrument contains hierarchically ordered items, the items can be
@@
 example, taller people would endorse more items than those who were
-shorter.\citep{watson2008hierarchy}
+shorter \citep{watson2008hierarchy}.
 
@@
 of coefficients \textbf{H\textsubscript{i}} in the Mokken scale by 0.05
-increments).\citep{molenaar2000mps5} Monotone homogeneity and invariant
+increments) \citep{molenaar2000mps5}. Monotone homogeneity and invariant
 item ordering were investigated at the whole scale level, as no
@@
 
-\hypertarget{results-1}{%
-\subsection{Results}\label{results-1}}
+The R statistical program was used to conduct all analyses
+\citep{rcore2020}. All data and R code, along with instructions for how
+to completely reproduce results of the analyses, is available
+\href{https://github.com/awconway/ptds}{here}.
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
@@
 the analysis. A summary of participant characteristics is displayed in
-Table @ref(tab:tab1). The median age was 62 (IQR 48 - 72) and 58\%
+Table \ref{tab:tab1pdf}. The median age was 62 (IQR 48 - 72) and 58\%
 (n=111) were female. 43\% of patients were scheduled to have their
@@
 
-\hypertarget{mokken-scale-analysis}{%
-\subsubsection{Mokken scale analysis}\label{mokken-scale-analysis}}
+\begin{table}
 
-No items required removal from the PTDS-7 based on the condition that
-scalability coefficients (\emph{H\textsubscript{i}} and lower bound of
-95\% CI) should be over 0.3 (Appendix Table @ref(tab:Hvaluesptds7)). We
-used a lower bound for \emph{c}, starting from 0.05 and increasing to
-0.80 in 0.05 increments, to explore the dimensionality of the PTDS-7.
-From 0.05 to 0.45, all of the items formed a single scale. The item,
-`\emph{I have a bad taste in my mouth}', was excluded by selecting the
-remaining items which showed uni-dimensionality at a threshold level of
-.50. The 6 remaining items (PTDS-6) were further examined for the
-criteria of a Mokken scale and item invariant ordering. The PTDS-6 item
-set has a homogeneity value \emph{H(se)} of 0.599, (0.034), which
-indicated a strong Mokken scale. Results indicated that there were no
-significant violations of the assumption of monotonicity (z-sig). As
-such, no items were removed based on this criteria.
-
-\hypertarget{invariant-item-ordering}{%
-\paragraph{Invariant item ordering}\label{invariant-item-ordering}}
-
-Analysis of invariant item ordering for the PTDS-6 revealed violations
-for the dry throat/dry lips item pair (Appendix Table
-@ref(tab:iioptds6)). Although there were no \emph{significant}
-intersections between items, due to the intersection between the dry
-throat and dry lips items, crit values were above 40. For this reason,
-we excluded one of these items (dry throat), and re-assessed the
-remaining 5 items for the assumptions of a Mokken scale. With the dry
-throat item removed, none of the items violated the assumption of
-invariant item ordering (Appendix Table @ref(tab:iioptds5)).
-
-\hypertarget{accuracy-and-reliability}{%
-\paragraph{Accuracy and reliability}\label{accuracy-and-reliability}}
-
-The accuracy of the item ordering was found to be acceptable with a
-\emph{H\textsubscript{T}} coefficient of 0.397. Scale reliability was
-adequate with a coefficient score (Rho) of 0.84.
-
-\hypertarget{item-difficulty-hierarchy}{%
-\paragraph{Item difficulty hierarchy}\label{item-difficulty-hierarchy}}
-
-The item difficult hierarchy for the PTDS-5 is presented in Table
-@ref(tab:tab2). The most highly endorsed symptom (i.e.~the least severe
-symptom of thirst discomfort in the PTDS-5) was being bothered by a
-desire to drink water. The least endorsed symptom (i.e.~the most severe
-symptom of thirst discomfort in the PTDS-5) was being bothered by the
-perception of a `thick tongue'. The items in the PTDS-5 form a strong
-Mokken scale (H\textsubscript{s} = 0.605).
-
-\begin{ShadedResult}
-\begin{verbatim}
-#                     item
-#  1 I want to drink water
-#  2       My mouth is dry
-#  3       My lips are dry
-#  4    My saliva is thick
-#  5    My tongue is thick
-\end{verbatim}
-\end{ShadedResult}
-
-\hypertarget{construct-and-convergent-validity}{%
-\subsubsection{Construct and convergent
-validity}\label{construct-and-convergent-validity}}
-
-The median PTDS-5 score was 3 (IQR 1, 6). The distribution of PTDS-5
-scores was right-skewed (Figure @ref(fig:distribution)). There was a
-positive correlation between PTDS-5 scores and thirst intensity (Figure
-@ref(fig:intensity)), as well as the global thirst discomfort rating
-(Figure @ref(fig:globalthirst)). The associations between PTDS-5 scores
-with food and fluid duration is displayed in Figure @ref(fig:fasting).
-Longer duration of fasting was not linearly associated with PTDS-5
-scores. Likewise, the correlation between pain and PTDS-5 scores was not
-significant (Spearman's rho = 0.07; 95\% CI = -0.07 to 0.21).
-
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
-
-Results indicate that the PTDS-5 is a reliable and precise way to order
-patients according to their thirst discomfort in the
-\emph{pre-procedure} period. In addition, an acceptable accuracy for
-invariant item ordering was identified, meaning patients are likely to
-rank the PTDS-5 symptoms of thirst discomfort in the same way. As such,
-the item difficulty hierarchy, presented in Table @ref(tab:tab2) can be
-used as a guide to easily determine which patients are expriencing worse
-thirst discomfort. The hierarchical pattern of responses to items in the
-Mokken scale is interpretable in terms of respondents more easily
-endorsing items related to a general desire to drink water through to
-first endorsing more specific symptoms related to dryness of the mouth
-and lips prior to endorsing discomfort associated with the sensations of
-`thick' saliva and `thick' tongue. In more practical terms, a patient
-who reports feeling bothered by a sensation that saliva is thick, could
-be interpreted as experiencing worse thirst discomfort than a patient
-who did not endorse any of the items lower down on the item difficulty
-hierarchy (i.e.~`my lips are dry' and `my mouth is dry').
-
-The original (7-item) perioperative thirst-discomfort scale has
-previously been used to determine the relationship between thirst
-intensity and thirst-discomfort in the immediate postoperative period
-during anesthesia recovery.\citep{Pierotti_2018} In a non-probabilistic
-sample of 203 adult participants in post-operative recovery following
-elective, urgent and emergency surgeries, the investigators found no
-association between fasting duration, or patient age, and the intensity
-of thirst and degree of discomfort.\citep{Pierotti_2018} However, the
-mean fasting duration among participants was uniformly high (mean 16.2
-{[}SD 8.7{]} hours), which may have prevented the researchers from
-discerning a correlation between fasting duration. Even in the subgroup
-of participant who reported 8 hours of fasting, all reported
-thirst.\citep{aroni2012assessment, Pierotti_2018}
-
-It should be noted that an investigation of the dimensionality of the
-7-item PTDS had not been performed prior to our study. A uni-dimensional
-solution was identified in our analysis. This should be expected
-considering the small number of items that were evaluated.
-
-We did not observe an association between PTDS-5 and fasting duration
-for either food or fluids. This would suggest that it is important to
-assess thirst discomfort periodically, regardless of the duration of
-fasting. If thirst discomfort is present, and fasting is still
-indicated, interventions to alleviate symptoms should be implemented.
-There is a lack of evidence regarding the efficacy of any such
-interventions.
-
-\hypertarget{limitations}{%
-\subsubsection{Limitations}\label{limitations}}
-
-Selection bias should be considered because a convenience sample was
-used for this study. Also, it is uncertain if similar results would be
-obtained for patients who are fasted prior to undergoing procedures in
-other settings. Only one patient was excluded from the analysis due to
-missing data, so it is unlikely attrition bias would have exerted a
-major influence on the results. Further research is required to evaluate
-the minimal detectable difference for the PTDS-5 as well as the minimal
-clinically important difference prior to use of this scale as an outcome
-measure in research.
-
-\hypertarget{conclusion-1}{%
-\subsubsection{Conclusion}\label{conclusion-1}}
-
-The items in the PTDS-5 form a reliable and precise way to order
-patients according to their thirst discomfort. Nurses may use this scale
-to assess the severity of thirst discomfort in patients who are fasted
-prior to procedures and initiate interventions according to individual
-needs.
-
-\begin{tabular}{ll}
+\caption{\label{tab:tab1pdf}Participant characteristics}
+\centering
+\begin{tabular}[t]{ll}
 \toprule
@@
 \hspace{1em}Port-a-cath & 4 (2.1\%)\\
-\hspace{1em}Gall bladder stone removal & 0 (0\%)\\
-\hspace{1em}Endovascular aneurysm repair (EVAR) & 0 (0\%)\\
+\hspace{1em}Radiofrequency ablation of a tumour (RFA) & 9 (4.7\%)\\
+\hspace{1em}Vascular embolization (e.g., renal) & 1 (0.5\%)\\
 \addlinespace
-\hspace{1em}Radiofrequency ablation of a tumour (RFA) & 9 (4.7\%)\\
-\hspace{1em}Fistula repair/de-clotting & 0 (0\%)\\
-\hspace{1em}Vascular embolization (e.g., renal) & 1 (0.5\%)\\
 \hspace{1em}Other procedure & 16 (8.3\%)\\
 Time since last clear fluids (hours) & 9.5 (5.0, 13.0)\\
-\addlinespace
 Time since last food (hours) & 13.0 (11.0, 15.0)\\
@@
 \end{tabular}
+\end{table}
+
+\hypertarget{mokken-scale-analysis}{%
+\subsection{Mokken scale analysis}\label{mokken-scale-analysis}}
+
+No items required removal from the PTDS-7 based on the condition that
+scalability coefficients (\emph{H\textsubscript{i}} and lower bound of
+95\% CI) should be over 0.3. We used a lower bound for \emph{c},
+starting from 0.05 and increasing to 0.80 in 0.05 increments, to explore
+the dimensionality of the PTDS-7. From 0.05 to 0.45, all of the items
+formed a single scale. The item, `\emph{I have a bad taste in my
+mouth}', was excluded by selecting the remaining items which showed
+uni-dimensionality at a threshold level of .50. The 6 remaining items
+(PTDS-6) were further examined for the criteria of a Mokken scale and
+item invariant ordering. The PTDS-6 item set had a homogeneity value
+\emph{H(se)} of 0.599, which indicated a strong Mokken scale. No items
+were removed based on the criteria that there should be no significant
+violations of the assumption of monotonicity.
+
+\begin{table}
+
+\caption{\label{tab:hierarchytabpdf}Item hierarchy of the PTDS-5}
+\centering
+\begin{tabular}[t]{lrrr}
+\toprule
+item & mean & Hi & se\\
+\midrule
+I want to drink water & 1.10 & 0.56 & 0.054\\
+My mouth is dry & 0.80 & 0.62 & 0.044\\
+My lips are dry & 0.69 & 0.63 & 0.041\\
+My saliva is thick & 0.51 & 0.57 & 0.050\\
+My tongue is thick & 0.43 & 0.65 & 0.042\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\hypertarget{invariant-item-ordering}{%
+\subsubsection{Invariant item ordering}\label{invariant-item-ordering}}
+
+Analysis of invariant item ordering for the PTDS-6 revealed violations
+for the dry throat/dry lips item pair. Although there were no
+\emph{significant} intersections between items, due to the intersection
+between the dry throat and dry lips items, crit values were above 40.
+For this reason, we excluded one of these items (dry throat), and
+re-assessed the remaining 5 items for the assumptions of a Mokken scale.
+With the dry throat item removed, none of the items violated the
+assumption of invariant item ordering.
+
+\hypertarget{accuracy-and-reliability}{%
+\subsubsection{Accuracy and
+reliability}\label{accuracy-and-reliability}}
+
+The accuracy of the item ordering was found to be acceptable with a
+\emph{H\textsubscript{T}} coefficient of 0.397. Scale reliability was
+adequate with a coefficient score (Rho) of 0.84.
+
+\hypertarget{item-difficulty-hierarchy}{%
+\subsubsection{Item difficulty
+hierarchy}\label{item-difficulty-hierarchy}}
+
+The item difficult hierarchy for the PTDS-5 is presented in Table
+\ref{tab:hierarchytabpdf}. The most highly endorsed symptom (i.e.~the
+least severe symptom of thirst discomfort in the PTDS-5) was being
+bothered by a desire to drink water. The least endorsed symptom
+(i.e.~the most severe symptom of thirst discomfort in the PTDS-5) was
+being bothered by the perception of a `thick tongue'. The items in the
+PTDS-5 form a strong Mokken scale (H\textsubscript{s} = 0.605).
+
+\hypertarget{construct-and-convergent-validity}{%
+\subsection{Construct and convergent
+validity}\label{construct-and-convergent-validity}}
+
+The distribution of PTDS-5 scores in the sample was right-skewed, with a
+median score of 3 (IQR 1, 6). (Figure \ref{fig:distributionpdf}). As
+shown in Figure \ref{fig:globalthirst}, there was a positive correlation
+between PTDS-5 scores and the global thirst discomfort rating
+(\emph{tau} = 0.54; 95\% CI = 0.43 to 0.63). The correlation between
+PTDS-5 scores and thirst intensity was also high (\emph{tau} = 0.49;
+95\% CI = 0.38 to 0.59). Likewise, the correlation between pain and
+PTDS-5 scores was not significant (Spearman's rho = 0.07; 95\% CI =
+-0.07 to 0.21).
 
@@
 
-{\centering \includegraphics[width=252px]{manuscript_files/figure-latex/distribution-1}
+{\centering \includegraphics[width=252px]{manuscript_files/figure-latex/distributionpdf-1}
 
@@
 
-\caption{Distribution of PTDS-5 scores}\label{fig:distribution}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=252px]{manuscript_files/figure-latex/intensity-1}
-
-}
-
-\caption{Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 0 to 10 with higher scores indicating worse thirst discomfort.}\label{fig:intensity}
+\caption{Distribution of PTDS-5 scores (higher scores indicate worse thirst discomfort)}\label{fig:distributionpdf}
 \end{figure}
@@
 
+The associations between PTDS-5 scores with food and fluid duration are
+displayed in Figure \ref{fig:fasting}. Longer duration of fasting was
+not linearly associated with PTDS-5 scores for fluids (Spearman's rho =
+0.1; 95\% CI = -0.05 to 0.23) or food (Spearman's rho = 0.13; 95\% CI =
+-0.01 to 0.27).
+
 \begin{figure}
@@
 
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+Results indicate that the PTDS-5 is a reliable and precise way to order
+patients according to their thirst discomfort in the
+\emph{pre-procedure} period. In addition, an acceptable accuracy for
+invariant item ordering was identified, meaning patients are likely to
+rank the PTDS-5 symptoms of thirst discomfort in the same way. As such,
+the item difficulty hierarchy, presented in Table
+\ref{tab:hierarchytabpdf}, can be used as a guide to easily determine
+which patients are experiencing worse thirst discomfort. The
+hierarchical pattern of responses to items in the Mokken scale is
+interpretable in terms of respondents more easily endorsing items
+related to a general desire to drink water through to first endorsing
+more specific symptoms related to dryness of the mouth and lips prior to
+endorsing discomfort associated with the sensations of `thick' saliva
+and `thick' tongue. In more practical terms, a patient who reports
+feeling bothered by a sensation that saliva is thick, could be
+interpreted as experiencing worse thirst discomfort than a patient who
+did not endorse any of the items lower down on the item difficulty
+hierarchy (i.e.~`my lips are dry' and `my mouth is dry').
+
+The original (7-item) perioperative thirst-discomfort scale has
+previously been used to determine the relationship between thirst
+intensity and thirst-discomfort in the immediate postoperative period
+during anesthesia recovery.\citep{Pierotti_2018} In a non-probabilistic
+sample of 203 adult participants in post-operative recovery following
+elective, urgent and emergency surgeries, the investigators found no
+association between fasting duration, or patient age, and the intensity
+of thirst and degree of discomfort \citep{Pierotti_2018}. However, the
+mean fasting duration among participants was uniformly high (mean 16.2
+{[}SD 8.7{]} hours), which may have prevented the researchers from
+discerning a correlation between fasting duration. Even in the subgroup
+of participant who reported 8 hours of fasting, all reported thirst
+\citep{aroni2012assessment, Pierotti_2018}.
+
+It should be noted that an investigation of the dimensionality of the
+7-item PTDS had not been performed prior to our study. A uni-dimensional
+solution was identified in our analysis. This should be expected
+considering the small number of items that were evaluated.
+
+We did not observe an association between PTDS-5 and fasting duration
+for either food or fluids. This would suggest that it is important to
+assess thirst discomfort periodically, regardless of the duration of
+fasting. If thirst discomfort is present, and fasting is still
+indicated, interventions to alleviate symptoms should be implemented.
+There is a lack of evidence regarding the efficacy of interventions to
+alleviate perioperative thirst discomfort. However, recent randomized
+controlled trials have identified clinically relevant redictions in
+thirst discomfort with menthol chewing gum and 30mL menthol popsicles
+\citep{garcia2019, nascimento2020}.
+
+\hypertarget{limitations}{%
+\subsection{Limitations}\label{limitations}}
+
+Selection bias should be considered because a convenience sample was
+used for this study. Also, it is uncertain if similar results would be
+obtained for patients who are fasted prior to undergoing procedures in
+other settings. Only one patient was excluded from the analysis due to
+missing data, so it is unlikely attrition bias would have exerted a
+major influence on the results. Further research is required to evaluate
+the minimal detectable difference for the PTDS-5 as well as the minimal
+clinically important difference prior to use of this scale as an outcome
+measure in research.
+
+\hypertarget{conclusion}{%
+\subsection{Conclusion}\label{conclusion}}
+
+The items in the PTDS-5 form a reliable and precise way to order
+patients according to their thirst discomfort. Nurses may use this scale
+to assess the severity of thirst discomfort in patients who are fasted
+prior to procedures and initiate interventions according to individual
+needs.
+
 %\showmatmethods
-\showacknow
 
-\bibliography{references}
+\pnasbreak
+
+\bibliography{references.bib}
 \bibliographystyle{jss}

### 2020-08-13 R/create_global_thirst_plot.R

@@ -8,19 +8,20 @@
 #' labs element_blank theme ggsave
 #' @importFrom ggtext geom_richtext
 
-create_global_thirst_plot <- function(data,
-                                      global_thirst_corr) {
+create_global_thirst_plot <- function(data
+                                      # global_thirst_corr
+                                      ) {
 
-  tau <- round(global_thirst_corr$tau, 2)
-  tau_low <- round(global_thirst_corr$CI_low, 2)
-  tau_high <- round(global_thirst_corr$CI_high, 2)
-
-  cor <- tibble(
-    ptds = 8,
-    global_thirst_factor = "Extremely uncomfortable",
-    # text label containing r^2 value
-    label = glue("*tau* = {tau}; (95% CI = {tau_low} to {tau_high})")
-  )
+  # tau <- round(global_thirst_corr$tau, 2)
+  # tau_low <- round(global_thirst_corr$CI_low, 2)
+  # tau_high <- round(global_thirst_corr$CI_high, 2)
+  #
+  # cor <- tibble(
+  #   ptds = 8,
+  #   global_thirst_factor = "Extremely uncomfortable",
+  #   # text label containing r^2 value
+  #   label = glue("*tau* = {tau}; (95% CI = {tau_low} to {tau_high})")
+  # )
 
   data %>%
     ggplot(aes(
@@ -33,16 +34,17 @@ create_global_thirst_plot <- function(data,
       y = element_blank(),
       x = "\nPTDS-5"
     ) +
+    theme_minimal()+
     theme(
       legend.position = "none",
       plot.title.position = "plot",
       text = element_text(size=9)
     ) +
-    geom_richtext(
-      data = cor,
-      aes(label = label),
-      hjust = 1, vjust = 2
-    ) +
+    # geom_richtext(
+    #   data = cor,
+    #   aes(label = label),
+    #   hjust = 1, vjust = 2
+    # ) +
     scale_x_continuous(
       breaks = seq(0, 10, by = 2)
     )

### 2020-08-13 R/create_summary_table.R

@@ -7,6 +7,7 @@
 create_summary_table <- function(data_ptds) {
   data_ptds %>%
     select(age, sex_factor, procedure_factor, fluids_duration, food_duration) %>%
+    droplevels() %>%
     tbl_summary(
       missing = "no",
       label = list(

### 2020-08-13 R/plan.R

@@ -41,8 +41,8 @@ get_analysis_plan <- function() {
     data_ptds = add_ptds_total(ptds5, participation_df), # data with total ptds score added
     summary_table = create_summary_table(data_ptds),
     global_thirst_plot = create_global_thirst_plot(
-      data = data_ptds,
-      global_thirst_corr = global_thirst_corr
+      data = data_ptds
+      # global_thirst_corr = global_thirst_corr
     ),
     global_thirst_plot_png = ggsave(
       plot = global_thirst_plot,
@@ -67,8 +67,8 @@ get_analysis_plan <- function() {
     ),
 
     fasting_duration_ptds_plot = create_fasting_duration_ptds_plot(
-      data_ptds,
-      fluids_corr, food_corr
+      data_ptds
+      # fluids_corr, food_corr
     ),
 
     fasting_duration_ptds_plot_png = ggsave(

### 2020-08-17 manuscript/index.Rmd
plan with word manuscript

@@
 knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.retina = 5)
-# tint options
-library(tint)
-knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
 options(htmltools.dir.version = FALSE)
@@
 drake::loadd(summary_table)
-# gtsummary::as_flextable(summary_table)
-gtsummary::as_kable_extra(summary_table, booktabs = T)
+gtsummary::as_flextable(summary_table)
+# gtsummary::as_kable_extra(summary_table, booktabs = T)
 ```
@@
 ```{r tab.id = "tab2", tab.cap= "Item hierarchy of the PTDS-5"}
-# flextable::flextable(drake::readd(item_hierarchy_df_ptds5)) %>% 
-#   autofit()
-drake::readd(item_hierarchy_df_ptds5) %>% 
-  kableExtra::kable(booktabs = T)
+flextable::flextable(drake::readd(item_hierarchy_df_ptds5)) %>%
+  autofit()
+# drake::readd(item_hierarchy_df_ptds5) %>% 
+#   kableExtra::kable(booktabs = T)
 ```
@@
 drake::readd(H_values_ptds7_table) %>% 
-  # flextable::flextable() %>% 
-  # autofit()
-  kableExtra::kable(booktabs = T)
+  flextable::flextable() %>%
+  autofit()
+  # kableExtra::kable(booktabs = T)
 ```
@@
 
-# flextable::flextable(drake::readd(monotonicity_df_ptds6_table)) %>%  
-#   autofit()
- drake::readd(monotonicity_df_ptds6_table) %>% 
-kableExtra::kable(booktabs = T)
+flextable::flextable(drake::readd(monotonicity_df_ptds6_table)) %>%
+  autofit()
+#  drake::readd(monotonicity_df_ptds6_table) %>% 
+# kableExtra::kable(booktabs = T)
 ```
@@
 ```{r tab.id='aisp', tab.cap.pre="Appendix Table ", tab.cap = "aisp for increasing H thresholds (*c*)"}
-# flextable::flextable(drake::readd(aisp)) %>% 
-  # autofit()
-drake::readd(aisp) %>% 
-  kableExtra::kable(booktabs = T)
+flextable::flextable(drake::readd(aisp)) %>%
+autofit()
+# drake::readd(aisp) %>% 
+#   kableExtra::kable(booktabs = T)
 ```
@@
 drake::readd(iio_ptds6_table) %>% 
-  # flextable() %>% 
-  # autofit()
-  kableExtra::kable(booktabs = T)
+  flextable() %>%
+  autofit()
+  # kableExtra::kable(booktabs = T)
 ```
@@
 drake::readd(iio_ptds5_table) %>%
-    # flextable() %>%
-    # autofit()
-  kableExtra::kable(booktabs = T)
+    flextable() %>%
+    autofit()
+  # kableExtra::kable(booktabs = T)
 ```

### 2020-08-17 R/plan.R

@@ -149,17 +149,17 @@ get_analysis_plan <- function() {
     #  device = "svg"
     #),
 
-    # manuscript = target(
-    #   command = {
-    #     rmarkdown::render(knitr_in("manuscript/index.Rmd"))
-    #     file_out("manuscript/index.docx")
-    #   }
-    # ),
-    manuscript_pdf = target(
+    manuscript = target(
       command = {
-        rmarkdown::render(knitr_in("manuscript/manuscript.Rmd"))
-        file_out("manuscript/manuscript.pdf")
+        rmarkdown::render(knitr_in("manuscript/index.Rmd"))
+        file_out("manuscript/index.docx")
       }
-    )
+    ),
+    # manuscript_pdf = target(
+    #   command = {
+    #     rmarkdown::render(knitr_in("manuscript/manuscript.Rmd"))
+    #     file_out("manuscript/manuscript.pdf")
+    #   }
+    # )
   )
 }

### 2020-09-16 manuscript/manuscript.Rmd
submission to jrn

@@
 ---
-title: Mokken scale analysis of the perioperative thirst discomfort scale
+title: Validating the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures
 author:
@@
 There is a lack of evidence regarding the efficacy of interventions to alleviate perioperative thirst discomfort.
-However, recent randomized controlled trials have identified clinically relevant redictions in thirst discomfort with menthol chewing gum and 30mL menthol popsicles [@garcia2019; @nascimento2020].
+However, recent randomized controlled trials have identified clinically relevant reductions in thirst discomfort with menthol chewing gum and 30mL menthol popsicles [@garcia2019; @nascimento2020].
 

### 2020-09-16 R/plan.R

@@ -149,17 +149,17 @@ get_analysis_plan <- function() {
     #  device = "svg"
     #),
 
-    manuscript = target(
-      command = {
-        rmarkdown::render(knitr_in("manuscript/index.Rmd"))
-        file_out("manuscript/index.docx")
-      }
-    ),
-    # manuscript_pdf = target(
+    # manuscript = target(
     #   command = {
-    #     rmarkdown::render(knitr_in("manuscript/manuscript.Rmd"))
-    #     file_out("manuscript/manuscript.pdf")
+    #     rmarkdown::render(knitr_in("manuscript/index.Rmd"))
+    #     file_out("manuscript/index.docx")
     #   }
-    # )
+    # ),
+    manuscript_pdf = target(
+      command = {
+        rmarkdown::render(knitr_in("manuscript/manuscript.Rmd"))
+        file_out("manuscript/manuscript.pdf")
+      }
+    )
   )
 }

### 2020-09-17 manuscript/manuscript.Rmd
remove footer

@@
 lead_author_surname: Blinded for peer review
-doi_footer: "Pre-print submitted to *Applied Nursing Research*"
+doi_footer: ""
 abstract: |

### 2020-10-09 manuscript/cover-letter/cover-letter.Rmd
R1 for jrn

@@
 ---
-toname: Applied Nursing Research
-toaddress: Editorial Board
-opening: To the editor, 
-closing: Regards,
-spacing: 1
-output: ConwayLetterTemplate::letter
+# toname: Journal of Radiology Nursing
+# toaddress: Editorial Board
+# opening: To the editor, 
+# closing: Regards,
+# spacing: 1
+# output: ConwayLetterTemplate::letter
+output: pdf_document
 ---
 
-On behalf of my co-authors, I am pleased to submit our manuscript entitled “Mokken scale analysis of the perioperative thirst discomfort scale".
+Thank you for providing us with the opportunity to revise out manuscript. We have made revisions to the paper and provide detailed responses to the reviewers' comments below.
 
-Prior research indicates that most patients do not spontaneously verbalize thirst discomfort even when experiencing intense distress.
-As such, providing perioperative nurses with a validated tool that accurately measures thirst discomfort is vital.
-Results from our study provide additional insight regarding the validity of perioperative thirst discomfort scale in patients who were fasted before diagnostic or interventional procedures.
-We have not published this data previously.
+## Reviewer 1
+
+*Minor note not affecting publication decision: Your acronym (PTDS) is very close to PTSD (post-traumatic stress disorder): would you consider shortening it to TDS (thirst discomfort score)?
+Just a suggestion in an effort not to confuse this with something else widely noted in the professional literature.*
+
+> Thank you for the suggestion. We have chosen not to remove the 'P' as it stands for perioperative, which is an important concept for this scale.
+
+*Pages 1-2: Excellent introduction, clear, easy to understand, prepares the reader for the balance of the paper.*
+
+> Thank you
+
+*Page 3 Line 54: You wrote, "The Cronbach alpha score was 0.91." This should be in the results section, it is the only outcome datum you have in the methods section.*
+
+> This Cronbach alpha reported in the methods section was that reported by the previous validation study of the original version of the Perioperative Thirst Discomfort Scale.
+We had already provided an overview of the previous validation studies for the scale in the Introduction so we have removed it from the methods to reduce duplication of content.
+
+*Page 4 line 241 you wrote, "To investigate construct validity we hypothesized that higher scores on the PTDS would be associated with greater fasting duration." Very nicely done, this is helpful*
+
+> Thank you
+
+*Page 4 line 253 you wrote, "All data and R code, along with instructions for how to completely reproduce results of the analyses, is available here." This is quite helpful allowing for replication of your study.*
+
+> Thank you
+
+*Page 6 line 318 you wrote, "As shown in Figure 2, there was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (tau = 0.54; 95% CI = 0.43 to 0.63).
+The correlation between PTDS-5 scores and thirst intensity was also high (tau =0.49; 95% CI = 0.38 to 0.59)." Were these correlations statistically significant?
+Therefore, since n = 193, then for your first tau, the z-score exceeds 11!
+Very low p-score (p \< 0.0001), and for the second test the z-score exceeds 10, also a very low p score...I'd go back to your software and doublecheck my math, but would include these points.*
+
+> Your math is correct - the p-values for these correlations are very low.
+We prefer however to report the confidence intervals for the correlations as they are more informative than the very low p-value.
+Confidence intervals have been reported for all correlations.
+
+*Page 6, line 329, you wrote, "Longer duration of fasting was not linearly associated with PTDS-5 scores for fluids (Spearman's rho = 0.1; 95% CI = -0.05 to 0.23) or food (Spearman's rho = 0.13; 95% CI = -0.01 to 0.27)." Did you look at curvilinear or exponential relationships?
+Just curious, as not all physiologic mechanisms (indeed, quite few) are linear in nature.
+Your scatterplots in figure 3 reflect the lack of linear association but there might be a sine wave type line from lower left to upper right, you might consider checking this.*
+
+> Thank you for the suggestion.
+We have not explored other potential relationships between fasting duration and PTDS-5 scores because this analysis was performed to test our stated hypothesis to evaluate construct validity that higher scores on the PTDS-5 would be associated with longer fasting duration.
+We believe a linear association is correct to evaluate this correlation.
+
+*Page 7 line 382 you wrote, "We did not observe an association between PTDS-5 and fasting duration for either food or fluids.
+This would suggest that it is important to assess thirst discomfort periodically, regardless of the duration of fasting.
+If thirst discomfort is present, and fasting is still indicated, interventions to alleviate symptoms should be implemented." This is an excellent take home message for your reader, very practical and useful.*
+
+> Thank you very much for your detailed review and suggestions for our paper.
+
+*Overall: this could go to press as is...I would add the p scores and run the scatterplot curves myself, but that does not preclude publication of the MS. Very nicely done.*
+
+## Reviewer 2
+
+*Thank you for the opportunity to review your document. This is an interesting study and is well-written. Just a few issues that need attention.*
+
+*Page 1, line 12 - I would use a newer reference unless this is a landmark publication.*
+
+> There is surprisingly little research on the patient experience of thirst in the perioperative literature.
+We have no reason to believe that the patient experience of thirst before surgery has changed since this paper was published because excessive fasting prior to procedures is still prevalent, as demonstrated in our study and the other recent studies we have cited.
+
+*Page 1, line 15 - See reference below. If you have an article that is not dated, (n.d.) is used.*
+
+*Dobson G, Chow L, Filteau L, Flexman A, Hurdle H, Kurrek M, Milkovich R, Perrault MA, Sparrow K, Swart PA, Wong M.
+Guidelines to the Practice of Anesthesia - Revised Edition 2020.
+Can J Anaesth.
+2020 Jan;67(1):64-99.
+doi: 10.1007/s12630-019-01507-4.
+Epub 2019 Nov 27.*
+
+> The date for this reference has been corrected
+
+*Page 1, line 22 - I would reword this sentence. It reads disjointed.*
+
+> Changed to: "It is not common for fasting instructions to be updated even when there are significant delays in procedure start time. "
+
+*Page 1, line 26-34 - I would agree with the thesis of this paragraph if it was written before 2017; however, the fasting guidelines dramatically changed in 2017 and all these articles predate the change.
+I would provide more recent evidence that patients are fasting for long periods despite the changes.
+Also, you should discuss what the current fasting recommendations are.
+The current ASA fasting guidelines allow for clear liquids up to 2 hours before the procedure so I would assume that 12.8 hours average fasting would decrease.
+NOTE: If long fasting times are specific to your institution, you would need to disclose this to support why this study was important to perform or even the utility of the instrument.*
+
+> Excessive pre-procedural fasting is a long-standing issue, which by no means is specific to the institution in which this study was conducted.
+Guidelines from the American Society of Anesthesiology as far back as 1999 recommended 2 hour fasting for clear fluids and 6 hours for a light meal (for healthy patients who are undergiong elective procedures.
+(Practice Guidelines for Preoperative Fasting and the Use of Pharmacologic Agents to Reduce the Risk of Pulmonary Aspiration: Application to Healthy Patients Undergoing Elective Procedures: A Report by the American Society of Anesthesiologists Task Force on Preoperative Fasting.
+Anesthesiology.
+1999;90(3):896-905.)
+
+*Page 2, line 52-75 - I would explain in more detail the limitations of the Martins' study.
+For example, postoperative instrument use is confounded by the surgical procedure where insensible fluid losses can occur or the effect of medications like glycopyrrolate have on thirst.
+This will validate the aim of your study.
+NOTE: I would clarify with the authors of the Martins' study that this was indeed only used and validated postoperatively.
+I read the article and it is unclear to me whether they performed ALL instrument interviewing in the postoperative period.*
+
+> We have included some additional content to describe the potential influence of events that occur during surgery on the psychometric properties of the PTDS:
+>
+> "It should be noted that the initial validation of the PTDS was conducted in one specific point in the perioperative journey (i.e.
+post-surgical patients).
+The psychometric properties of the scale may be influenced by events that occur during the surgical procedure, such as fluid loss and administrations of medications that induce thirst, such as glycopyrrolate.
+A recent large study showed that use of glycopyrrolate was an independent risk factor for moderate to severe post-operative thirst.(Lee et al., 2020) As such, validation of the scale for measuring thirst discomfort before procedures is required."
+>
+> It stated in the article by Martins et al, that participants were recruited from the Post-Anesthetic Care Unit.
+
+*Page 2, line 102 - I would write this in paragraph form or put it into a table.*
+
+> Converted to a paragraph:
+>
+> "Patients were not eligible for the study if they were under 16 years of age, scheduled for an emergency procedure, unable to understand or speak English (unless a translator was available) or if the nurse in pre-procedure bay considered that there was insufficient time prior to anticipated commencement of the procedure for participation in the study."
+
+*Page 3, line 135 - Need reference and it would be nice to actually have a visual depicting the instrument somewhere in the article.*
+
+> We have included the reference to the initial scale validation study. We have also provided a visual of the PTDS-5 (Table 3).
+
+*Page 4, line 256 - Is there any reason that recruiting stopped at 198?
+An explanation as to how you came about this sample size is warranted.
+A recent article discussing sample size choice and Mokken scaling have been published.
+(see attachment)*
+
+> The sample size was decided upon in terms of the resources available. We have added to the statistics section that a formal sample size calculation was not conducted.
+
+*Page 7, line 396 - You may want to note that sample size may be an issue.*
+
+> We have included a comment in the limitations section about the potential that the sample size was insufficient:
+>
+> "A recent study empirically evaluated the sample size required for a scale and found that a minimum of 250 participants would be required to establish scalability of the whole scale.(Watson et al., 2018) It was unclear, however, if this minimum threshold would apply to other scales.
+
+*Page 8, line 409 - This is a strong statement which I believe needs to be reworded due to issues related to external validity.
+I would use "may provide a way to measure thirst" instead of what is written.*
+
+> Changed to 'may provide'.
+
+*The main issue I struggle with is the utility of this instrument especially with new fasting guidelines.
+Some would suggest updating your NPO strategies.
+I can see reordering patients as its primary use making it indeed important in that context.
+However, I would provide more compelling evidence as to why measuring thirst and case reordering is needed in your background section.*
+
+> We totally agree that NPO strategies should be addressed in order to reduce the potential for thirst discomfort in patients scheduled for procedures.
+Unfortunately, this is a difficult problem to solve (Hewson, David W., and Iain Moppett.
+"Preoperative fasting and prevention of pulmonary aspiration in adults: research feast, quality improvement famine."British Journal of Anaesthesia 124.4 (2020): 361-363.) We conducted this study to validate the PTDS as the first step in an ongoing program of research aiming to address this issue.
+However, we believe that for the purposes of this current paper, it is best to restrict the discussion to the validation of the scale rather than about the need to address the causes of thirst.
+>
+> We hadn't considered the potential that cases could be re-ordered based on severity of thirst, but agree this is one way that standardized assessment of thirst discomfort could be useful for practice.
+We have added content to the discussion:
+>
+> "One additional potential course of action could be to rearrange the order of procedures on a schedule based on results of periodic thirst discomfort assessments using the PTDS.
+For example, circumstance may arise where a patient who reports severe thirst discomfort is scheduled to have their procedure later in the day than a patient who is not currently bothered by thirst symptoms.
+Reordering the procedure schedule would reduce the duration of fasting for the patient who is bothered by symptoms of thirst.
+Further research would be required to confirm the effectiveness of this strategy in practice on overall patient satisfaction."

### 2020-10-09 manuscript/manuscript.Rmd
R1 for jrn

@@
 footer_contents: "Perioperative thirst discomfort scale"
-#one_column: true
+one_column: true
 fontsize: 12pt
@@
   pdf: true
-header-includes: >
-  \usepackage[switch]{lineno}
-  \linenumbers
-  \usepackage{setspace}
-  \onehalfspacing
-
+# header-includes: >
+#   \usepackage[switch]{lineno}
+#   \linenumbers
+#   \usepackage{setspace}
+#   \onehalfspacing
 ---
@@
 For example, 'no eating or drinking after midnight' is most common.
-This 'standardized' instruction will not be changed regardless of whether or not there are alterations in scheduling throughout the day that result in significant delays in procedure start time.
+It is not common for fasting instructions to be updated even when there are significant delays in procedure start time.
 As a result, fasting durations far exceed the recommended requirement for most patients undergoing medical and surgical procedures [@de2014actual; @sorita2015frequency; @spitz2017impact].
@@
 Cronbach's alpha was used to test the internal consistency of the scale with a value of 0.91 [@Martins_2017].
+It should be noted that the initial validation of the PTDS was conducted in one specific point in the perioperative journey (i.e. *post*-surgical patients).
+The psychometric properties of the scale may be influenced by events that occur during the surgical procedure, such as fluid loss and administrations of medications that induce thirst, such as glycopyrrolate.
+A recent large study showed that use of glycopyrrolate was an independent risk factor for moderate to severe post-operative thirst.[@lee2020prevalence] As such, validation of the scale for measuring thirst discomfort before procedures is required.
 
-Although the initial validation of the PTDS was conducted in a different context (i.e. *post*-surgical patients), it demonstrates strong potential for accurately assessing thirst-discomfort more generally for peri-operative use.
+The PTDS demonstrates strong potential for accurately assessing thirst-discomfort more generally for peri-operative use.
 The aim of this study was to examine the validity and reliability of the PTDS for measuring thirst discomfort for patients who are fasting *before* medical procedures.
@@
 
-1.  Under 16 years of age
-2.  Scheduled for an emergency procedure
-3.  Unable to understand or speak English and a translator was not immediately available
-4.  Nurse in pre-procedure bay considered that there was insufficient time prior to anticipated commencement of the procedure for participation in the study.
+Patients were not eligible for the study if they were under 16 years of age, scheduled for an emergency procedure, unable to understand or speak English (unless a translator was available) or if the nurse in pre-procedure bay considered that there was insufficient time prior to anticipated commencement of the procedure for participation in the study.
 
@@
 
-The seven-item Perioperative Thirst Discomfort Scale (PTDS-7) is a 7-item self-reported, composite score evaluating the severity of dry mouth, lips, and throat, thick tongue and saliva, a bad taste in the mouth, and a desire to drink water.
-The total score can range from 0 to 14, with 14 representing the "most extreme" intensity of discomfort related to perioperative thirst.
-This instrument was developed and validated for perioperative use in a surgical centre at a public hospital in Brazil.
-Face validation was conducted by group consisting of nurses with PhDs, new graduate nurses, and patients.
-Content validation was conducted by a team consisting of specialists with perioperative nursing experience related to thirst or instrument validation.
-The reliability of the scale was assessed through inter-observer equivalence by a research nurse and a staff nurse from the Post Anesthesia Care Unit among 70 patients pre and postoperatively in an anesthetic recovery room.
-The Cronbach alpha score was 0.91.
-The time taken to complete the PTDS-7 was not evaluated.
+The seven-item Perioperative Thirst Discomfort Scale (PTDS-7) is a 7-item self-reported, composite score evaluating the severity of dry mouth, lips, and throat, thick tongue and saliva, a bad taste in the mouth, and a desire to drink water.[@Martins_2017] The total score can range from 0 to 14, with 14 representing the "most extreme" intensity of discomfort related to perioperative thirst.[@Martins_2017]
 
@@
 All data and R code, along with instructions for how to completely reproduce results of the analyses, is available [here](https://github.com/awconway/ptds).
+A formal sample size calculation was not conducted.
 
@@
 
-```{r hierarchytabpdf, eval=params$pdf}
-drake::readd(item_hierarchy_df_ptds5) %>% 
-  kableExtra::kable(booktabs=T, caption="Item hierarchy of the PTDS-5")
-```
-
 ### Invariant item ordering
@@
 
+```{r hierarchytabpdf, eval=params$pdf}
+drake::readd(item_hierarchy_df_ptds5) %>% 
+  kableExtra::kable(booktabs=T, caption="Item hierarchy of the PTDS-5")
+```
+
 ## Construct and convergent validity
@@
 (Figure <!--\@ref(fig:distribution)-->\ref{fig:distributionpdf}).
-As shown in Figure <!--\@ref(fig:globalthirst)-->\ref{fig:globalthirst}, there was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (*tau* = `r round(global_thirst_corr$tau, 2)`; 95% CI = `r round(global_thirst_corr$CI_low, 2)` to `r round(global_thirst_corr$CI_high, 2)`).
+As shown in Figure <!--\@ref(fig:globalthirst)-->\ref{fig:globalthirst}, there was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (*tau* = `r round(global_thirst_corr$tau, 2)`; 95% CI = `r round(global_thirst_corr$CI_low, 2)` to `r round(global_thirst_corr$CI_high, 2)`; `r round(global_thirst_corr$p, 3)`).
 The correlation between PTDS-5 scores and thirst intensity was also high (*tau* = `r round(intensity_corr$tau, 2)`; 95% CI = `r round(intensity_corr$CI_low, 2)` to `r round(intensity_corr$CI_high, 2)`).
@@
 However, recent randomized controlled trials have identified clinically relevant reductions in thirst discomfort with menthol chewing gum and 30mL menthol popsicles [@garcia2019; @nascimento2020].
+One additional potential course of action could be to rearrange the order of procedures on a schedule based on results of periodic thirst discomfort assessments using the PTDS.
+For example, circumstance may arise where a patient who reports severe thirst discomfort is scheduled to have their procedure later in the day than a patient who is not currently bothered by thirst symptoms.
+Reordering the procedure schedule would reduce the duration of fasting for the patient who is bothered by symptoms of thirst.
+Further research would be required to confirm the effectiveness of this strategy in practice on overall patient satisfaction.
 
@@
 Further research is required to evaluate the minimal detectable difference for the PTDS-5 as well as the minimal clinically important difference prior to use of this scale as an outcome measure in research.
+The potential that the sample size was insufficient for Mokken scale analysis should be considered.
+A recent study empirically evaluated the sample size required for a scale and found that a minimum of 250 participants would be required to establish scalability of the whole scale.[@watson2018] It was unclear, however, if this minimum threshold would apply to other scales.
 
@@
 
-The items in the PTDS-5 form a reliable and precise way to order patients according to their thirst discomfort.
+The items in the PTDS-5 may provide a reliable and precise way to order patients according to their thirst discomfort.
 Nurses may use this scale to assess the severity of thirst discomfort in patients who are fasted prior to procedures and initiate interventions according to individual needs.
+The PTDS-5 with the rating scale is presented in Table \ref{tab:tabptds}.
+
+```{r tabptds, eval=params$pdf}
+tibble::tribble(~"How bothered are you by the following?", ~"Not at all", ~"Slightly", ~"Extremely",
+                "I want to drink water", 0, 1, 2,
+                "My mouth is dry", 0, 1, 2,
+                "My lips is dry", 0, 1, 2,
+                "My saliva is thick", 0, 1, 2,
+                "My tongue is thick", 0, 1, 2
+) %>% 
+kableExtra::kable(booktabs=T, caption="5-item perioperative thirst discomfort scale (PTDS-5)")
+```
 

### 2020-10-09 manuscript/manuscript.tex
R1 for jrn

@@
-\documentclass[12pt,twocolumn,twoside,]{pinp}
+\documentclass[12pt,twoside,]{pinp}
 
@@
 
-\usepackage[switch]{lineno} \linenumbers \usepackage{setspace} \onehalfspacing
-
+\usepackage{setspace} \onehalfspacing
 \usepackage{booktabs}
@@
 
-\title{Mokken scale analysis of the perioperative thirst discomfort
-scale}
+\title{Validating the perioperative thirst discomfort scale for
+measuring thirst discomfort prior to procedures}
 
@@
 % new name is doi_footer
-\doifooter{Pre-print submitted to \emph{Applied Nursing Research}}
 
@@
 pre-specified time interval before procedures. For example, `no eating
-or drinking after midnight' is most common. This `standardized'
-instruction will not be changed regardless of whether or not there are
-alterations in scheduling throughout the day that result in significant
-delays in procedure start time. As a result, fasting durations far
-exceed the recommended requirement for most patients undergoing medical
-and surgical procedures
+or drinking after midnight' is most common. It is not common for fasting
+instructions to be updated even when there are significant delays in
+procedure start time. As a result, fasting durations far exceed the
+recommended requirement for most patients undergoing medical and
+surgical procedures
 \citep{de2014actual, sorita2015frequency, spitz2017impact}. For example,
@@
 alpha was used to test the internal consistency of the scale with a
-value of 0.91 \citep{Martins_2017}.
+value of 0.91 \citep{Martins_2017}. It should be noted that the initial
+validation of the PTDS was conducted in one specific point in the
+perioperative journey (i.e.~\emph{post}-surgical patients). The
+psychometric properties of the scale may be influenced by events that
+occur during the surgical procedure, such as fluid loss and
+administrations of medications that induce thirst, such as
+glycopyrrolate. A recent large study showed that use of glycopyrrolate
+was an independent risk factor for moderate to severe post-operative
+thirst.\citep{lee2020prevalence} As such, validation of the scale for
+measuring thirst discomfort before procedures is required.
 
-Although the initial validation of the PTDS was conducted in a different
-context (i.e.~\emph{post}-surgical patients), it demonstrates strong
-potential for accurately assessing thirst-discomfort more generally for
-peri-operative use. The aim of this study was to examine the validity
-and reliability of the PTDS for measuring thirst discomfort for patients
-who are fasting \emph{before} medical procedures.
+The PTDS demonstrates strong potential for accurately assessing
+thirst-discomfort more generally for peri-operative use. The aim of this
+study was to examine the validity and reliability of the PTDS for
+measuring thirst discomfort for patients who are fasting \emph{before}
+medical procedures.
 
@@
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Under 16 years of age
-\item
-  Scheduled for an emergency procedure
-\item
-  Unable to understand or speak English and a translator was not
-  immediately available
-\item
-  Nurse in pre-procedure bay considered that there was insufficient time
-  prior to anticipated commencement of the procedure for participation
-  in the study.
-\end{enumerate}
+Patients were not eligible for the study if they were under 16 years of
+age, scheduled for an emergency procedure, unable to understand or speak
+English (unless a translator was available) or if the nurse in
+pre-procedure bay considered that there was insufficient time prior to
+anticipated commencement of the procedure for participation in the
+study.
 
@@
 mouth, lips, and throat, thick tongue and saliva, a bad taste in the
-mouth, and a desire to drink water. The total score can range from 0 to
-14, with 14 representing the ``most extreme'' intensity of discomfort
-related to perioperative thirst. This instrument was developed and
-validated for perioperative use in a surgical centre at a public
-hospital in Brazil. Face validation was conducted by group consisting of
-nurses with PhDs, new graduate nurses, and patients. Content validation
-was conducted by a team consisting of specialists with perioperative
-nursing experience related to thirst or instrument validation. The
-reliability of the scale was assessed through inter-observer equivalence
-by a research nurse and a staff nurse from the Post Anesthesia Care Unit
-among 70 patients pre and postoperatively in an anesthetic recovery
-room. The Cronbach alpha score was 0.91. The time taken to complete the
-PTDS-7 was not evaluated.
+mouth, and a desire to drink water.\citep{Martins_2017} The total score
+can range from 0 to 14, with 14 representing the ``most extreme''
+intensity of discomfort related to perioperative
+thirst.\citep{Martins_2017}
 
@@
 to completely reproduce results of the analyses, is available
-\href{https://github.com/awconway/ptds}{here}.
+\href{https://github.com/awconway/ptds}{here}. A formal sample size
+calculation was not conducted.
 
@@
 
-\begin{table}
-
-\caption{\label{tab:hierarchytabpdf}Item hierarchy of the PTDS-5}
-\centering
-\begin{tabular}[t]{lrrr}
-\toprule
-item & mean & Hi & se\\
-\midrule
-I want to drink water & 1.10 & 0.56 & 0.054\\
-My mouth is dry & 0.80 & 0.62 & 0.044\\
-My lips are dry & 0.69 & 0.63 & 0.041\\
-My saliva is thick & 0.51 & 0.57 & 0.050\\
-My tongue is thick & 0.43 & 0.65 & 0.042\\
-\bottomrule
-\end{tabular}
-\end{table}
-
 \hypertarget{invariant-item-ordering}{%
@@
 
+\begin{table}
+
+\caption{\label{tab:hierarchytabpdf}Item hierarchy of the PTDS-5}
+\centering
+\begin{tabular}[t]{lrrr}
+\toprule
+item & mean & Hi & se\\
+\midrule
+I want to drink water & 1.10 & 0.56 & 0.054\\
+My mouth is dry & 0.80 & 0.62 & 0.044\\
+My lips are dry & 0.69 & 0.63 & 0.041\\
+My saliva is thick & 0.51 & 0.57 & 0.050\\
+My tongue is thick & 0.43 & 0.65 & 0.042\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \hypertarget{construct-and-convergent-validity}{%
@@
 between PTDS-5 scores and the global thirst discomfort rating
-(\emph{tau} = 0.54; 95\% CI = 0.43 to 0.63). The correlation between
+(\emph{tau} = 0.54; 95\% CI = 0.43 to 0.63; 0). The correlation between
 PTDS-5 scores and thirst intensity was also high (\emph{tau} = 0.49;
@@
 
-{\centering \includegraphics[width=252px]{manuscript_files/figure-latex/distributionpdf-1}
+{\centering \includegraphics[width=252px]{distributionpdf-1}
 
@@
 
-{\centering \includegraphics[width=252px]{manuscript_files/figure-latex/globalthirst-1}
+{\centering \includegraphics[width=252px]{globalthirst-1}
 
@@
 
-{\centering \includegraphics[width=252px]{manuscript_files/figure-latex/fasting-1}
+{\centering \includegraphics[width=252px]{fasting-1}
 
@@
 endorsing discomfort associated with the sensations of `thick' saliva
-and `thick' tongue. In more practical terms, a patient who reports
-feeling bothered by a sensation that saliva is thick, could be
-interpreted as experiencing worse thirst discomfort than a patient who
-did not endorse any of the items lower down on the item difficulty
-hierarchy (i.e.~`my lips are dry' and `my mouth is dry').
+and `thick' tongue.
+In more practical terms, a patient who reports feeling bothered by a sensation of thick saliva could be interpreted as experiencing worse thirst discomfort than a patient who did not endorse any of the items lower down on the item difficulty hierarchy (i.e.~`my lips are dry' and `my mouth is dry').
 
@@
 alleviate perioperative thirst discomfort. However, recent randomized
-controlled trials have identified clinically relevant redictions in
+controlled trials have identified clinically relevant reductions in
 thirst discomfort with menthol chewing gum and 30mL menthol popsicles
-\citep{garcia2019, nascimento2020}.
+\citep{garcia2019, nascimento2020}. One additional potential course of
+action could be to rearrange the order of procedures on a schedule based
+on results of periodic thirst discomfort assessments using the PTDS. For
+example, circumstance may arise where a patient who reports severe
+thirst discomfort is scheduled to have their procedure later in the day
+than a patient who is not currently bothered by thirst symptoms.
+Reordering the procedure schedule would reduce the duration of fasting
+for the patient who is bothered by symptoms of thirst. Further research
+would be required to confirm the effectiveness of this strategy in
+practice on overall patient satisfaction.
 
@@
 clinically important difference prior to use of this scale as an outcome
-measure in research.
+measure in research. The potential that the sample size was insufficient
+for Mokken scale analysis should be considered. A recent study
+empirically evaluated the sample size required for a scale and found
+that a minimum of 250 participants would be required to establish
+scalability of the whole scale.\citep{watson2018} It was unclear,
+however, if this minimum threshold would apply to other scales.
 
@@
 
-The items in the PTDS-5 form a reliable and precise way to order
+The items in the PTDS-5 may provide a reliable and precise way to order
 patients according to their thirst discomfort. Nurses may use this scale
@@
 prior to procedures and initiate interventions according to individual
-needs.
+needs. The PTDS-5 with the rating scale is presented in Table
+\ref{tab:tabptds}.
+
+\begin{table}
+
+\caption{\label{tab:tabptds}5-item perioperative thirst discomfort scale (PTDS-5)}
+\centering
+\begin{tabular}[t]{lrrr}
+\toprule
+How bothered are you by the following? & Not at all & Slightly & Extremely\\
+\midrule
+I want to drink water & 0 & 1 & 2\\
+My mouth is dry & 0 & 1 & 2\\
+My lips is dry & 0 & 1 & 2\\
+My saliva is thick & 0 & 1 & 2\\
+My tongue is thick & 0 & 1 & 2\\
+\bottomrule
+\end{tabular}
+\end{table}
 

### 2020-10-09 R/create_deviation_plot.R

@@ -229,7 +229,7 @@ create_deviation_plot <- function(data_ptds, coded_descriptions) {
             strip.background = element_blank(),
             plot.title = element_markdown(size = 14, margin = margin(0, 0, 20, 0)),
             plot.title.position = "plot",
-                      text = element_text(family = "BioRhyme", size = 18)
+                      text = element_text( size = 18)
 
         )
 

### 2020-10-09 R/plan.R

@@ -106,14 +106,14 @@ get_analysis_plan <- function() {
 
     # deviations
 
-    #coded_descriptions = read.csv(
-    #  here::here("Analysis/fasting_descriptions_complete.csv")
-    #),
+    coded_descriptions = read.csv(
+     here::here("data/fasting_descriptions_complete.csv")
+    ),
 
-    #deviation_plot = create_deviation_plot(
-    #  data_ptds = data_ptds,
-    #  coded_descriptions = coded_descriptions
-    #),
+    deviation_plot = create_deviation_plot(
+     data_ptds = data_ptds,
+     coded_descriptions = coded_descriptions
+    ),
     #deviation_plot_png = ggsave(
     #  plot = deviation_plot,
     #  filename = here("plots", "fasting-deviation-plot.png"),
